---
title: "1.Overall_associations"
output: html_document
date: "2025-12-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Overall Association

The purpose of this script is to analyze the global association between the transcriptome and the microbiome

## load data
```{r}
rnaseq_count <- read.csv("../../rnaseq/2.rnaseq_analysis/output/counts_uncorr_fin_woOutlier.csv", header = TRUE, row.names = 1)
rnaseq_meta <- read.csv("../../rnaseq/2.rnaseq_analysis/output/metadata_fin_woOutlier.csv", header = TRUE, row.names = 1)
micro_ps <- readRDS("../../microbiome/3.microbiome_analysis/output/phyloseq_sampleReads_unadj.rds")
```

## load packages
```{r}
library(ggplot2)
library(DESeq2)
library(phyloseq)
library(vegan)
library(lme4)
```

## load functions
```{r}
## microbiome correction
correct_microbiome_region <- function(micro_ordinated, region_ps) {
  # create empty result matrix with same dimensions
  micro_corrected <- as.data.frame(
    matrix(nrow = nrow(micro_ordinated), ncol = ncol(micro_ordinated))
  )
  colnames(micro_corrected) <- colnames(micro_ordinated)
  rownames(micro_corrected) <- rownames(micro_ordinated)

  # add variables
  micro_ordinated$i.Sex <- as.factor(region_ps@sam_data$i.Sex)
  micro_ordinated$i.Age <- as.numeric(region_ps@sam_data$i.Age)
  micro_ordinated$ex.ExtractionBatch <- as.factor(region_ps@sam_data$ex.ExtractionBatch)
  micro_ordinated$p.CleanedPCRConcentration <- as.numeric(region_ps@sam_data$p.CleanedPCRConcentration)

  n <- 1
  set.seed(123)
  for (i in colnames(micro_corrected)) {
    obj <- lm(
      as.formula(
        paste0("`", i, "`", " ~ i.Sex + i.Age + ex.ExtractionBatch + p.CleanedPCRConcentration")
      ),
      data = micro_ordinated
    )
    micro_corrected[, i] <- resid(obj)
    print(n)
    n <- n + 1
  }

  return(micro_corrected)
}


## rnaseq correction
correct_rnaseq_region <- function(rna_ordinated, region_meta) {
  # create empty result matrix with same dimensions
  rna_corrected <- as.data.frame(
    matrix(nrow = nrow(rna_ordinated), ncol = ncol(rna_ordinated))
  )
  colnames(rna_corrected) <- colnames(rna_ordinated)
  rownames(rna_corrected) <- rownames(rna_ordinated)

  # add variables
  rna_ordinated$i.Sex <- as.factor(region_meta$i.Sex)
  rna_ordinated$i.Age <- as.numeric(region_meta$i.Age)
  rna_ordinated$p.LibraryConcentration_ngul <- as.numeric(region_meta$p.LibraryConcentration_ngul)
  rna_ordinated$s.SequencingBatch <- as.factor(region_meta$s.SequencingBatch)
  rna_ordinated$ex.Extractedby <- as.factor(region_meta$ex.Extractedby)
  rna_ordinated$ex.RINscore <- as.numeric(region_meta$ex.RINscore)

  n <- 1
  set.seed(123)
  for (i in colnames(rna_corrected)) {
    obj <- lm(
      as.formula(
        paste0("`", i, "`", "~ i.Sex + i.Age + p.LibraryConcentration_ngul + ex.RINscore + ex.Extractedby + s.SequencingBatch")
      ),
      data = rna_ordinated
    )
    rna_corrected[, i] <- resid(obj)
    print(n)
    n <- n + 1
  }

  return(rna_corrected)
}
```

## filter to overlapping CD patients and adjust order
```{r}
micro_overlap_ps <- subset_samples(micro_ps, UNQ_ID != "NA")
CD_overlap_ps <- subset_samples(micro_overlap_ps, i.Disease == "Crohns")

rnaseq_meta_overlap <- subset(rnaseq_meta, UNQ_ID != "NA")
rna_meta_CD <- subset(rnaseq_meta_overlap, i.Disease == "Crohns")

## adjust order of rna metadata based on microbiome
rna_meta_CD_reorder <- rna_meta_CD[order(match(rna_meta_CD$UNQ_ID, CD_overlap_ps@sam_data$UNQ_ID)),]

rna_count_CD <- subset(rnaseq_count, rownames(rnaseq_count) %in% rna_meta_CD_reorder$p.LibraryID)

## adjust order of rna count based on rna metadata
rna_count_CD_reorder <- rna_count_CD[order(match(rownames(rna_count_CD), rna_meta_CD_reorder$p.LibraryID)),]

rownames(rna_count_CD_reorder) <- rna_meta_CD_reorder$UNQ_ID

rna_count_CD_reorder_t <- t(rna_count_CD_reorder)

## check order
identical(CD_overlap_ps@sam_data$UNQ_ID, rownames(rna_count_CD_reorder))
```

## calculate distance for procrustes across intestine
```{r}
# calculate aitchison distance for microbes and ordinate
CD_overlap_RA <- transform_sample_counts(CD_overlap_ps, function(otu) otu/sum(otu))

set.seed(123)
CD_overlap_otu <- as.data.frame(as.matrix(CD_overlap_RA@otu_table))
identical(CD_overlap_ps@sam_data$s.SequencingID, rownames(CD_overlap_otu))
rownames(CD_overlap_otu) <- CD_overlap_ps@sam_data$UNQ_ID

micro_uncorr_dist <- vegdist(CD_overlap_otu, method = "aitchison", pseudocount = 0.000001)
micro_pcoa <- ape::pcoa(micro_uncorr_dist)

micro_ord <- as.data.frame(micro_pcoa$vectors)


# calculate aitchison distance for rnaseq
## transform
set.seed(123)
dds_full <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ 1)

## normalize across at least 1/3 of the samples
smallestGroupSize <- ncol(rna_count_CD_reorder_t)/3
  keep <- rowSums(counts(dds_full) >= 10) >= smallestGroupSize
  dds_full <- dds_full[keep,]
set.seed(123)
vsd_full <- vst(dds_full, blind=FALSE)
mat_full <- assay(vsd_full)


## distance
mat_t <- t(mat_full)
set.seed(123)
rna_uncorr_dist <- vegdist(mat_t, method = "aitchison", pseudocount = 0.000001) 
rna_pcoa <- ape::pcoa(rna_uncorr_dist)

rnaseq_ord <- as.data.frame(rna_pcoa$vectors)
```

## correct matrices for covariates, including specimen
```{r}
# microbiome
micro_ord_corr <- as.data.frame(matrix(nrow = nrow(micro_ord), ncol = ncol(micro_ord)))

colnames(micro_ord_corr) <- colnames(micro_ord)
rownames(micro_ord_corr) <- rownames(micro_ord)

### check names
identical(CD_overlap_ps@sam_data$UNQ_ID, rownames(micro_ord_corr))

### add variable
micro_ord$i.Sex <- as.factor(CD_overlap_ps@sam_data$i.Sex)
micro_ord$i.Age <- as.numeric(CD_overlap_ps@sam_data$i.Age)
micro_ord$ex.ExtractionBatch <- as.factor(CD_overlap_ps@sam_data$ex.ExtractionBatch)
micro_ord$p.CleanedPCRConcentration <- as.numeric(CD_overlap_ps@sam_data$p.CleanedPCRConcentration)
micro_ord$i.PatientID <- as.factor(CD_overlap_ps@sam_data$i.PatientID)
micro_ord$i.Specimen <- as.factor(CD_overlap_ps@sam_data$i.Specimen)

n = 1

set.seed(123)
for(i in colnames(micro_ord_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + ex.ExtractionBatch + p.CleanedPCRConcentration + i.Specimen + (1 | i.PatientID)")),  data = micro_ord)
  micro_ord_corr[, i] <- resid(obj)
  print(n)
  n = n + 1
}


# rnaseq
rnaseq_ord_corr <- as.data.frame(matrix(nrow = nrow(rnaseq_ord), ncol = ncol(rnaseq_ord)))

colnames(rnaseq_ord_corr) <- colnames(rnaseq_ord)
rownames(rnaseq_ord_corr) <- rownames(rnaseq_ord)

### check order
identical(rna_meta_CD_reorder$UNQ_ID, rownames(rnaseq_ord_corr))

### add variable
rnaseq_ord$i.Sex <- as.factor(rna_meta_CD_reorder$i.Sex)
rnaseq_ord$i.Age <- as.numeric(rna_meta_CD_reorder$i.Age)
rnaseq_ord$p.LibraryConcentration_ngul <- as.numeric(rna_meta_CD_reorder$p.LibraryConcentration_ngul)
rnaseq_ord$s.SequencingBatch <- as.factor(rna_meta_CD_reorder$s.SequencingBatch)
rnaseq_ord$ex.Extractedby <- as.factor(rna_meta_CD_reorder$ex.Extractedby)
rnaseq_ord$ex.RINscore <- as.numeric(rna_meta_CD_reorder$ex.RINscore)
rnaseq_ord$i.PatientID <- as.factor(rna_meta_CD_reorder$i.PatientID)
rnaseq_ord$i.Specimen <- as.factor(rna_meta_CD_reorder$i.Specimen)

n = 1

set.seed(123)
for(i in colnames(rnaseq_ord_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.LibraryConcentration_ngul + ex.RINscore + ex.Extractedby + s.SequencingBatch + i.Specimen + (1 | i.PatientID)")),  data = rnaseq_ord)
  rnaseq_ord_corr[, i] <- resid(obj)
  print(n)
  n = n + 1
}

## check order
identical(rownames(micro_ord_corr), rownames(rnaseq_ord_corr))
```

## procrustes across intestine
```{r}
set.seed(123)
pro <- protest(X = rnaseq_ord_corr, Y = micro_ord_corr, scores = "sites", symmetric = TRUE, permutations = 99999)
# plot(pro)

pro_vals <- pro$t
pro_vals <- as.data.frame(pro_vals)
pro_nonper <- pro$t0

procrustes_dist_plot <- ggplot(data = pro_vals, aes(x = pro_vals)) +
  geom_histogram(binwidth = 0.01, fill = "dark grey") +
    geom_vline(xintercept = pro_nonper, lty = 2, color = "red") +
    labs(x = "Procrustes distribution", y = "Count") +
    theme(axis.text.x = element_text(hjust = 0.95, vjust=0.95)) + 
    theme(legend.position = "none")

ggsave(file="../figures/procrustes_dist_plot.pdf", plot=procrustes_dist_plot, width=6, height=3)


df_pro_Y <- data.frame(micro1 = pro$Yrot[,1], micro2 = pro$Yrot[,2])
df_pro_X <- data.frame(rna1 = pro$X[,1], rna2 = pro$X[,2])


procrustes_rna_plot <- ggplot(data = df_pro_X, aes(x = rna1, y = rna2)) +
  geom_point(color = c("#009194"), pch = 16, size = 3) +
  lims(x = c(-0.03, 0.05), y = c(-0.1, 0.11)) +
  labs(x = "rna PCOA1 (39.8%)",
       y = "rna PCOA2 (9.3%)")

ggsave(file="../figures/procrustes_rna_plot.pdf", plot=procrustes_rna_plot, width=3.5, height=3)
  

procrustes_mic_plot <- ggplot(data = df_pro_Y, aes(x = micro1, y = micro2)) +
  geom_point(color = c("black"), pch = 21, size = 3) +
  lims(x = c(-0.03, 0.05), y = c(-0.1, 0.11)) +
  labs(x = "micro PCOA1 (6.8%)",
       y = "micro PCOA2 (4.9%)")

ggsave(file="../figures/procrustes_mic_plot.pdf", plot=procrustes_mic_plot, width=3.5, height=3)


procrustes_rna_mic_plot <- ggplot() +
  geom_point(data = df_pro_X, color = c("#009194"), pch = 16, size = 3, aes(x = rna1, y = rna2)) +
  geom_point(data = df_pro_Y, color = c("black"), pch = 21, size = 3, aes(x = micro1, y = micro2)) +
  geom_segment(alpha = 0.3, aes(x = df_pro_Y$micro1, y = df_pro_Y$micro2), xend = df_pro_X$rna1, yend = df_pro_X$rna2, arrow = arrow(length=unit(0.2,"cm"))) +
  lims(x = c(-0.03, 0.05), y = c(-0.1, 0.11)) +
  labs(x = "Dim 1",
       y = "Dim 2")

ggsave(file="../figures/procrustes_rna_mic_plot.pdf", plot=procrustes_rna_mic_plot, width=4, height=3.5)
```

## filter data by region
```{r}
## filter by region
rna_meta_TI <- subset(rna_meta_CD_reorder, i.Specimen == "Terminal Ileum")
rna_meta_C <- subset(rna_meta_CD_reorder, i.Specimen == "Cecum")
rna_meta_RC <- subset(rna_meta_CD_reorder, i.Specimen == "Right Colon")

micro_TI_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Terminal Ileum")
micro_C_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Cecum")
micro_RC_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Right Colon")
```

## calculate distance for procrustes by region
```{r}
# calculate aitchison distance for microbes and ordinate
set.seed(123)
TI_ps_RA <- transform_sample_counts(micro_TI_ps, function(otu) otu/sum(otu))

TI_overlap_otu <- as.data.frame(as.matrix(TI_ps_RA@otu_table))
identical(micro_TI_ps@sam_data$s.SequencingID, rownames(TI_overlap_otu))
rownames(TI_overlap_otu) <- micro_TI_ps@sam_data$UNQ_ID

TI_micro_uncorr_dist <- vegdist(TI_overlap_otu, method = "aitchison", pseudocount = 0.000001)
TI_micro_pcoa <- ape::pcoa(TI_micro_uncorr_dist)

micro_ord_TI <- as.data.frame(TI_micro_pcoa$vectors)


set.seed(123)
C_ps_RA <- transform_sample_counts(micro_C_ps, function(otu) otu/sum(otu))

C_overlap_otu <- as.data.frame(as.matrix(C_ps_RA@otu_table))
identical(micro_C_ps@sam_data$s.SequencingID, rownames(C_overlap_otu))
rownames(C_overlap_otu) <- micro_C_ps@sam_data$UNQ_ID

C_micro_uncorr_dist <- vegdist(C_overlap_otu, method = "aitchison", pseudocount = 0.000001)
C_micro_pcoa <- ape::pcoa(C_micro_uncorr_dist)

micro_ord_C <- as.data.frame(C_micro_pcoa$vectors)


set.seed(123)
RC_ps_RA <- transform_sample_counts(micro_RC_ps, function(otu) otu/sum(otu))

RC_overlap_otu <- as.data.frame(as.matrix(RC_ps_RA@otu_table))
identical(micro_RC_ps@sam_data$s.SequencingID, rownames(RC_overlap_otu))
rownames(RC_overlap_otu) <- micro_RC_ps@sam_data$UNQ_ID

RC_micro_uncorr_dist <- vegdist(RC_overlap_otu, method = "aitchison", pseudocount = 0.000001)
RC_micro_pcoa <- ape::pcoa(RC_micro_uncorr_dist)

micro_ord_RC <- as.data.frame(RC_micro_pcoa$vectors)


# calculate aitchison distance
## transform
set.seed(123)
dds_full_spec <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ i.Specimen)

smallestGroupSize_spec <- 30
keep_spec <- rowSums(counts(dds_full_spec) >= 10) >= smallestGroupSize_spec
dds_full_spec <- dds_full_spec[keep_spec,]
set.seed(123)
vsd_spec <- vst(dds_full_spec, blind=FALSE)
mat_spec <- assay(vsd_spec)

mat_spec <- as.data.frame(mat_spec)

mat_TI <- mat_spec[,colnames(mat_spec) %in% rna_meta_TI$UNQ_ID]
mat_C <- mat_spec[,colnames(mat_spec) %in% rna_meta_C$UNQ_ID]
mat_RC <- mat_spec[,colnames(mat_spec) %in% rna_meta_RC$UNQ_ID]

## distance
## TI
mat_TI_t <- t(mat_TI)
set.seed(123)
rna_uncorr_TI_dist <- vegdist(mat_TI_t, method="aitchison") 
rna_pcoa_TI <- ape::pcoa(rna_uncorr_TI_dist)

rnaseq_ord_TI <- as.data.frame(rna_pcoa_TI$vectors)

## C
mat_C_t <- t(mat_C)
set.seed(123)
rna_uncorr_C_dist <- vegdist(mat_C_t, method="aitchison") 
rna_pcoa_C <- ape::pcoa(rna_uncorr_C_dist)

rnaseq_ord_C <- as.data.frame(rna_pcoa_C$vectors)

## RC
mat_RC_t <- t(mat_RC)
set.seed(123)
rna_uncorr_RC_dist <- vegdist(mat_RC_t, method = "aitchison") 
rna_pcoa_RC <- ape::pcoa(rna_uncorr_RC_dist)

rnaseq_ord_RC <- as.data.frame(rna_pcoa_RC$vectors)
```

## correct matrices
```{r}
# TI
identical(rownames(micro_ord_TI), rownames(rnaseq_ord_TI))
micro_ord_corr_TI <- correct_microbiome_region(micro_ord_TI, TI_ps_RA)
rnaseq_ord_corr_TI <- correct_rnaseq_region(rnaseq_ord_TI, rna_meta_TI)

# C
identical(rownames(micro_ord_C), rownames(rnaseq_ord_C))
micro_ord_corr_C <- correct_microbiome_region(micro_ord_C, C_ps_RA)
rnaseq_ord_corr_C <- correct_rnaseq_region(rnaseq_ord_C, rna_meta_C)

# RC
identical(rownames(micro_ord_RC), rownames(rnaseq_ord_RC))
micro_ord_corr_RC <- correct_microbiome_region(micro_ord_RC, RC_ps_RA)
rnaseq_ord_corr_RC <- correct_rnaseq_region(rnaseq_ord_RC, rna_meta_RC)
```

## procrustes per region
```{r}
## TI
set.seed(123)
pro_TI <- protest(X = rnaseq_ord_corr_TI, Y = micro_ord_corr_TI, scores = "sites", symmetric = TRUE, permutations = 99999)
# plot(pro)

pro_vals_TI <- pro_TI$t
pro_vals_TI <- as.data.frame(pro_vals_TI)
pro_nonper_TI <- pro_TI$t0

procrustes_dist_plot_TI <- ggplot(data = pro_vals_TI, aes(x = pro_vals_TI)) +
  geom_histogram(binwidth = 0.01, fill = "dark grey") +
    geom_vline(xintercept = pro_nonper_TI, lty = 2, color = "red") +
    labs(x = "Procrustes distribution", y = "Count") +
    theme(axis.text.x = element_text(hjust = 0.95, vjust=0.95)) + 
    theme(legend.position = "none")

ggsave(file="../figures/procrustes_dist_plot_TI.pdf", plot=procrustes_dist_plot_TI, width=4, height=3)


df_pro_Y_TI <- data.frame(micro1 = pro_TI$Yrot[,1], micro2 = pro_TI$Yrot[,2])
df_pro_X_TI <- data.frame(rna1 = pro_TI$X[,1], rna2 = pro_TI$X[,2])


procrustes_rna_plot_TI <- ggplot(data = df_pro_X_TI, aes(x = rna1, y = rna2)) +
  geom_point(color = c("#009194"), pch = 16, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "rna PCOA1 (19.3%)",
       y = "rna PCOA2 (10.2%)")

ggsave(file="../figures/procrustes_rna_plot_TI.pdf", plot=procrustes_rna_plot_TI, width=3.5, height=3)
  

procrustes_mic_plot_TI <- ggplot(data = df_pro_Y_TI, aes(x = micro1, y = micro2)) +
  geom_point(color = c("black"), pch = 21, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "micro PCOA1 (9.0%)",
       y = "micro PCOA2 (6.9%)")

ggsave(file="../figures/procrustes_mic_plot_TI.pdf", plot=procrustes_mic_plot_TI, width=3.5, height=3)


procrustes_rna_mic_plot_TI <- ggplot() +
  geom_point(data = df_pro_X_TI, color = c("#009194"), pch = 16, size = 3, aes(x = rna1, y = rna2)) +
  geom_point(data = df_pro_Y_TI, color = c("black"), pch = 21, size = 3, aes(x = micro1, y = micro2)) +
  geom_segment(alpha = 0.3, aes(x = df_pro_Y_TI$micro1, y = df_pro_Y_TI$micro2), xend = df_pro_X_TI$rna1, yend = df_pro_X_TI$rna2, arrow = arrow(length=unit(0.2,"cm"))) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "Dim 1",
       y = "Dim 2")

ggsave(file="../figures/procrustes_rna_mic_plot_TI.pdf", plot=procrustes_rna_mic_plot_TI, width=4, height=3.5)



## C
set.seed(123)
pro_C <- protest(X = rnaseq_ord_corr_C, Y = micro_ord_corr_C, scores = "sites", symmetric = TRUE, permutations = 99999)
# plot(pro)

pro_vals_C <- pro_C$t
pro_vals_C <- as.data.frame(pro_vals_C)
pro_nonper_C <- pro_C$t0

procrustes_dist_plot_C <- ggplot(data = pro_vals_C, aes(x = pro_vals_C)) +
  geom_histogram(binwidth = 0.01, fill = "dark grey") +
    geom_vline(xintercept = pro_nonper_C, lty = 2, color = "red") +
    labs(x = "Procrustes distribution", y = "Count") +
    theme(axis.text.x = element_text(hjust = 0.95, vjust=0.95)) + 
    theme(legend.position = "none")

ggsave(file="../figures/procrustes_dist_plot_C.pdf", plot=procrustes_dist_plot_C, width=4, height=3)


df_pro_Y_C <- data.frame(micro1 = pro_C$Yrot[,1], micro2 = pro_C$Yrot[,2])
df_pro_X_C <- data.frame(rna1 = pro_C$X[,1], rna2 = pro_C$X[,2])


procrustes_rna_plot_C <- ggplot(data = df_pro_X_C, aes(x = rna1, y = rna2)) +
  geom_point(color = c("#009194"), pch = 16, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "rna PCOA1 (25.7%)",
       y = "rna PCOA2 (10.2%)")

ggsave(file="../figures/procrustes_rna_plot_C.pdf", plot=procrustes_rna_plot_C, width=3.5, height=3)
  

procrustes_mic_plot_C <- ggplot(data = df_pro_Y_C, aes(x = micro1, y = micro2)) +
  geom_point(color = c("black"), pch = 21, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "micro PCOA1 (7.8%)",
       y = "micro PCOA2 (5.7%)")

ggsave(file="../figures/procrustes_mic_plot_C.pdf", plot=procrustes_mic_plot_C, width=3.5, height=3)


procrustes_rna_mic_plot_C <- ggplot() +
  geom_point(data = df_pro_X_C, color = c("#009194"), pch = 16, size = 3, aes(x = rna1, y = rna2)) +
  geom_point(data = df_pro_Y_C, color = c("black"), pch = 21, size = 3, aes(x = micro1, y = micro2)) +
  geom_segment(alpha = 0.3, aes(x = df_pro_Y_C$micro1, y = df_pro_Y_C$micro2), xend = df_pro_X_C$rna1, yend = df_pro_X_C$rna2, arrow = arrow(length=unit(0.2,"cm"))) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "Dim 1",
       y = "Dim 2")

ggsave(file="../figures/procrustes_rna_mic_plot_C.pdf", plot=procrustes_rna_mic_plot_C, width=4, height=3.5)



## RC
set.seed(123)
pro_RC <- protest(X = rnaseq_ord_corr_RC, Y = micro_ord_corr_RC, scores = "sites", symmetric = TRUE, permutations = 99999)
# plot(pro)

pro_vals_RC <- pro_RC$t
pro_vals_RC <- as.data.frame(pro_vals_RC)
pro_nonper_RC <- pro_RC$t0

procrustes_dist_plot_RC <- ggplot(data = pro_vals_RC, aes(x = pro_vals_RC)) +
  geom_histogram(binwidth = 0.01, fill = "dark grey") +
    geom_vline(xintercept = pro_nonper_RC, lty = 2, color = "red") +
    labs(x = "Procrustes distribution", y = "Count") +
    theme(axis.text.x = element_text(hjust = 0.95, vjust=0.95)) + 
    theme(legend.position = "none")

ggsave(file="../figures/procrustes_dist_plot_RC.pdf", plot=procrustes_dist_plot_RC, width=4, height=3)


df_pro_Y_RC <- data.frame(micro1 = pro_RC$Yrot[,1], micro2 = pro_RC$Yrot[,2])
df_pro_X_RC <- data.frame(rna1 = pro_RC$X[,1], rna2 = pro_RC$X[,2])


procrustes_rna_plot_RC <- ggplot(data = df_pro_X_RC, aes(x = rna1, y = rna2)) +
  geom_point(color = c("#009194"), pch = 16, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "rna PCOA1 (14.0%)",
       y = "rna PCOA2 (10.4%)")

ggsave(file="../figures/procrustes_rna_plot_RC.pdf", plot=procrustes_rna_plot_RC, width=3.5, height=3)
  

procrustes_mic_plot_RC <- ggplot(data = df_pro_Y_RC, aes(x = micro1, y = micro2)) +
  geom_point(color = c("black"), pch = 21, size = 3) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "micro PCOA1 (8.4%)",
       y = "micro PCOA2 (7.8%)")

ggsave(file="../figures/procrustes_mic_plot_RC.pdf", plot=procrustes_mic_plot_RC, width=3.5, height=3)


procrustes_rna_mic_plot_RC <- ggplot() +
  geom_point(data = df_pro_X_RC, color = c("#009194"), pch = 16, size = 3, aes(x = rna1, y = rna2)) +
  geom_point(data = df_pro_Y_RC, color = c("black"), pch = 21, size = 3, aes(x = micro1, y = micro2)) +
  geom_segment(alpha = 0.3, aes(x = df_pro_Y_RC$micro1, y = df_pro_Y_RC$micro2), xend = df_pro_X_RC$rna1, yend = df_pro_X_RC$rna2, arrow = arrow(length=unit(0.2,"cm"))) +
  lims(x = c(-0.15, 0.2), y = c(-0.15, 0.16)) +
  labs(x = "Dim 1",
       y = "Dim 2")

ggsave(file="../figures/procrustes_rna_mic_plot_RC.pdf", plot=procrustes_rna_mic_plot_RC, width=4, height=3.5)
```
