---
title: "2.group_association"
output: html_document
date: "2025-12-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2. Group Association

The purpose of this script is to analyze the group-level association between co-expressed genes and co-abundant microbes

## load data
```{r}
rnaseq_count <- read.csv("../../rnaseq/2.rnaseq_analysis/output/counts_uncorr_fin_woOutlier.csv", header = TRUE, row.names = 1)
rnaseq_meta <- read.csv("../../rnaseq/2.rnaseq_analysis/output/metadata_fin_woOutlier.csv", header = TRUE, row.names = 1)
micro_ps <- readRDS("../../microbiome/3.microbiome_analysis/output/phyloseq_sampleReads_unadj.rds")

# protein coding genes
anno_protein_coding <- read.table("../data/protein_coding.txt", header = TRUE)
hsa_to_path <- read.csv("../data/hsa_to_path.csv", row.names = 1, header = TRUE)
pathway_list <- read.csv("../data/pathway_list.csv", header = TRUE)
PID_path <- read.csv("../data/PID_path_entrez.csv", header = FALSE)
```

## load packages
```{r}
library(phyloseq)
library(PMA)
library(DESeq2)
library(lme4)
library(data.table)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(clusterProfiler)
```

## sparseCCA functions
```{r}
## this code is modified from Priya et al: https://github.com/blekhmanlab/host_gene_microbiome_interactions/tree/main/Tutorial/sparseCCA_tutorial
tune_params_grid_search <- function( X, Y){
  scoreXcv <- c()
  scoreYcv <- c()
  penaltyX <- seq(0.15,0.4,length=10)
  penaltyY <- seq(0.1,0.4,length=10)
  corr_demo <- matrix(nrow = length(penaltyX), ncol =  length(penaltyY))
  num_samples <- nrow(X)
  start_time <- Sys.time()
  for( i in 1:length(penaltyX)){
    for(j in 1:length(penaltyY)){
      # print(paste0("Index: i = ",i,", j =", j)); flush.console()
      for(k in 1:num_samples){
       
        print(paste0("Index: i = ",i,", j =", j," k = ",k)); flush.console()
        #compute weights with sample k held out:
        #Default niter = 15 edited to 5 to speed this up.
        res <- PMA::CCA(X[-k,],Y[-k,], penaltyx = penaltyX[i], penaltyz = penaltyY[j], K=1, niter = 5, trace = F, standardize = T, typex = c("standard"), typez = c("standard"))
        ## Compute scores for k'th sample for first pair of canonical variables
        ## Take weight of features (res$u and res$v) computed using all except 
        ## the kth sample and multiple it by values for the kth sample in the 
        ## feature matrix X and Y.
        scoreXcv[k] <- X[k,]%*%res$u ## single value
        scoreYcv[k] <- Y[k,]%*%res$v ## single value
      }
      ## correlation between scores for X and Y for all held out samples.
      corr_demo[i,j] = cor(scoreXcv,scoreYcv) 
    }
  }
  end_time <- Sys.time()
  time_elapsed <- end_time - start_time
  print(paste0("Time elapsed for param tuning = ", time_elapsed)); flush.console()
  
  row.names(corr_demo) <- as.character(penaltyX)
  colnames(corr_demo) <- as.character(penaltyY)
  
  corr_demo_df <- as.data.frame(corr_demo)
  rownames(corr_demo_df)
  colnames(corr_demo_df)
  
  ##identify best penalty parameters
  # find index with max absolute corr
  bestpenalty <- which(abs(corr_demo) == max(abs(corr_demo)), arr.ind = TRUE)
  bestpenalty
  bestpenaltyX <- penaltyX[bestpenalty[1]]
  bestpenaltyY <- penaltyY[bestpenalty[2]]
  
  return (c(bestpenaltyX,bestpenaltyY))
}

run_sparseCCA <- function(X, Z, CCA.K, penaltyX, penaltyZ, vInit=NULL, outputFile=NULL){
  CCA.out <-  PMA::CCA(X,Z,typex="standard",typez="standard",K=CCA.K,
                  penaltyx=penaltyX,penaltyz=penaltyZ,
                  v=vInit)
  if(!is.null(outputFile)){
    sink(outputFile)
    print(CCA.out)
    sink()
  }
  
  ## add rownames to output factors
  rownames(CCA.out$u) <- colnames(X)
  rownames(CCA.out$v) <- colnames(Z)
  ## compute contribution of selected features to each of the samples.
  CCA_var_genes <- X %*% CCA.out$u ## canonical variance for genes 
  CCA_var_microbes <- Z %*% CCA.out$v ## canonical variance for microbes
  
  return(list(CCA.out, CCA_var_genes, CCA_var_microbes))
  
}

get_avg_features <- function(cca_cov, CCA.K){
  num_features <- 0
  for(k in 1:CCA.K){
    num_features <- num_features + length(which(cca_cov[,k]!=0))
  }
  avg_features <- num_features/CCA.K
}

test_significance_LOOCV <- function(X, Y, bestpenaltyX, bestpenaltyY, num_components){
  cca.k = num_components
  scoresXcv <- matrix(nrow = nrow(X), ncol = cca.k)
  scoresYcv <-  matrix(nrow = nrow(Y), ncol = cca.k)
  corr_pval <- c()
  corr_r <- c()
  for(i in 1:nrow(X)){ #n = no. of samples
    #compute weights with sample i held out:
    res <- PMA::CCA(X[-i,],Y[-i,], penaltyx=bestpenaltyX, penaltyz=bestpenaltyY, K=cca.k, trace = F, typex = c("standard"), typez = c("standard")) ## default niter = 15 which is spit out when trace = T (default)
    ###compute scores for i'th sample for each component (pair of canonical variables)
    for(j in 1:cca.k){
      #print(paste0("i = ", i," K = ", j)); flush.console()
      scoresXcv[i,j] <- X[i,]%*%res$u[,j]
      scoresYcv[i,j] <- Y[i,]%*%res$v[,j]
    }
  }
  ## Test for each components
  for(j in 1:cca.k){
    corr <- cor.test(scoresXcv[,j],scoresYcv[,j]) ## Pearson correlation.
    corr_pval[j] <- corr$p.value
  }
  corr_pval
}

save_CCA_components <- function(CCA.out, CCA.K, dirname){
  ## Print canonical covariates in files 
  for(i in CCA.K){
    print(paste0("Writing significant component = ", i))
    selected_X <- which(CCA.out$u[,i]!=0) 
    selected_X <- rownames(CCA.out$u)[selected_X]
    coeff_X <- unname(CCA.out$u[selected_X,i])
    selected_Z <- which(CCA.out$v[,i]!=0)
    selected_Z <- rownames(CCA.out$v)[selected_Z]
    coeff_Z <- unname(CCA.out$v[selected_Z,i])
    ## Make all vectors of same length to avoid repetition of elements from shorter vectors.
    n <- max(length(selected_X), length(selected_Z))
    length(selected_X) <- n                      
    length(selected_Z) <- n
    length(coeff_X) <- n
    length(coeff_Z) <- n
    selected_XZ <- as.data.frame(cbind(gene = selected_X, gene_coeff = coeff_X,
                                       taxa = selected_Z, taxa_coeff = coeff_Z))
    write.table(selected_XZ, file=paste0(dirname,"gene_taxa_component_",i,".txt"), sep = "\t", col.names = NA)
  }
  
}

```

## functions for corrections within region
```{r}
correct_rnaseq_data_within_region <- function(rna_transformed_region, rna_metadata) {
  # filter to protein-coding genes
  mat_inte_df <- rna_transformed_region[, colnames(rna_transformed_region) %in% anno_protein_coding$symbol]

  # Create an empty matrix for corrected data
  mat_inte_corr <- matrix(ncol = ncol(mat_inte_df), nrow = nrow(mat_inte_df))
  colnames(mat_inte_corr) <- colnames(mat_inte_df)
  rownames(mat_inte_corr) <- rownames(mat_inte_df)

  # Add variables to mat_inte_df
  mat_inte_df$lpconc <- as.numeric(rna_metadata$p.LibraryConcentration_ngul)
  mat_inte_df$extractor <- as.factor(rna_metadata$ex.Extractedby)
  mat_inte_df$sbatch <- as.factor(rna_metadata$s.SequencingBatch)
  mat_inte_df$RINscore <- as.numeric(rna_metadata$ex.RINscore)
  mat_inte_df$sex <- as.factor(rna_metadata$i.Sex)
  mat_inte_df$age <- as.numeric(rna_metadata$i.Age)

  # Set up progress bar
  total_cols <- ncol(mat_inte_corr)
  pb <- txtProgressBar(min = 0, max = total_cols, style = 3)

  # Run linear models and store residuals
  set.seed(123)
  for(i in seq_along(colnames(mat_inte_corr))){
    col_name <- colnames(mat_inte_corr)[i]
    obj <- lm(as.formula(paste0("`", col_name, "`", "~ lpconc + sbatch + RINscore + extractor + sex + age")),  data = mat_inte_df)
    mat_inte_corr[, col_name] <- resid(obj)
    
    # Update progress bar
    setTxtProgressBar(pb, i)
  }

  # Close progress bar
  close(pb)
  # filter low variance genes
  mat_inte_corr_t <- t(mat_inte_corr)
  ## calculate variance for each row
  gene_variance <- apply(mat_inte_corr_t, 1, var)
  ## calculate the 25% quantile of variance
  variance_cutoff <- quantile(gene_variance, 0.25)
  ## filter to genes above the 25% quantile of variance
  mat_inte_filt <-  mat_inte_corr_t[gene_variance > variance_cutoff, ]
  # Convert the corrected matrix to a dataframe
  mat_inte_filt <- as.data.frame(t(mat_inte_filt))

  return(mat_inte_filt)
}


correct_microbiome_data_within_region <- function(ps) {
  # Agglomerate taxa at Genus level
  phy_genus_overlap <- tax_glom(ps, "Genus", NArm = FALSE)
  phy_genus_overlap_RA <- transform_sample_counts(phy_genus_overlap, function(otu) otu/sum(otu))
  phy_genus_overlap_RA <- filter_taxa(phy_genus_overlap_RA, function(x) sum(x >= 0.0001) >= (nsamples(ps)/10), TRUE)
  phy_genus_overlap_filt <- microbiome::transform(phy_genus_overlap_RA, transform = "clr")
  
  # Convert OTU table to dataframe
  otu_table_overlap <- as.data.frame(as.matrix(phy_genus_overlap_filt@otu_table))
  setDT(otu_table_overlap, keep.rownames = TRUE)[]
  colnames(otu_table_overlap)[1] <- "SampleID"
  
  # Process OTU table
  otu_table2_overlap <- otu_table_overlap %>% data.frame(check.names=FALSE) %>% set_rownames(.$SampleID)
  otu_table2_overlap <- otu_table2_overlap[,-1]
  otu_table2_overlap <- t(otu_table2_overlap)
  
  # Get taxonomy table
  tax_table_overlap <- as.data.frame(as.matrix(phy_genus_overlap_filt@tax_table))
  
  # Merge OTU and taxonomy tables
  otu_tax_overlap <- merge(otu_table2_overlap, tax_table_overlap, by = 0, all = TRUE)
  
  # Process taxonomy information
  otu_tax_overlap$Genus <- replace_na(otu_tax_overlap$Genus, "_unclassified")
  otu_tax_overlap$Phylum <- ifelse(is.na(otu_tax_overlap$Phylum), otu_tax_overlap$Kingdom, otu_tax_overlap$Phylum)
  otu_tax_overlap$Class <- ifelse(is.na(otu_tax_overlap$Class), otu_tax_overlap$Phylum, otu_tax_overlap$Class)
  otu_tax_overlap$Order <- ifelse(is.na(otu_tax_overlap$Order), otu_tax_overlap$Class, otu_tax_overlap$Order)
  otu_tax_overlap$Family <- ifelse(is.na(otu_tax_overlap$Family), otu_tax_overlap$Order, otu_tax_overlap$Family)
  otu_tax_overlap$Genus <- ifelse(otu_tax_overlap$Genus == "_unclassified", paste(otu_tax_overlap$Family, otu_tax_overlap$Genus, sep = ""), otu_tax_overlap$Genus)
  otu_tax_overlap$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax_overlap$Genus)
  otu_tax_overlap$Family <- ifelse(otu_tax_overlap$Family == "uncultured", paste0("uncultured_", otu_tax_overlap$Order), otu_tax_overlap$Family)
  otu_tax_overlap$Genus <- ifelse(grepl("uncultured", otu_tax_overlap$Family), otu_tax_overlap$Family, otu_tax_overlap$Genus)
  otu_tax_overlap$Genus <- ifelse(otu_tax_overlap$Genus == "uncultured", paste0("uncultured_", otu_tax_overlap$Family), otu_tax_overlap$Genus)
  
  # Set genus as rownames and remove taxonomic columns
  rownames(otu_tax_overlap) <- otu_tax_overlap$Genus
  otu_tax_overlap <- subset(otu_tax_overlap, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus, Species))
  otu_tax_overlap <- as.data.frame(t(otu_tax_overlap))
  
  # Fix names
  rownames(otu_tax_overlap) <- phy_genus_overlap@sam_data$UNQ_ID
  
  # Prepare for correlation
  overlap_otu_corr <- matrix(ncol = ncol(otu_tax_overlap), nrow = nrow(otu_tax_overlap))
  colnames(overlap_otu_corr) <- colnames(otu_tax_overlap)
  rownames(overlap_otu_corr) <- rownames(otu_tax_overlap)
  
  # Get sample data
  overlap_sam <- as.data.frame(as.matrix(phy_genus_overlap_filt@sam_data))
  
  # Add variables
  otu_tax_overlap$PCRconc <- as.numeric(overlap_sam$p.CleanedPCRConcentration)
  otu_tax_overlap$exbatch <- as.factor(overlap_sam$ex.ExtractionBatch)
  otu_tax_overlap$sex <- as.factor(overlap_sam$i.Sex)
  otu_tax_overlap$age <- as.numeric(overlap_sam$i.Age)
  
  # Run linear models with progress bar
  total_cols <- ncol(overlap_otu_corr)
  pb <- txtProgressBar(min = 0, max = total_cols, style = 3)
  
  set.seed(123)
  for(i in seq_along(colnames(overlap_otu_corr))){
    col_name <- colnames(overlap_otu_corr)[i]
    obj <- lm(as.formula(paste0("`", col_name, "`", "~ PCRconc + exbatch + sex + age")),  data = otu_tax_overlap)
    overlap_otu_corr[, col_name] <- resid(obj)
    setTxtProgressBar(pb, i)
  }
  
  close(pb)
  
  # Convert to dataframe and return
  overlap_otu_corr <- as.data.frame(overlap_otu_corr)
  return(overlap_otu_corr)
}
```

## filter samples and adjust order
```{r}
micro_overlap_ps <- subset_samples(micro_ps, UNQ_ID != "NA")
CD_overlap_ps <- subset_samples(micro_overlap_ps, i.Disease == "Crohns")

rnaseq_meta_overlap <- subset(rnaseq_meta, UNQ_ID != "NA")
rna_meta_CD <- subset(rnaseq_meta_overlap, i.Disease == "Crohns")

## adjust order of rna metadata based on microbiome
rna_meta_CD_reorder <- rna_meta_CD[order(match(rna_meta_CD$UNQ_ID, CD_overlap_ps@sam_data$UNQ_ID)),]

rna_count_CD <- subset(rnaseq_count, rownames(rnaseq_count) %in% rna_meta_CD_reorder$p.LibraryID)

## adjust order of rna count based on rna metadata
rna_count_CD_reorder <- rna_count_CD[order(match(rownames(rna_count_CD), rna_meta_CD_reorder$p.LibraryID)),]

rownames(rna_count_CD_reorder) <- rna_meta_CD_reorder$UNQ_ID

rna_count_CD_reorder_t <- t(rna_count_CD_reorder)

## check order
identical(CD_overlap_ps@sam_data$UNQ_ID, rownames(rna_count_CD_reorder))
```

## correct data for covariates, including specimen
```{r}
# RNAseq
## checks
identical(rownames(rna_count_CD_reorder), rna_meta_CD_reorder$UNQ_ID)

## normalize
set.seed(123)
dds <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ 1)
smallestGroupSize <- ncol(rna_meta_CD)/3
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

set.seed(123)
vsd <- vst(dds, blind=FALSE)
mat_inte <- assay(vsd)

## Convert mat_inte to a dataframe and transpose
mat_inte_df <- as.data.frame(t(mat_inte))

## filter to protein coding genes only
mat_inte_df <- mat_inte_df[, colnames(mat_inte_df) %in% anno_protein_coding$symbol]

## Create an empty matrix for corrected data
mat_inte_corr <- matrix(ncol = ncol(mat_inte_df), nrow = nrow(mat_inte_df))
colnames(mat_inte_corr) <- colnames(mat_inte_df)
rownames(mat_inte_corr) <- rownames(mat_inte_df)

## Add covariates
mat_inte_df$lpconc <- as.numeric(rna_meta_CD_reorder$p.LibraryConcentration_ngul)
mat_inte_df$extractor <- as.factor(rna_meta_CD_reorder$ex.Extractedby)
mat_inte_df$sbatch <- as.factor(rna_meta_CD_reorder$s.SequencingBatch)
mat_inte_df$RINscore <- as.numeric(rna_meta_CD_reorder$ex.RINscore)
mat_inte_df$sex <- as.factor(rna_meta_CD_reorder$i.Sex)
mat_inte_df$age <- as.numeric(rna_meta_CD_reorder$i.Age)
mat_inte_df$spec <- as.factor(rna_meta_CD_reorder$i.Specimen)
mat_inte_df$id <- as.factor(rna_meta_CD_reorder$i.PatientID)

## Run linear models and store residuals
set.seed(123)
for(i in colnames(mat_inte_corr)){
  
  obj <- lmer(as.formula(paste0("`", i, "`", "~ lpconc + sbatch + RINscore + extractor +
                                sex + age + spec + (1|id)")),  data = mat_inte_df, 
              control = lmerControl(optCtrl = list(maxfun = 1000000), 
                                    check.conv.singular = .makeCC(action = "ignore", tol = formals(isSingular)$tol)))
  mat_inte_corr[,i] <- resid(obj)
}

## filter low variance genes
mat_inte_corr_t <- t(mat_inte_corr)
## calculate variance for each row
gene_variance <- apply(mat_inte_corr_t, 1, var)
## calculate the 25% quantile of variance
variance_cutoff <- quantile(gene_variance, 0.25)
## filter to genes above the 25% quantile of variance
mat_inte_filt <-  mat_inte_corr_t[gene_variance > variance_cutoff, ]
# Convert the corrected matrix to a dataframe
rna_corr_ALL <- as.data.frame(t(mat_inte_filt))


# microbiome
## agglomerate to Genus, transform to RA, and then CLR transform
CD_genus_overlap <- tax_glom(CD_overlap_ps, "Genus", NArm = FALSE)
CD_genus_overlap_RA <- transform_sample_counts(CD_genus_overlap, function(otu) otu/sum(otu))
CD_genus_overlap_RA <- filter_taxa(CD_genus_overlap_RA, function(x) sum(x >= 0.0001) >= (nsamples(CD_overlap_ps)/10), TRUE)
CD_genus_overlap_filt <- microbiome::transform(CD_genus_overlap_RA, transform = "clr")
  
## Convert OTU table to dataframe
CD_otu_table_overlap <- as.data.frame(as.matrix(CD_genus_overlap_filt@otu_table))
setDT(CD_otu_table_overlap, keep.rownames = TRUE)[]
colnames(CD_otu_table_overlap)[1] <- "SampleID"
  
## fix formatting
CD_otu_table2_overlap <- CD_otu_table_overlap %>% data.frame(check.names=FALSE) %>% set_rownames(.$SampleID)
CD_otu_table2_overlap <- CD_otu_table2_overlap[,-1]
CD_otu_table2_overlap <- t(CD_otu_table2_overlap)
  
## Get taxonomy table
CD_tax_table_overlap <- as.data.frame(as.matrix(CD_genus_overlap_filt@tax_table))
  
## Merge OTU and taxonomy tables
CD_otu_tax_overlap <- merge(CD_otu_table2_overlap, CD_tax_table_overlap, by = 0, all = TRUE)
  
## fix taxonomy information
CD_otu_tax_overlap$Genus <- replace_na(CD_otu_tax_overlap$Genus, "_unclassified")
CD_otu_tax_overlap$Phylum <- ifelse(is.na(CD_otu_tax_overlap$Phylum), CD_otu_tax_overlap$Kingdom, CD_otu_tax_overlap$Phylum)
CD_otu_tax_overlap$Class <- ifelse(is.na(CD_otu_tax_overlap$Class), CD_otu_tax_overlap$Phylum, CD_otu_tax_overlap$Class)
CD_otu_tax_overlap$Order <- ifelse(is.na(CD_otu_tax_overlap$Order), CD_otu_tax_overlap$Class, CD_otu_tax_overlap$Order)
CD_otu_tax_overlap$Family <- ifelse(is.na(CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Order, CD_otu_tax_overlap$Family)
CD_otu_tax_overlap$Genus <- ifelse(CD_otu_tax_overlap$Genus == "_unclassified", paste(CD_otu_tax_overlap$Family, CD_otu_tax_overlap$Genus, sep = ""), CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Genus <- gsub("NA_unclassified", "Unclassified", CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Family <- ifelse(CD_otu_tax_overlap$Family == "uncultured", paste0("uncultured_", CD_otu_tax_overlap$Order), CD_otu_tax_overlap$Family)
CD_otu_tax_overlap$Genus <- ifelse(grepl("uncultured", CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Family, CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Genus <- ifelse(CD_otu_tax_overlap$Genus == "uncultured", paste0("uncultured_", CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Genus)
  
## Set genus as rownames and remove taxonomic columns
rownames(CD_otu_tax_overlap) <- CD_otu_tax_overlap$Genus
CD_otu_tax_overlap <- subset(CD_otu_tax_overlap, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus, Species))
CD_otu_tax_overlap <- as.data.frame(t(CD_otu_tax_overlap))
  
## Fix names
rownames(CD_otu_tax_overlap) <- CD_genus_overlap_filt@sam_data$UNQ_ID
  
## Prepare for correlation
CD_overlap_otu_corr <- matrix(ncol = ncol(CD_otu_tax_overlap), nrow = nrow(CD_otu_tax_overlap))
colnames(CD_overlap_otu_corr) <- colnames(CD_otu_tax_overlap)
rownames(CD_overlap_otu_corr) <- rownames(CD_otu_tax_overlap)
  
## Get sample data
CD_overlap_sam <- as.data.frame(as.matrix(CD_genus_overlap_filt@sam_data))
  
## Add variables
CD_otu_tax_overlap$PCRconc <- as.numeric(CD_overlap_sam$p.CleanedPCRConcentration)
CD_otu_tax_overlap$exbatch <- as.factor(CD_overlap_sam$ex.ExtractionBatch)
CD_otu_tax_overlap$sex <- as.factor(CD_overlap_sam$i.Sex)
CD_otu_tax_overlap$age <- as.numeric(CD_overlap_sam$i.Age)
CD_otu_tax_overlap$id <- as.factor(CD_overlap_sam$i.PatientID)
CD_otu_tax_overlap$spec <- as.factor(CD_overlap_sam$i.Specimen)
  
## Run linear models
set.seed(123)
for(i in colnames(CD_overlap_otu_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ PCRconc + exbatch + sex + age + spec + (1|id)")),  
              data = CD_otu_tax_overlap, control = lmerControl(optCtrl = list(maxfun = 1000000), 
                                    check.conv.singular = .makeCC(action = "ignore", tol =
                                                                    formals(isSingular)$tol)))
  CD_overlap_otu_corr[,i] <- resid(obj)
}

## Convert to dataframe
CD_overlap_otu_corr <- as.data.frame(CD_overlap_otu_corr)

## checks
identical(rownames(CD_overlap_otu_corr), rownames(rna_corr_ALL))
```

## format data in prep for sparse-CCA
```{r}
rna_corr_mat <- as.matrix(rna_corr_ALL)
micro_corr_mat <- as.matrix(CD_overlap_otu_corr)
```

## run sparse CCA
```{r}
current_dir <- dirname(rstudioapi::getSourceEditorContext()$path)

##### what is the reasoning for this value?
cca.k = 10

set.seed(123)
bestPenalty <- tune_params_grid_search(rna_corr_mat, micro_corr_mat)
bestpenaltyX <- bestPenalty[1]
bestpenaltyY <- bestPenalty[2]

set.seed(123)
cca <- run_sparseCCA(rna_corr_mat, micro_corr_mat, cca.k, bestpenaltyX, 
                            bestpenaltyY, outputFile = paste0(current_dir, "/../output/CCA_output_",
                                                                     bestpenaltyX, "_" , 
                                                                     bestpenaltyY, ".txt"))

## average number of genes and microbes in resulting components
avg_genes <- get_avg_features(cca[[1]]$u, cca.k)
avg_genes

avg.microbes <- get_avg_features(cca[[1]]$v, cca.k)
avg.microbes


## check significance
CCA_pval <- test_significance_LOOCV(rna_corr_mat, micro_corr_mat, bestpenaltyX, bestpenaltyY, cca.k)

length(which(CCA_pval < 0.1))
which(CCA_pval < 0.1)

CCA_padj <- p.adjust(CCA_pval, method = "BH")
CCA_padj

length(which(CCA_padj < 0.1))
which(CCA_padj < 0.1)

## save values
CCA_sig <- cbind(CCA_pval, CCA_padj)

rownames(CCA_sig) <- paste0("comp_", 1:10)

# write.csv(CCA_sig, "../output/CCA_sig.csv")
```

## plot components
```{r}
rna_comp <- as.data.frame(cca[[2]])
micro_comp <- as.data.frame(cca[[3]])

colnames(rna_comp) <- sprintf("Comp%02d", 1:10)
rna_comp$sample <- rownames(rna_comp)

rna_comp <- pivot_longer(rna_comp, cols = !sample, names_to = "component", values_to = "rna")

colnames(micro_comp) <- sprintf("Comp%02d", 1:10)
micro_comp$sample <- rownames(micro_comp)

micro_comp <- pivot_longer(micro_comp, cols = !sample, names_to = "component", values_to = "micro")

df_comp <- cbind(rna_comp, micro_comp$micro)
df_comp <- as.data.frame(df_comp)
colnames(df_comp) <- c("sample", "component", "rna", "microbe")


sparseCCA_intestine <- ggplot(df_comp, aes(x = rna, y = microbe), color = black) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  facet_wrap(~component, scales = "free", , ncol = 5) +
  labs(x = "RNA canonical variate", y = "Microbe canonical variate")

ggsave("../figures/sparseCCA_intestine.pdf", sparseCCA_intestine, width = 20, height = 7)
```

## save components
```{r}
sig_cutoff <- 0.1
sig <- which(CCA_pval < sig_cutoff)
dirname <- paste0(current_dir,"/../output/ALL_REGION")
save_CCA_components(cca[[1]],sig,dirname)
getwd()

saveRDS(cca, file = "../output/ALL_REGION_cca.RDS")
```

## extract significant components and switch to entrez IDs
```{r}
comp1 <- cca[[1]][["u"]][,1]
comp1 <- as.data.frame(comp1)
comp1 <- filter(comp1, comp1 != 0)
comp1 <- rownames(comp1)

comp2 <- cca[[1]][["u"]][,2]
comp2 <- as.data.frame(comp2)
comp2 <- filter(comp2, comp2 != 0)
comp2 <- rownames(comp2)

comp3 <- cca[[1]][["u"]][,3]
comp3 <- as.data.frame(comp3)
comp3 <- filter(comp3, comp3 != 0)
comp3 <- rownames(comp3)

comp4 <- cca[[1]][["u"]][,4]
comp4 <- as.data.frame(comp4)
comp4 <- filter(comp4, comp4 != 0)
comp4 <- rownames(comp4)

comp6 <- cca[[1]][["u"]][,6]
comp6 <- as.data.frame(comp6)
comp6 <- filter(comp6, comp6 != 0)
comp6 <- rownames(comp6)

comp10 <- cca[[1]][["u"]][,10]
comp10 <- as.data.frame(comp10)
comp10 <- filter(comp10, comp10 != 0)
comp10 <- rownames(comp10)

# universe
all_entrez <- colnames(rna_corr_mat)

# entrez ID
comp1_entrez <- as.character(anno_protein_coding$entrezID[match(comp1, anno_protein_coding$symbol)])
comp2_entrez <- as.character(anno_protein_coding$entrezID[match(comp2, anno_protein_coding$symbol)])
comp3_entrez <- as.character(anno_protein_coding$entrezID[match(comp3, anno_protein_coding$symbol)])
comp4_entrez <- as.character(anno_protein_coding$entrezID[match(comp4, anno_protein_coding$symbol)])
comp6_entrez <- as.character(anno_protein_coding$entrezID[match(comp6, anno_protein_coding$symbol)])
comp10_entrez <- as.character(anno_protein_coding$entrezID[match(comp10, anno_protein_coding$symbol)])

all_entrez <- as.character(anno_protein_coding$entrezID[match(all_entrez, anno_protein_coding$symbol)])
```

## combine all pathway info
```{r}
colnames(hsa_to_path) <- c("pathway", "entrez_ID") 
colnames(PID_path) <- c("pathway", "entrez_ID")

path_input <- bind_rows(hsa_to_path, PID_path)
```

## enrichment
```{r}
comp1_enrich <- enricher(comp1_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp2_enrich <- enricher(comp2_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp3_enrich <- enricher(comp3_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp4_enrich <- enricher(comp4_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp6_enrich <- enricher(comp6_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp10_enrich <- enricher(comp10_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp1_enrich_out <- comp1_enrich@result
comp2_enrich_out <- comp2_enrich@result
comp3_enrich_out <- comp3_enrich@result
comp4_enrich_out <- comp4_enrich@result
comp6_enrich_out <- comp6_enrich@result
comp10_enrich_out <- comp10_enrich@result

# filter to significant pathways
comp1_enrich_filt <- filter(comp1_enrich_out, p.adjust < 0.1 & Count >= 10)
comp2_enrich_filt <- filter(comp2_enrich_out, p.adjust < 0.1 & Count >= 10)
comp3_enrich_filt <- filter(comp3_enrich_out, p.adjust < 0.1 & Count >= 10)
comp4_enrich_filt <- filter(comp4_enrich_out, p.adjust < 0.1 & Count >= 10)
comp6_enrich_filt <- filter(comp6_enrich_out, p.adjust < 0.1 & Count >= 10)
comp10_enrich_filt <- filter(comp10_enrich_out, p.adjust < 0.1 & Count >= 10)
```

## plot enrichments
```{r}
barplot(comp1_enrich, showCategory=20) 
barplot(comp2_enrich, showCategory=20)
barplot(comp3_enrich, showCategory=20)
barplot(comp4_enrich, showCategory=20)
```

## examine KEGG pathways
```{r}
comp1_wkegg_path <- comp1_enrich_filt
comp1_wkegg_path$Description <- pathway_list$description[match(comp1_wkegg_path$Description, pathway_list$map)]

comp3_wkegg_path <- comp3_enrich_filt
comp3_wkegg_path$Description <- pathway_list$description[match(comp3_wkegg_path$Description, pathway_list$map)]

```

## save
```{r}
# ## save
# write.csv(comp1_wkegg_path, "../output/comp1_wkegg_path.csv")
# write.csv(comp3_wkegg_path, "../output/comp3_wkegg_path.csv")
```

## filter data by region
```{r}
rna_meta_TI <- subset(rna_meta_CD_reorder, i.Specimen == "Terminal Ileum")
rna_meta_C <- subset(rna_meta_CD_reorder, i.Specimen == "Cecum")
rna_meta_RC <- subset(rna_meta_CD_reorder, i.Specimen == "Right Colon")

micro_TI_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Terminal Ileum")
micro_C_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Cecum")
micro_RC_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Right Colon")
```

## transform RNAseq data
```{r}
set.seed(123)
dds_full_spec <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ i.Specimen)

smallestGroupSize_spec <- 30
keep_spec <- rowSums(counts(dds_full_spec) >= 10) >= smallestGroupSize_spec
dds_full_spec <- dds_full_spec[keep_spec,]
set.seed(123)
vsd_spec <- vst(dds_full_spec, blind=FALSE)
mat_spec <- assay(vsd_spec)

mat_spec <- as.data.frame(t(mat_spec))

mat_TI <- mat_spec[rownames(mat_spec) %in% rna_meta_TI$UNQ_ID,]
mat_C <- mat_spec[rownames(mat_spec) %in% rna_meta_C$UNQ_ID,]
mat_RC <- mat_spec[rownames(mat_spec) %in% rna_meta_RC$UNQ_ID,]
```

## correct regional datasets
```{r}
# checks
identical(rownames(mat_TI), rna_meta_TI$UNQ_ID)
identical(rownames(mat_C), rna_meta_C$UNQ_ID)
identical(rownames(mat_RC), rna_meta_RC$UNQ_ID)

# correct data
set.seed(123)
rna_corr_TI <- correct_rnaseq_data_within_region(mat_TI, rna_meta_TI)
set.seed(123)
rna_corr_C <- correct_rnaseq_data_within_region(mat_C, rna_meta_C)
set.seed(123)
rna_corr_RC <- correct_rnaseq_data_within_region(mat_RC, rna_meta_RC)
set.seed(123)
micro_corr_TI <- correct_microbiome_data_within_region(micro_TI_ps)
set.seed(123)
micro_corr_C <- correct_microbiome_data_within_region(micro_C_ps)
set.seed(123)
micro_corr_RC <- correct_microbiome_data_within_region(micro_RC_ps)

# format for sparse-CCA
rna_corr_TI_mat <- as.matrix(rna_corr_TI)
rna_corr_C_mat <- as.matrix(rna_corr_C)
rna_corr_RC_mat <- as.matrix(rna_corr_RC)
micro_corr_TI_mat <- as.matrix(micro_corr_TI)
micro_corr_C_mat <- as.matrix(micro_corr_C)
micro_corr_RC_mat <- as.matrix(micro_corr_RC)

# check
identical(rownames(rna_corr_TI_mat), rownames(micro_corr_TI_mat))
identical(rownames(rna_corr_C_mat), rownames(micro_corr_C_mat))
identical(rownames(rna_corr_RC_mat), rownames(micro_corr_RC_mat))
```

## run sparse CCA
```{r}
current_dir <- dirname(rstudioapi::getSourceEditorContext()$path)

cca.k = 10
# CLR
## TI
set.seed(100)
bestPenalty_TI <- tune_params_grid_search(rna_corr_TI_mat, micro_corr_TI_mat)
bestpenaltyX_TI <- bestPenalty_TI[1]
bestpenaltyY_TI <- bestPenalty_TI[2]

bestpenaltyX_TI
bestpenaltyY_TI

set.seed(100)
cca_TI <- run_sparseCCA(rna_corr_TI_mat, micro_corr_TI_mat, cca.k, bestpenaltyX_TI, 
                            bestpenaltyY_TI, outputFile = paste0(current_dir,
                                                                     "/../output/CCA_TI_output_",
                                                                     bestpenaltyX_TI, "_" , 
                                                                     bestpenaltyY_TI, ".txt"))

## average number of genes and microbes in resulting components
avg_genes_TI <- get_avg_features(cca_TI[[1]]$u, cca.k)
avg_genes_TI

avg.microbes_TI <- get_avg_features(cca_TI[[1]]$v, cca.k)
avg.microbes_TI


## C 
set.seed(100)
bestPenalty_C <- tune_params_grid_search(rna_corr_C_mat, micro_corr_C_mat)
bestpenaltyX_C <- bestPenalty_C[1]
bestpenaltyY_C <- bestPenalty_C[2]

set.seed(100)
cca_C <- run_sparseCCA(rna_corr_C_mat, micro_corr_C_mat, cca.k, bestpenaltyX_C, 
                            bestpenaltyY_C, outputFile = paste0(current_dir,
                                                                    "/../output/CCA_C_output_",
                                                                     bestpenaltyX_C, "_" , 
                                                                     bestpenaltyY_C, ".txt"))

## average number of genes and microbes in resulting components
avg_genes_C <- get_avg_features(cca_C[[1]]$u, cca.k)
avg_genes_C

avg.microbes_C <- get_avg_features(cca_C[[1]]$v, cca.k)
avg.microbes_C


## RC
set.seed(100)
bestPenalty_RC <- tune_params_grid_search(rna_corr_RC_mat, micro_corr_RC_mat)
bestpenaltyX_RC <- bestPenalty_RC[1]
bestpenaltyY_RC <- bestPenalty_RC[2]

set.seed(100)
cca_RC <- run_sparseCCA(rna_corr_RC_mat, micro_corr_RC_mat, cca.k, bestpenaltyX_RC, 
                            bestpenaltyY_RC, outputFile = paste0(current_dir,
                                                                     "/../output/CCA_RC_output_",
                                                                     bestpenaltyX_RC, "_" , 
                                                                     bestpenaltyY_RC, ".txt"))

## average number of genes and microbes in resulting components
avg_genes_RC <- get_avg_features(cca_RC[[1]]$u, cca.k)
avg_genes_RC

avg.microbes_RC <- get_avg_features(cca_RC[[1]]$v, cca.k)
avg.microbes_RC
```

## check significance
```{r}
## TI
CCA_pval_TI <- test_significance_LOOCV(rna_corr_TI_mat, micro_corr_TI_mat, bestpenaltyX_TI, bestpenaltyY_TI, cca.k)

length(which(CCA_pval_TI < 0.1))
which(CCA_pval_TI < 0.1)

CCA_padj_TI <- p.adjust(CCA_pval_TI, method = "BH")
CCA_padj_TI

length(which(CCA_padj_TI < 0.1))
which(CCA_padj_TI < 0.1)

## C
CCA_pval_C <- test_significance_LOOCV(rna_corr_C_mat, micro_corr_C_mat, bestpenaltyX_C, bestpenaltyY_C, cca.k)

length(which(CCA_pval_C < 0.1))
which(CCA_pval_C < 0.1)

CCA_padj_C <- p.adjust(CCA_pval_C, method = "BH")
CCA_padj_C

length(which(CCA_padj_C < 0.1))
which(CCA_padj_C < 0.1)

## RC
CCA_pval_RC <- test_significance_LOOCV(rna_corr_RC_mat, micro_corr_RC_mat, bestpenaltyX_RC, bestpenaltyY_RC, cca.k)

length(which(CCA_pval_RC < 0.1))
which(CCA_pval_RC < 0.1)

CCA_padj_RC <- p.adjust(CCA_pval_RC, method = "BH")
CCA_padj_RC

length(which(CCA_padj_RC < 0.1))
which(CCA_padj_RC < 0.1)

## save values
CCA_sig_TI <- cbind(CCA_pval_TI, CCA_padj_TI)
rownames(CCA_sig_TI) <- paste0("comp_", 1:10)

CCA_sig_C <- cbind(CCA_pval_C, CCA_padj_C)
rownames(CCA_sig_C) <- paste0("comp_", 1:10)

CCA_sig_RC <- cbind(CCA_pval_RC, CCA_padj_RC)
rownames(CCA_sig_RC) <- paste0("comp_", 1:10)

# write.csv(CCA_sig_TI, "../output/CCA_sig_TI.csv")
# write.csv(CCA_sig_C, "../output/CCA_sig_C.csv")
# write.csv(CCA_sig_RC, "../output/CCA_sig_RC.csv")
```

## save components
```{r}
sig_cutoff <- 0.1
sig_TI <- which(CCA_pval_TI < sig_cutoff)
dirname_TI <- paste0(current_dir,"/../output/TI_")
save_CCA_components(cca_TI[[1]],sig_TI,dirname_TI)

sig_C <- which(CCA_pval_C < sig_cutoff)
dirname_C <- paste0(current_dir,"/../output/C_")
save_CCA_components(cca_C[[1]],sig_C,dirname_C)

sig_RC <- which(CCA_pval_RC < sig_cutoff)
dirname_RC <- paste0(current_dir,"/../output/RC_")
save_CCA_components(cca_RC[[1]],sig_RC,dirname_RC)
```

## plot components
```{r}
# TI
rna_TI_comp <- as.data.frame(cca_TI[[2]])
micro_TI_comp <- as.data.frame(cca_TI[[3]])

colnames(rna_TI_comp) <- sprintf("Comp%02d", 1:10)
rna_TI_comp$sample <- rownames(rna_TI_comp)

rna_TI_comp <- pivot_longer(rna_TI_comp, cols = !sample, names_to = "component", values_to = "rna")

colnames(micro_TI_comp) <- sprintf("Comp%02d", 1:10)
micro_TI_comp$sample <- rownames(micro_TI_comp)

micro_TI_comp <- pivot_longer(micro_TI_comp, cols = !sample, names_to = "component", values_to = "micro")

df_TI_comp <- cbind(rna_TI_comp, micro_TI_comp$micro)
df_TI_comp <- as.data.frame(df_TI_comp)
colnames(df_TI_comp) <- c("sample", "component", "rna", "microbe")

sparseCCA_TI <- ggplot(df_TI_comp, aes(x = rna, y = microbe), color = black) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  facet_wrap(~component, scales = "free", , ncol = 2) +
  labs(x = "RNA canonical variate", y = "Microbe canonical variate")

ggsave("../figures/sparseCCA_TI.pdf", sparseCCA_TI, width = 2.75, height = 8)


# C
rna_C_comp <- as.data.frame(cca_C[[2]])
micro_C_comp <- as.data.frame(cca_C[[3]])

colnames(rna_C_comp) <- sprintf("Comp%02d", 1:10)
rna_C_comp$sample <- rownames(rna_C_comp)

rna_C_comp <- pivot_longer(rna_C_comp, cols = !sample, names_to = "component", values_to = "rna")

colnames(micro_C_comp) <- sprintf("Comp%02d", 1:10)
micro_C_comp$sample <- rownames(micro_C_comp)

micro_C_comp <- pivot_longer(micro_C_comp, cols = !sample, names_to = "component", values_to = "micro")

df_C_comp <- cbind(rna_C_comp, micro_C_comp$micro)
df_C_comp <- as.data.frame(df_C_comp)
colnames(df_C_comp) <- c("sample", "component", "rna", "microbe")

sparseCCA_C <- ggplot(df_C_comp, aes(x = rna, y = microbe), color = black) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  facet_wrap(~component, scales = "free", , ncol = 2) +
  labs(x = "RNA canonical variate", y = "Microbe canonical variate")

ggsave("../figures/sparseCCA_C.pdf", sparseCCA_C, width = 2.75, height = 8)


# RC
rna_RC_comp <- as.data.frame(cca_RC[[2]])
micro_RC_comp <- as.data.frame(cca_RC[[3]])

colnames(rna_RC_comp) <- sprintf("Comp%02d", 1:10)
rna_RC_comp$sample <- rownames(rna_RC_comp)

rna_RC_comp <- pivot_longer(rna_RC_comp, cols = !sample, names_to = "component", values_to = "rna")

colnames(micro_RC_comp) <- sprintf("Comp%02d", 1:10)
micro_RC_comp$sample <- rownames(micro_RC_comp)

micro_RC_comp <- pivot_longer(micro_RC_comp, cols = !sample, names_to = "component", values_to = "micro")

df_RC_comp <- cbind(rna_RC_comp, micro_RC_comp$micro)
df_RC_comp <- as.data.frame(df_RC_comp)
colnames(df_RC_comp) <- c("sample", "component", "rna", "microbe")

sparseCCA_RC <- ggplot(df_RC_comp, aes(x = rna, y = microbe), color = black) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  facet_wrap(~component, scales = "free", , ncol = 2) +
  labs(x = "RNA canonical variate", y = "Microbe canonical variate")

ggsave("../figures/sparseCCA_RC.pdf", sparseCCA_RC, width = 2.75, height = 8)
```

## extract significant components and switch to entrez IDs
```{r}
comp1_TI <- cca_TI[[1]][["u"]][,1]
comp1_TI <- as.data.frame(comp1_TI)
comp1_TI <- filter(comp1_TI, comp1_TI != 0)
comp1_TI <- rownames(comp1_TI)

comp2_TI <- cca_TI[[1]][["u"]][,2]
comp2_TI <- as.data.frame(comp2_TI)
comp2_TI <- filter(comp2_TI, comp2_TI != 0)
comp2_TI <- rownames(comp2_TI)

comp3_TI <- cca_TI[[1]][["u"]][,3]
comp3_TI <- as.data.frame(comp3_TI)
comp3_TI <- filter(comp3_TI, comp3_TI != 0)
comp3_TI <- rownames(comp3_TI)

comp9_TI <- cca_TI[[1]][["u"]][,9]
comp9_TI <- as.data.frame(comp9_TI)
comp9_TI <- filter(comp9_TI, comp9_TI != 0)
comp9_TI <- rownames(comp9_TI)

comp1_C <- cca_C[[1]][["u"]][,1]
comp1_C <- as.data.frame(comp1_C)
comp1_C <- filter(comp1_C, comp1_C != 0)
comp1_C <- rownames(comp1_C)

comp6_C <- cca_C[[1]][["u"]][,6]
comp6_C <- as.data.frame(comp6_C)
comp6_C <- filter(comp6_C, comp6_C != 0)
comp6_C <- rownames(comp6_C)

# universe
all_TI_entrez <- colnames(rna_corr_TI_mat)
all_C_entrez <- colnames(rna_corr_C_mat)
all_RC_entrez <- colnames(rna_corr_RC_mat)

# entrez ID
comp1_TI_entrez <- as.character(anno_protein_coding$entrezID[match(comp1_TI, anno_protein_coding$symbol)])
comp2_TI_entrez <- as.character(anno_protein_coding$entrezID[match(comp2_TI, anno_protein_coding$symbol)])
comp3_TI_entrez <- as.character(anno_protein_coding$entrezID[match(comp3_TI, anno_protein_coding$symbol)])
comp9_TI_entrez <- as.character(anno_protein_coding$entrezID[match(comp9_TI, anno_protein_coding$symbol)])
comp1_C_entrez <- as.character(anno_protein_coding$entrezID[match(comp1_C, anno_protein_coding$symbol)])
comp6_C_entrez <- as.character(anno_protein_coding$entrezID[match(comp6_C, anno_protein_coding$symbol)])

all_TI_entrez <- as.character(anno_protein_coding$entrezID[match(all_TI_entrez, anno_protein_coding$symbol)])
all_C_entrez <- as.character(anno_protein_coding$entrezID[match(all_C_entrez, anno_protein_coding$symbol)])
```

## enrichment
```{r}
comp1_TI_enrich <- enricher(comp1_TI_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_TI_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp2_TI_enrich <- enricher(comp2_TI_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_TI_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp3_TI_enrich <- enricher(comp3_TI_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_TI_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp9_TI_enrich <- enricher(comp9_TI_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_TI_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp1_C_enrich <- enricher(comp1_C_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_C_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp6_C_enrich <- enricher(comp6_C_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_C_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

comp1_TI_enrich_out <- comp1_TI_enrich@result
comp2_TI_enrich_out <- comp2_TI_enrich@result
comp3_TI_enrich_out <- comp3_TI_enrich@result
comp9_TI_enrich_out <- comp9_TI_enrich@result
comp1_C_enrich_out <- comp1_C_enrich@result
comp6_C_enrich_out <- comp6_C_enrich@result

# filter to significant pathways
comp1_TI_enrich_filt <- filter(comp1_TI_enrich_out, p.adjust < 0.1 & Count >= 10)
comp2_TI_enrich_filt <- filter(comp2_TI_enrich_out, p.adjust < 0.1 & Count >= 10)
comp3_TI_enrich_filt <- filter(comp3_TI_enrich_out, p.adjust < 0.1 & Count >= 10)
comp9_TI_enrich_filt <- filter(comp9_TI_enrich_out, p.adjust < 0.1 & Count >= 10)
comp1_C_enrich_filt <- filter(comp1_C_enrich_out, p.adjust < 0.1 & Count >= 10)
comp6_C_enrich_filt <- filter(comp6_C_enrich_out, p.adjust < 0.1 & Count >= 10)

## save
# write.csv(comp1_TI_enrich_filt, "../output/comp1_TI_enrich_filt.csv")
# write.csv(comp2_TI_enrich_filt, "../output/comp2_TI_enrich_filt.csv")
# write.csv(comp3_TI_enrich_filt, "../output/comp3_TI_enrich_filt.csv")
# write.csv(comp9_TI_enrich_filt, "../output/comp9_TI_enrich_filt.csv")
# write.csv(comp1_C_enrich_filt, "../output/comp1_C_enrich_filt.csv")
# write.csv(comp6_C_enrich_filt, "../output/comp6_C_enrich_filt.csv")
```

## plot
```{r}
barplot(comp1_TI_enrich, showCategory=20)
barplot(comp2_TI_enrich, showCategory=20)
barplot(comp3_TI_enrich, showCategory=20)
barplot(comp9_TI_enrich, showCategory=20)
barplot(comp1_C_enrich, showCategory=20) 
barplot(comp6_C_enrich, showCategory=20) 
```

## examine KEGG pathways
```{r}
comp2_TI_wkegg_path <- comp2_TI_enrich_filt
comp2_TI_wkegg_path$Description <- pathway_list$description[match(comp2_TI_wkegg_path$Description, pathway_list$map)]

comp9_TI_wkegg_path <- comp9_TI_enrich_filt
comp9_TI_wkegg_path$Description <- pathway_list$description[match(comp9_TI_wkegg_path$Description, pathway_list$map)]
```

## save
```{r}
# ## save
# write.csv(comp2_TI_wkegg_path, "../output/comp2_TI_wkegg_path.csv")
# write.csv(comp9_TI_wkegg_path, "../output/comp9_TI_wkegg_path.csv")
```