---
title: "3.ind_association"
output: html_document
date: "2025-12-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3.1 Individual Association Data Prep

The purpose of this script is to prep data for identifying individual associations between genes and microbes

## load data
```{r}
rnaseq_count <- read.csv("../../rnaseq/2.rnaseq_analysis/output/counts_uncorr_fin_woOutlier.csv", header = TRUE, row.names = 1)
rnaseq_meta <- read.csv("../../rnaseq/2.rnaseq_analysis/output/metadata_fin_woOutlier.csv", header = TRUE, row.names = 1)
micro_ps <- readRDS("../../microbiome/3.microbiome_analysis/output/phyloseq_sampleReads_unadj.rds")

# protein coding genes
anno_protein_coding <- read.table("../data/protein_coding.txt", header = TRUE)
hsa_to_path <- read.csv("../data/hsa_to_path.csv", row.names = 1, header = TRUE)
pathway_list <- read.csv("../data/pathway_list.csv", header = TRUE)
PID_path <- read.csv("../data/PID_path_entrez.csv", header = FALSE)
```

## load packages
```{r}
library(phyloseq)
library(DESeq2)
library(lme4)
library(data.table)
library(dplyr)
library(tidyr)
library(magrittr)
```

## functions for corrections within region
```{r}
correct_rnaseq_data_within_region <- function(rna_transformed_region, rna_metadata) {
  # filter to protein-coding genes
  mat_inte_df <- rna_transformed_region[, colnames(rna_transformed_region) %in% anno_protein_coding$symbol]

  # Create an empty matrix for corrected data
  mat_inte_corr <- matrix(ncol = ncol(mat_inte_df), nrow = nrow(mat_inte_df))
  colnames(mat_inte_corr) <- colnames(mat_inte_df)
  rownames(mat_inte_corr) <- rownames(mat_inte_df)

  # Add variables to mat_inte_df
  mat_inte_df$lpconc <- as.numeric(rna_metadata$p.LibraryConcentration_ngul)
  mat_inte_df$extractor <- as.factor(rna_metadata$ex.Extractedby)
  mat_inte_df$sbatch <- as.factor(rna_metadata$s.SequencingBatch)
  mat_inte_df$RINscore <- as.numeric(rna_metadata$ex.RINscore)
  mat_inte_df$sex <- as.factor(rna_metadata$i.Sex)
  mat_inte_df$age <- as.numeric(rna_metadata$i.Age)

  # Set up progress bar
  total_cols <- ncol(mat_inte_corr)
  pb <- txtProgressBar(min = 0, max = total_cols, style = 3)

  # Run linear models and store residuals
  set.seed(123)
  for(i in seq_along(colnames(mat_inte_corr))){
    col_name <- colnames(mat_inte_corr)[i]
    obj <- lm(as.formula(paste0("`", col_name, "`", "~ lpconc + sbatch + RINscore + extractor + sex + age")),  data = mat_inte_df)
    mat_inte_corr[, col_name] <- resid(obj)
    
    # Update progress bar
    setTxtProgressBar(pb, i)
  }

  # Close progress bar
  close(pb)
  # filter low variance genes
  mat_inte_corr_t <- t(mat_inte_corr)
  ## calculate variance for each row
  gene_variance <- apply(mat_inte_corr_t, 1, var)
  ## calculate the 25% quantile of variance
  variance_cutoff <- quantile(gene_variance, 0.25)
  ## filter to genes above the 25% quantile of variance
  mat_inte_filt <-  mat_inte_corr_t[gene_variance > variance_cutoff, ]
  # Convert the corrected matrix to a dataframe
  mat_inte_filt <- as.data.frame(t(mat_inte_filt))

  return(mat_inte_filt)
}


correct_microbiome_data_within_region <- function(ps) {
  # Agglomerate taxa at Genus level
  phy_genus_overlap <- tax_glom(ps, "Genus", NArm = FALSE)
  phy_genus_overlap_RA <- transform_sample_counts(phy_genus_overlap, function(otu) otu/sum(otu))
  phy_genus_overlap_RA <- filter_taxa(phy_genus_overlap_RA, function(x) sum(x >= 0.0001) >= (nsamples(ps)/10), TRUE)
  phy_genus_overlap_filt <- microbiome::transform(phy_genus_overlap_RA, transform = "clr")
  
  # Convert OTU table to dataframe
  otu_table_overlap <- as.data.frame(as.matrix(phy_genus_overlap_filt@otu_table))
  setDT(otu_table_overlap, keep.rownames = TRUE)[]
  colnames(otu_table_overlap)[1] <- "SampleID"
  
  # Process OTU table
  otu_table2_overlap <- otu_table_overlap %>% data.frame(check.names=FALSE) %>% set_rownames(.$SampleID)
  otu_table2_overlap <- otu_table2_overlap[,-1]
  otu_table2_overlap <- t(otu_table2_overlap)
  
  # Get taxonomy table
  tax_table_overlap <- as.data.frame(as.matrix(phy_genus_overlap_filt@tax_table))
  
  # Merge OTU and taxonomy tables
  otu_tax_overlap <- merge(otu_table2_overlap, tax_table_overlap, by = 0, all = TRUE)
  
  # Process taxonomy information
  otu_tax_overlap$Genus <- replace_na(otu_tax_overlap$Genus, "_unclassified")
  otu_tax_overlap$Phylum <- ifelse(is.na(otu_tax_overlap$Phylum), otu_tax_overlap$Kingdom, otu_tax_overlap$Phylum)
  otu_tax_overlap$Class <- ifelse(is.na(otu_tax_overlap$Class), otu_tax_overlap$Phylum, otu_tax_overlap$Class)
  otu_tax_overlap$Order <- ifelse(is.na(otu_tax_overlap$Order), otu_tax_overlap$Class, otu_tax_overlap$Order)
  otu_tax_overlap$Family <- ifelse(is.na(otu_tax_overlap$Family), otu_tax_overlap$Order, otu_tax_overlap$Family)
  otu_tax_overlap$Genus <- ifelse(otu_tax_overlap$Genus == "_unclassified", paste(otu_tax_overlap$Family, otu_tax_overlap$Genus, sep = ""), otu_tax_overlap$Genus)
  otu_tax_overlap$Genus <- gsub("NA_unclassified", "Unclassified", otu_tax_overlap$Genus)
  otu_tax_overlap$Family <- ifelse(otu_tax_overlap$Family == "uncultured", paste0("uncultured_", otu_tax_overlap$Order), otu_tax_overlap$Family)
  otu_tax_overlap$Genus <- ifelse(grepl("uncultured", otu_tax_overlap$Family), otu_tax_overlap$Family, otu_tax_overlap$Genus)
  otu_tax_overlap$Genus <- ifelse(otu_tax_overlap$Genus == "uncultured", paste0("uncultured_", otu_tax_overlap$Family), otu_tax_overlap$Genus)
  
  # Set genus as rownames and remove taxonomic columns
  rownames(otu_tax_overlap) <- otu_tax_overlap$Genus
  otu_tax_overlap <- subset(otu_tax_overlap, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus, Species))
  otu_tax_overlap <- as.data.frame(t(otu_tax_overlap))
  
  # Fix names
  rownames(otu_tax_overlap) <- phy_genus_overlap@sam_data$UNQ_ID
  
  # Prepare for correlation
  overlap_otu_corr <- matrix(ncol = ncol(otu_tax_overlap), nrow = nrow(otu_tax_overlap))
  colnames(overlap_otu_corr) <- colnames(otu_tax_overlap)
  rownames(overlap_otu_corr) <- rownames(otu_tax_overlap)
  
  # Get sample data
  overlap_sam <- as.data.frame(as.matrix(phy_genus_overlap_filt@sam_data))
  
  # Add variables
  otu_tax_overlap$PCRconc <- as.numeric(overlap_sam$p.CleanedPCRConcentration)
  otu_tax_overlap$exbatch <- as.factor(overlap_sam$ex.ExtractionBatch)
  otu_tax_overlap$sex <- as.factor(overlap_sam$i.Sex)
  otu_tax_overlap$age <- as.numeric(overlap_sam$i.Age)
  
  # Run linear models with progress bar
  total_cols <- ncol(overlap_otu_corr)
  pb <- txtProgressBar(min = 0, max = total_cols, style = 3)
  
  set.seed(123)
  for(i in seq_along(colnames(overlap_otu_corr))){
    col_name <- colnames(overlap_otu_corr)[i]
    obj <- lm(as.formula(paste0("`", col_name, "`", "~ PCRconc + exbatch + sex + age")),  data = otu_tax_overlap)
    overlap_otu_corr[, col_name] <- resid(obj)
    setTxtProgressBar(pb, i)
  }
  
  close(pb)
  
  # Convert to dataframe and return
  overlap_otu_corr <- as.data.frame(overlap_otu_corr)
  return(overlap_otu_corr)
}
```

## filter samples and adjust order
```{r}
micro_overlap_ps <- subset_samples(micro_ps, UNQ_ID != "NA")
CD_overlap_ps <- subset_samples(micro_overlap_ps, i.Disease == "Crohns")

rnaseq_meta_overlap <- subset(rnaseq_meta, UNQ_ID != "NA")
rna_meta_CD <- subset(rnaseq_meta_overlap, i.Disease == "Crohns")

## adjust order of rna metadata based on microbiome
rna_meta_CD_reorder <- rna_meta_CD[order(match(rna_meta_CD$UNQ_ID, CD_overlap_ps@sam_data$UNQ_ID)),]

rna_count_CD <- subset(rnaseq_count, rownames(rnaseq_count) %in% rna_meta_CD_reorder$p.LibraryID)

## adjust order of rna count based on rna metadata
rna_count_CD_reorder <- rna_count_CD[order(match(rownames(rna_count_CD), rna_meta_CD_reorder$p.LibraryID)),]

rownames(rna_count_CD_reorder) <- rna_meta_CD_reorder$UNQ_ID

rna_count_CD_reorder_t <- t(rna_count_CD_reorder)

## check order
identical(CD_overlap_ps@sam_data$UNQ_ID, rownames(rna_count_CD_reorder))
```

## correct data for covariates, including specimen
```{r}
# RNAseq
## checks
identical(rownames(rna_count_CD_reorder), rna_meta_CD_reorder$UNQ_ID)

## normalize
set.seed(123)
dds <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ 1)
smallestGroupSize <- ncol(rna_meta_CD)/3
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

set.seed(123)
vsd <- vst(dds, blind=FALSE)
mat_inte <- assay(vsd)

## Convert mat_inte to a dataframe and transpose
mat_inte_df <- as.data.frame(t(mat_inte))

## filter to protein coding genes only
mat_inte_df <- mat_inte_df[, colnames(mat_inte_df) %in% anno_protein_coding$symbol]

## Create an empty matrix for corrected data
mat_inte_corr <- matrix(ncol = ncol(mat_inte_df), nrow = nrow(mat_inte_df))
colnames(mat_inte_corr) <- colnames(mat_inte_df)
rownames(mat_inte_corr) <- rownames(mat_inte_df)

## Add covariates
mat_inte_df$lpconc <- as.numeric(rna_meta_CD_reorder$p.LibraryConcentration_ngul)
mat_inte_df$extractor <- as.factor(rna_meta_CD_reorder$ex.Extractedby)
mat_inte_df$sbatch <- as.factor(rna_meta_CD_reorder$s.SequencingBatch)
mat_inte_df$RINscore <- as.numeric(rna_meta_CD_reorder$ex.RINscore)
mat_inte_df$sex <- as.factor(rna_meta_CD_reorder$i.Sex)
mat_inte_df$age <- as.numeric(rna_meta_CD_reorder$i.Age)
mat_inte_df$spec <- as.factor(rna_meta_CD_reorder$i.Specimen)
mat_inte_df$id <- as.factor(rna_meta_CD_reorder$i.PatientID)

## Run linear models and store residuals
set.seed(123)
for(i in colnames(mat_inte_corr)){
  
  obj <- lmer(as.formula(paste0("`", i, "`", "~ lpconc + sbatch + RINscore + extractor +
                                sex + age + spec + (1|id)")),  data = mat_inte_df, 
              control = lmerControl(optCtrl = list(maxfun = 1000000), 
                                    check.conv.singular = .makeCC(action = "ignore", tol = formals(isSingular)$tol)))
  mat_inte_corr[,i] <- resid(obj)
}

## filter low variance genes
mat_inte_corr_t <- t(mat_inte_corr)
## calculate variance for each row
gene_variance <- apply(mat_inte_corr_t, 1, var)
## calculate the 25% quantile of variance
variance_cutoff <- quantile(gene_variance, 0.25)
## filter to genes above the 25% quantile of variance
mat_inte_filt <-  mat_inte_corr_t[gene_variance > variance_cutoff, ]
# Convert the corrected matrix to a dataframe
rna_corr_ALL <- as.data.frame(t(mat_inte_filt))


# microbiome
## agglomerate to Genus, transform to RA, and then CLR transform
CD_genus_overlap <- tax_glom(CD_overlap_ps, "Genus", NArm = FALSE)
CD_genus_overlap_RA <- transform_sample_counts(CD_genus_overlap, function(otu) otu/sum(otu))
CD_genus_overlap_RA <- filter_taxa(CD_genus_overlap_RA, function(x) sum(x >= 0.0001) >= (nsamples(CD_overlap_ps)/10), TRUE)
CD_genus_overlap_filt <- microbiome::transform(CD_genus_overlap_RA, transform = "clr")
  
## Convert OTU table to dataframe
CD_otu_table_overlap <- as.data.frame(as.matrix(CD_genus_overlap_filt@otu_table))
setDT(CD_otu_table_overlap, keep.rownames = TRUE)[]
colnames(CD_otu_table_overlap)[1] <- "SampleID"
  
## fix formatting
CD_otu_table2_overlap <- CD_otu_table_overlap %>% data.frame(check.names=FALSE) %>% set_rownames(.$SampleID)
CD_otu_table2_overlap <- CD_otu_table2_overlap[,-1]
CD_otu_table2_overlap <- t(CD_otu_table2_overlap)
  
## Get taxonomy table
CD_tax_table_overlap <- as.data.frame(as.matrix(CD_genus_overlap_filt@tax_table))
  
## Merge OTU and taxonomy tables
CD_otu_tax_overlap <- merge(CD_otu_table2_overlap, CD_tax_table_overlap, by = 0, all = TRUE)
  
## fix taxonomy information
CD_otu_tax_overlap$Genus <- replace_na(CD_otu_tax_overlap$Genus, "_unclassified")
CD_otu_tax_overlap$Phylum <- ifelse(is.na(CD_otu_tax_overlap$Phylum), CD_otu_tax_overlap$Kingdom, CD_otu_tax_overlap$Phylum)
CD_otu_tax_overlap$Class <- ifelse(is.na(CD_otu_tax_overlap$Class), CD_otu_tax_overlap$Phylum, CD_otu_tax_overlap$Class)
CD_otu_tax_overlap$Order <- ifelse(is.na(CD_otu_tax_overlap$Order), CD_otu_tax_overlap$Class, CD_otu_tax_overlap$Order)
CD_otu_tax_overlap$Family <- ifelse(is.na(CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Order, CD_otu_tax_overlap$Family)
CD_otu_tax_overlap$Genus <- ifelse(CD_otu_tax_overlap$Genus == "_unclassified", paste(CD_otu_tax_overlap$Family, CD_otu_tax_overlap$Genus, sep = ""), CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Genus <- gsub("NA_unclassified", "Unclassified", CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Family <- ifelse(CD_otu_tax_overlap$Family == "uncultured", paste0("uncultured_", CD_otu_tax_overlap$Order), CD_otu_tax_overlap$Family)
CD_otu_tax_overlap$Genus <- ifelse(grepl("uncultured", CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Family, CD_otu_tax_overlap$Genus)
CD_otu_tax_overlap$Genus <- ifelse(CD_otu_tax_overlap$Genus == "uncultured", paste0("uncultured_", CD_otu_tax_overlap$Family), CD_otu_tax_overlap$Genus)
  
## Set genus as rownames and remove taxonomic columns
rownames(CD_otu_tax_overlap) <- CD_otu_tax_overlap$Genus
CD_otu_tax_overlap <- subset(CD_otu_tax_overlap, select = -c(Row.names, Kingdom, Phylum, Class, Order, Family, Genus, Species))
CD_otu_tax_overlap <- as.data.frame(t(CD_otu_tax_overlap))
  
## Fix names
rownames(CD_otu_tax_overlap) <- CD_genus_overlap_filt@sam_data$UNQ_ID
  
## Prepare for correlation
CD_overlap_otu_corr <- matrix(ncol = ncol(CD_otu_tax_overlap), nrow = nrow(CD_otu_tax_overlap))
colnames(CD_overlap_otu_corr) <- colnames(CD_otu_tax_overlap)
rownames(CD_overlap_otu_corr) <- rownames(CD_otu_tax_overlap)
  
## Get sample data
CD_overlap_sam <- as.data.frame(as.matrix(CD_genus_overlap_filt@sam_data))
  
## Add variables
CD_otu_tax_overlap$PCRconc <- as.numeric(CD_overlap_sam$p.CleanedPCRConcentration)
CD_otu_tax_overlap$exbatch <- as.factor(CD_overlap_sam$ex.ExtractionBatch)
CD_otu_tax_overlap$sex <- as.factor(CD_overlap_sam$i.Sex)
CD_otu_tax_overlap$age <- as.numeric(CD_overlap_sam$i.Age)
CD_otu_tax_overlap$id <- as.factor(CD_overlap_sam$i.PatientID)
CD_otu_tax_overlap$spec <- as.factor(CD_overlap_sam$i.Specimen)
  
## Run linear models
set.seed(123)
for(i in colnames(CD_overlap_otu_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ PCRconc + exbatch + sex + age + spec + (1|id)")),  
              data = CD_otu_tax_overlap, control = lmerControl(optCtrl = list(maxfun = 1000000), 
                                    check.conv.singular = .makeCC(action = "ignore", tol =
                                                                    formals(isSingular)$tol)))
  CD_overlap_otu_corr[,i] <- resid(obj)
}

## Convert to dataframe
CD_overlap_otu_corr <- as.data.frame(CD_overlap_otu_corr)

## checks
identical(rownames(CD_overlap_otu_corr), rownames(rna_corr_ALL))
```

## save intestine-wide individual datasets
```{r}
write.csv(CD_overlap_otu_corr, "../output/micro_indiv_corrected.csv")
write.csv(rna_corr_ALL, "../output/rna_indiv_corrected.csv")
```

## filter data by region
```{r}
rna_meta_TI <- subset(rna_meta_CD_reorder, i.Specimen == "Terminal Ileum")
rna_meta_C <- subset(rna_meta_CD_reorder, i.Specimen == "Cecum")
rna_meta_RC <- subset(rna_meta_CD_reorder, i.Specimen == "Right Colon")

micro_TI_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Terminal Ileum")
micro_C_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Cecum")
micro_RC_ps <- subset_samples(CD_overlap_ps, i.Specimen == "Right Colon")
```

## transform RNAseq data by region
```{r}
set.seed(123)
dds_full_spec <- DESeqDataSetFromMatrix(countData = rna_count_CD_reorder_t, colData = rna_meta_CD_reorder, design = ~ i.Specimen)

smallestGroupSize_spec <- 30
keep_spec <- rowSums(counts(dds_full_spec) >= 10) >= smallestGroupSize_spec
dds_full_spec <- dds_full_spec[keep_spec,]
set.seed(123)
vsd_spec <- vst(dds_full_spec, blind=FALSE)
mat_spec <- assay(vsd_spec)

mat_spec <- as.data.frame(t(mat_spec))

mat_TI <- mat_spec[rownames(mat_spec) %in% rna_meta_TI$UNQ_ID,]
mat_C <- mat_spec[rownames(mat_spec) %in% rna_meta_C$UNQ_ID,]
mat_RC <- mat_spec[rownames(mat_spec) %in% rna_meta_RC$UNQ_ID,]
```

## correct regional datasets
```{r}
# checks
identical(rownames(mat_TI), rna_meta_TI$UNQ_ID)
identical(rownames(mat_C), rna_meta_C$UNQ_ID)
identical(rownames(mat_RC), rna_meta_RC$UNQ_ID)

# correct data
set.seed(123)
rna_corr_TI <- correct_rnaseq_data_within_region(mat_TI, rna_meta_TI)
set.seed(123)
rna_corr_C <- correct_rnaseq_data_within_region(mat_C, rna_meta_C)
set.seed(123)
rna_corr_RC <- correct_rnaseq_data_within_region(mat_RC, rna_meta_RC)
set.seed(123)
micro_corr_TI <- correct_microbiome_data_within_region(micro_TI_ps)
set.seed(123)
micro_corr_C <- correct_microbiome_data_within_region(micro_C_ps)
set.seed(123)
micro_corr_RC <- correct_microbiome_data_within_region(micro_RC_ps)

# check
identical(rownames(rna_corr_TI), rownames(micro_corr_TI))
identical(rownames(rna_corr_C), rownames(micro_corr_C))
identical(rownames(rna_corr_RC), rownames(micro_corr_RC))
```

## save regional individual datasets
```{r}
write.csv(rna_corr_TI, "../output/rna_indiv_corrected_TI.csv")
write.csv(micro_corr_TI, "../output/micro_indiv_corrected_TI.csv")
write.csv(rna_corr_C, "../output/rna_indiv_corrected_C.csv")
write.csv(micro_corr_C, "../output/micro_indiv_corrected_C.csv")
write.csv(rna_corr_RC, "../output/rna_indiv_corrected_RC.csv")
write.csv(micro_corr_RC, "../output/micro_indiv_corrected_RC.csv")
```

