---
title: "3.2.ind_association_analysis"
output: html_document
date: "2025-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3.3 Individual Association Analysis

The purpose of this script is to identify individual associations between genes and microbes

## load data
```{r}
# elastic net results
intestine_wide_enet <- readRDS("../output/df_indiv.RDS")
TI_enet <- readRDS("../output/df_TI.RDS")
C_enet <- readRDS("../output/df_C.RDS")
RC_enet <- readRDS("../output/df_RC.RDS")

# microbial data
micro_ps <- readRDS("../../microbiome/3.microbiome_analysis/output/phyloseq_sampleReads_unadj.rds")

# protein coding genes
anno_protein_coding <- read.table("../data/protein_coding.txt", header = TRUE)
hsa_to_path <- read.csv("../data/hsa_to_path.csv", row.names = 1, header = TRUE)
pathway_list <- read.csv("../data/pathway_list.csv", header = TRUE)
PID_path <- read.csv("../data/PID_path_entrez.csv", header = FALSE)

# universe input genes
enet_uni_genes <- read.csv("../output/rna_indiv_corrected.csv", row.names = 1, header = TRUE)
enet_uni_genes_TI <- read.csv("../output/rna_indiv_corrected_TI.csv", row.names = 1, header = TRUE)
enet_uni_genes_C <- read.csv("../output/rna_indiv_corrected_C.csv", row.names = 1, header = TRUE)
enet_uni_genes_RC <- read.csv("../output/rna_indiv_corrected_RC.csv", row.names = 1, header = TRUE)

# microbe input
enet_microbe <- read.csv("../output/micro_indiv_corrected.csv", row.names = 1, header = TRUE) 
enet_microbe_TI <- read.csv("../output/micro_indiv_corrected_TI.csv", row.names = 1, header = TRUE) 
enet_microbe_C <- read.csv("../output/micro_indiv_corrected_C.csv", row.names = 1, header = TRUE)
enet_microbe_RC <- read.csv("../output/micro_indiv_corrected_RC.csv", row.names = 1, header = TRUE)
```

## load packages
```{r}
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(clusterProfiler)
library(phyloseq)
library(lme4)
library(data.table)
library(dplyr)
library(tidyr)
library(magrittr)
library(tidyverse)
library(org.Hs.eg.db)
```

## extract enet results and format
```{r}
intestine_wide_enet_param <- intestine_wide_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

intestine_wide_enet_coef <- intestine_wide_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

# remove 0s
enet_coef_filt <- intestine_wide_enet_coef[!rownames(intestine_wide_enet_coef) %in% c("(Intercept)"),]
enet_coef_filt_new <- enet_coef_filt[rowSums(enet_coef_filt != 0) > 0,]
enet_coef_filt_new <- enet_coef_filt_new[,colSums(enet_coef_filt_new != 0) > 0]

# format into gene - microbe tables
enet_genes_mic <- matrix(ncol = ncol(enet_coef_filt_new), nrow = nrow(enet_coef_filt_new))
colnames(enet_genes_mic) <- colnames(enet_coef_filt_new)

for(i in colnames(enet_genes_mic)){
  nonzero <- subset(enet_coef_filt_new, enet_coef_filt_new[[i]] != 0)
  genes <- rownames(nonzero)
  length(genes) <- nrow(enet_coef_filt_new)
  enet_genes_mic[,i] <- genes
}

# clean up columns and remove duplicates
enet_genes_mic <- as.data.frame(enet_genes_mic)
enet_genes_clean <- pivot_longer(enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
enet_genes_clean <- enet_genes_clean[!is.na(enet_genes_clean$gene), ]
enet_genes_only <- enet_genes_clean$gene
enet_genes_only <- unique(enet_genes_only)
```

## identify relationships
```{r}
enet_coef_freq <- as.data.frame(sapply(enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(enet_coef_freq) <- rownames(enet_coef_filt_new)

enet_assoc <- enet_coef_freq
enet_assoc$gene <- rownames(enet_assoc)
enet_assoc <- pivot_longer(enet_assoc, !gene, names_to = "microbe", values_to = "rel")
enet_assoc <- filter(enet_assoc, rel == "Yes")

enet_assoc_all <- paste(enet_assoc$gene, enet_assoc$microbe, sep = "_")

enet_micro <- as.data.frame(table(enet_assoc$microbe))
colnames(enet_micro) <- c("microbe", "num_assoc")

ggplot(enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))
```

## extract regional enet results and format
```{r}
# TI
TI_enet_param <- TI_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

TI_enet_coef <- TI_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
TI_enet_coef_filt <- TI_enet_coef[!rownames(TI_enet_coef) %in% c("(Intercept)"),]
TI_enet_coef_filt_new <- TI_enet_coef_filt[rowSums(TI_enet_coef_filt != 0) > 0,]
TI_enet_coef_filt_new <- TI_enet_coef_filt_new[,colSums(TI_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
TI_enet_genes_mic <- matrix(ncol = ncol(TI_enet_coef_filt_new), nrow = nrow(TI_enet_coef_filt_new))
colnames(TI_enet_genes_mic) <- colnames(TI_enet_coef_filt_new)

for(i in colnames(TI_enet_genes_mic)){
  TI_nonzero <- subset(TI_enet_coef_filt_new, TI_enet_coef_filt_new[[i]] != 0)
  TI_genes <- rownames(TI_nonzero)
  length(TI_genes) <- nrow(TI_enet_coef_filt_new)
  TI_enet_genes_mic[,i] <- TI_genes
}

## clean up columns and remove duplicates
TI_enet_genes_mic <- as.data.frame(TI_enet_genes_mic)
TI_enet_genes_clean <- pivot_longer(TI_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
TI_enet_genes_clean <- TI_enet_genes_clean[!is.na(TI_enet_genes_clean$gene), ]
TI_enet_genes_only <- TI_enet_genes_clean$gene
TI_enet_genes_only <- unique(TI_enet_genes_only)



# C
C_enet_param <- C_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

C_enet_coef <- C_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
C_enet_coef_filt <- C_enet_coef[!rownames(C_enet_coef) %in% c("(Intercept)"),]
C_enet_coef_filt_new <- C_enet_coef_filt[rowSums(C_enet_coef_filt != 0) > 0,]
C_enet_coef_filt_new <- C_enet_coef_filt_new[,colSums(C_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
C_enet_genes_mic <- matrix(ncol = ncol(C_enet_coef_filt_new), nrow = nrow(C_enet_coef_filt_new))
colnames(C_enet_genes_mic) <- colnames(C_enet_coef_filt_new)

for(i in colnames(C_enet_genes_mic)){
  C_nonzero <- subset(C_enet_coef_filt_new, C_enet_coef_filt_new[[i]] != 0)
  C_genes <- rownames(C_nonzero)
  length(C_genes) <- nrow(C_enet_coef_filt_new)
  C_enet_genes_mic[,i] <- C_genes
}

## clean up columns and remove duplicates
C_enet_genes_mic <- as.data.frame(C_enet_genes_mic)
C_enet_genes_clean <- pivot_longer(C_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
C_enet_genes_clean <- C_enet_genes_clean[!is.na(C_enet_genes_clean$gene), ]
C_enet_genes_only <- C_enet_genes_clean$gene
C_enet_genes_only <- unique(C_enet_genes_only)



# RC
RC_enet_param <- RC_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

RC_enet_coef <- RC_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
RC_enet_coef_filt <- RC_enet_coef[!rownames(RC_enet_coef) %in% c("(Intercept)"),]
RC_enet_coef_filt_new <- RC_enet_coef_filt[rowSums(RC_enet_coef_filt != 0) > 0,]
RC_enet_coef_filt_new <- RC_enet_coef_filt_new[,colSums(RC_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
RC_enet_genes_mic <- matrix(ncol = ncol(RC_enet_coef_filt_new), nrow = nrow(RC_enet_coef_filt_new))
colnames(RC_enet_genes_mic) <- colnames(RC_enet_coef_filt_new)

for(i in colnames(RC_enet_genes_mic)){
  RC_nonzero <- subset(RC_enet_coef_filt_new, RC_enet_coef_filt_new[[i]] != 0)
  RC_genes <- rownames(RC_nonzero)
  length(RC_genes) <- nrow(RC_enet_coef_filt_new)
  RC_enet_genes_mic[,i] <- RC_genes
}

## clean up columns and remove duplicates
RC_enet_genes_mic <- as.data.frame(RC_enet_genes_mic)
RC_enet_genes_clean <- pivot_longer(RC_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
RC_enet_genes_clean <- RC_enet_genes_clean[!is.na(RC_enet_genes_clean$gene), ]
RC_enet_genes_only <- RC_enet_genes_clean$gene
RC_enet_genes_only <- unique(RC_enet_genes_only)
```

## identify relationships by region
```{r}
# TI
TI_enet_coef_freq <- as.data.frame(sapply(TI_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(TI_enet_coef_freq) <- rownames(TI_enet_coef_filt_new)

TI_enet_assoc <- TI_enet_coef_freq
TI_enet_assoc$gene <- rownames(TI_enet_assoc)
TI_enet_assoc <- pivot_longer(TI_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
TI_enet_assoc <- filter(TI_enet_assoc, rel == "Yes")

TI_enet_assoc_all <- paste(TI_enet_assoc$gene, TI_enet_assoc$microbe, sep = "_")

TI_enet_micro <- as.data.frame(table(TI_enet_assoc$microbe))
colnames(TI_enet_micro) <- c("microbe", "num_assoc")

ggplot(TI_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))


# C
C_enet_coef_freq <- as.data.frame(sapply(C_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(C_enet_coef_freq) <- rownames(C_enet_coef_filt_new)

C_enet_assoc <- C_enet_coef_freq
C_enet_assoc$gene <- rownames(C_enet_assoc)
C_enet_assoc <- pivot_longer(C_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
C_enet_assoc <- filter(C_enet_assoc, rel == "Yes")

C_enet_assoc_all <- paste(C_enet_assoc$gene, C_enet_assoc$microbe, sep = "_")

C_enet_micro <- as.data.frame(table(C_enet_assoc$microbe))
colnames(C_enet_micro) <- c("microbe", "num_assoc")

ggplot(C_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))



# RC
RC_enet_coef_freq <- as.data.frame(sapply(RC_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(RC_enet_coef_freq) <- rownames(RC_enet_coef_filt_new)

RC_enet_assoc <- RC_enet_coef_freq
RC_enet_assoc$gene <- rownames(RC_enet_assoc)
RC_enet_assoc <- pivot_longer(RC_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
RC_enet_assoc <- filter(RC_enet_assoc, rel == "Yes")

RC_enet_assoc_all <- paste(RC_enet_assoc$gene, RC_enet_assoc$microbe, sep = "_")

RC_enet_micro <- as.data.frame(table(RC_enet_assoc$microbe))
colnames(RC_enet_micro) <- c("microbe", "num_assoc")

ggplot(RC_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))
```

## prep data for saving
```{r}
# intestine-wide
enet_coef_filt_new_prep <- rownames_to_column(enet_coef_filt_new, var = "gene")
enet_coef_df <- pivot_longer(enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
enet_coef_df <- filter(enet_coef_df, coef != 0)

# TI
TI_enet_coef_filt_new_prep <- rownames_to_column(TI_enet_coef_filt_new, var = "gene")
TI_enet_coef_df <- pivot_longer(TI_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
TI_enet_coef_df <- filter(TI_enet_coef_df, coef != 0)

# C
C_enet_coef_filt_new_prep <- rownames_to_column(C_enet_coef_filt_new, var = "gene")
C_enet_coef_df <- pivot_longer(C_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
C_enet_coef_df <- filter(C_enet_coef_df, coef != 0)

# RC
RC_enet_coef_filt_new_prep <- rownames_to_column(RC_enet_coef_filt_new, var = "gene")
RC_enet_coef_df <- pivot_longer(RC_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
RC_enet_coef_df <- filter(RC_enet_coef_df, coef != 0)
```

## generate table of associations
```{r}
enet_coef_df_prep <- enet_coef_df
TI_enet_coef_df_prep <- TI_enet_coef_df
C_enet_coef_df_prep <- C_enet_coef_df
RC_enet_coef_df_prep <- RC_enet_coef_df

enet_coef_df_prep$type <- "intestine"
TI_enet_coef_df_prep$type <- "TI"
C_enet_coef_df_prep$type <- "C"
RC_enet_coef_df_prep$type <- "RC"

all_enet_coef_df <- rbind(enet_coef_df_prep, TI_enet_coef_df_prep)
all_enet_coef_df <- rbind(all_enet_coef_df, C_enet_coef_df_prep)
all_enet_coef_df <- rbind(all_enet_coef_df, RC_enet_coef_df_prep)

# write.csv(all_enet_coef_df, "../output/all_enet_coef_df.csv")
```

## plot associations of interest
```{r}
# TI
lacto_acat1_TI_plot <- ggplot(TI_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_TI_plot <- ggplot(TI_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

# C
lacto_acat1_C_plot <- ggplot(C_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_C_plot <- ggplot(C_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

# RC
lacto_acat1_RC_plot <- ggplot(RC_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_RC_plot <- ggplot(RC_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

## correlation values
cor.test(TI_df$ACAT1, TI_df$Lactococcus, method = c("pearson"))
cor.test(TI_df$PCDH20, TI_df$Faecalitalea, method = c("pearson"))
cor.test(C_df$ACAT1, C_df$Lactococcus, method = c("pearson"))
cor.test(C_df$PCDH20, C_df$Faecalitalea, method = c("pearson"))
cor.test(RC_df$ACAT1, RC_df$Lactococcus, method = c("pearson"))
cor.test(RC_df$PCDH20, RC_df$Faecalitalea, method = c("pearson"))

## save
ggsave(file="../figures/lacto_acat1_TI_plot.pdf", plot=lacto_acat1_TI_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_TI_plot.pdf", plot=faecal_pcdh20_TI_plot, width=2.5, height=2)
ggsave(file="../figures/lacto_acat1_C_plot.pdf", plot=lacto_acat1_C_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_C_plot.pdf", plot=faecal_pcdh20_C_plot, width=2.5, height=2)
ggsave(file="../figures/lacto_acat1_RC_plot.pdf", plot=lacto_acat1_RC_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_RC_plot.pdf", plot=faecal_pcdh20_RC_plot, width=2.5, height=2)
```
