---
title: "3.2.ind_association_analysis"
output: html_document
date: "2025-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3.3 Individual Association Analysis

The purpose of this script is to identify individual associations between genes and microbes

## load data
```{r}
# elastic net results
intestine_wide_enet <- readRDS("../output/df_indiv.RDS")
TI_enet <- readRDS("../output/df_TI.RDS")
C_enet <- readRDS("../output/df_C.RDS")
RC_enet <- readRDS("../output/df_RC.RDS")

# microbial data
micro_ps <- readRDS("../../microbiome/3.microbiome_analysis/output/phyloseq_sampleReads_unadj.rds")

# protein coding genes
anno_protein_coding <- read.table("../data/protein_coding.txt", header = TRUE)
hsa_to_path <- read.csv("../data/hsa_to_path.csv", row.names = 1, header = TRUE)
pathway_list <- read.csv("../data/pathway_list.csv", header = TRUE)
PID_path <- read.csv("../data/PID_path_entrez.csv", header = FALSE)

# universe input genes
enet_uni_genes <- read.csv("../output/rna_indiv_corrected.csv", row.names = 1, header = TRUE)
enet_uni_genes_TI <- read.csv("../output/rna_indiv_corrected_TI.csv", row.names = 1, header = TRUE)
enet_uni_genes_C <- read.csv("../output/rna_indiv_corrected_C.csv", row.names = 1, header = TRUE)
enet_uni_genes_RC <- read.csv("../output/rna_indiv_corrected_RC.csv", row.names = 1, header = TRUE)

# microbe input
enet_microbe <- read.csv("../output/micro_indiv_corrected.csv", row.names = 1, header = TRUE) 
enet_microbe_TI <- read.csv("../output/micro_indiv_corrected_TI.csv", row.names = 1, header = TRUE) 
enet_microbe_C <- read.csv("../output/micro_indiv_corrected_C.csv", row.names = 1, header = TRUE)
enet_microbe_RC <- read.csv("../output/micro_indiv_corrected_RC.csv", row.names = 1, header = TRUE)

# GSEA results
GSEA_output <- read.table("../output/GSEA_tab.tsv", header = TRUE, sep = "\t")
```

## load packages
```{r}
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(clusterProfiler)
library(phyloseq)
library(lme4)
library(data.table)
library(dplyr)
library(tidyr)
library(magrittr)
library(tidyverse)
library(org.Hs.eg.db)
```

## extract enet results and format
```{r}
intestine_wide_enet_param <- intestine_wide_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

intestine_wide_enet_coef <- intestine_wide_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

# remove 0s
enet_coef_filt <- intestine_wide_enet_coef[!rownames(intestine_wide_enet_coef) %in% c("(Intercept)"),]
enet_coef_filt_new <- enet_coef_filt[rowSums(enet_coef_filt != 0) > 0,]
enet_coef_filt_new <- enet_coef_filt_new[,colSums(enet_coef_filt_new != 0) > 0]

# format into gene - microbe tables
enet_genes_mic <- matrix(ncol = ncol(enet_coef_filt_new), nrow = nrow(enet_coef_filt_new))
colnames(enet_genes_mic) <- colnames(enet_coef_filt_new)

for(i in colnames(enet_genes_mic)){
  nonzero <- subset(enet_coef_filt_new, enet_coef_filt_new[[i]] != 0)
  genes <- rownames(nonzero)
  length(genes) <- nrow(enet_coef_filt_new)
  enet_genes_mic[,i] <- genes
}

# clean up columns and remove duplicates
enet_genes_mic <- as.data.frame(enet_genes_mic)
enet_genes_clean <- pivot_longer(enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
enet_genes_clean <- enet_genes_clean[!is.na(enet_genes_clean$gene), ]
enet_genes_only <- enet_genes_clean$gene
enet_genes_only <- unique(enet_genes_only)
```

## identify relationships
```{r}
enet_coef_freq <- as.data.frame(sapply(enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(enet_coef_freq) <- rownames(enet_coef_filt_new)

enet_assoc <- enet_coef_freq
enet_assoc$gene <- rownames(enet_assoc)
enet_assoc <- pivot_longer(enet_assoc, !gene, names_to = "microbe", values_to = "rel")
enet_assoc <- filter(enet_assoc, rel == "Yes")

enet_assoc_all <- paste(enet_assoc$gene, enet_assoc$microbe, sep = "_")

enet_micro <- as.data.frame(table(enet_assoc$microbe))
colnames(enet_micro) <- c("microbe", "num_assoc")

ggplot(enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))
```

## extract genes and switch to entrez IDs
```{r}
# universe
all_genes <- colnames(enet_uni_genes)

# entrez ID
enet_entrez <- as.character(anno_protein_coding$entrezID[match(enet_genes_only, anno_protein_coding$symbol)])

all_entrez <- as.character(anno_protein_coding$entrezID[match(all_genes, anno_protein_coding$symbol)])
```

## combine all pathway info
```{r}
colnames(hsa_to_path) <- c("pathway", "entrez_ID") 
colnames(PID_path) <- c("pathway", "entrez_ID")

path_input <- bind_rows(hsa_to_path, PID_path)
```

## enrichment
```{r}
enet_enrich <- enricher(enet_entrez,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

enet_enrich_out <- enet_enrich@result

# filter to significant pathways
enet_enrich_filt <- filter(enet_enrich_out, p.adjust < 0.1 & Count >= 10)

# plot
barplot(enet_enrich, showCategory=20)

# examine KEGG pathways
enet_kegg_path <- pathway_list[pathway_list$map %in% rownames(enet_enrich_filt),]

## save
# write.csv(enet_enrich_filt, "../output/enet_enrich_filt.csv")
```

## extract top model hits
```{r}
enet_coef_filt_new_prep <- rownames_to_column(enet_coef_filt_new, var = "gene")
enet_coef_df <- pivot_longer(enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
enet_coef_df <- filter(enet_coef_df, coef != 0)

enet_coef_tophits <- filter(enet_coef_df, coef < -0.1 | coef > 0.1)

# plot distribution
ggplot(enet_coef_df, aes(x = coef)) +
  geom_histogram()

ggplot(enet_coef_tophits, aes(x = coef)) +
  geom_histogram()
```

## generate LMs from enet results intestine-wide
```{r}
enet_df <- cbind(enet_microbe, enet_uni_genes)
enet_pruned <- enet_assoc

for (i in 1:nrow(enet_pruned)) {
  g <- enet_pruned$gene[i]
  m <- enet_pruned$microbe[i]

  # Subset to the samples where both are measured
  df_subset <- enet_df[ , c(g, m), drop = FALSE]

  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = df_subset)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  enet_pruned$rel[i] <- lm_p
}

enet_pruned_BH <- enet_pruned

enet_pruned_BH$rel <- p.adjust(enet_pruned_BH$rel, method = "BH")
```

## generate plots based on top lm results
```{r}
ggplot(enet_df, aes(x = 	ZNF674, y = Gluconacetobacter)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "ZNF674 and Gluconacetobacter")

ggplot(enet_df, aes(x = 	TBC1D9, y = Escherichia.Shigella)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "TBC1D9 and Escherichia.Shigella")

ggplot(enet_df, aes(x = MACO1, y = Gluconobacter)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "MACO1 and Gluconobacter")

# not present in the enet results
ggplot(enet_df, aes(x = SLC6A12, y = NK4A214_group)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "SLC6A12 and NK4A214_group")

ggplot(enet_df, aes(x = CNR2, y = Fusicatenibacter)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "CNR2 and Fusicatenibacter")
```

## extract genes in prep for GSEA
```{r}
enet_rank <- enet_coef_df
enet_rank <- subset(enet_rank, select=-c(microbe))
enet_rank <- enet_rank[order(enet_rank$coef, decreasing = TRUE),]  

# write.table(enet_rank, "../output/enet_rank.rnk", sep = "\t", col.names = FALSE, row.names = FALSE)

```

## classification of top hits
```{r}
enet_tophits_entrez <- as.character(anno_protein_coding$entrezID[match(enet_coef_tophits$gene, anno_protein_coding$symbol)])

# classification
enet_tophits_class_CC <- groupGO(gene = enet_tophits_entrez,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "CC",
                                level    = 3,
                                readable = TRUE)

enet_tophits_class_BP <- groupGO(gene = enet_tophits_entrez,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)

enet_tophits_class_CC_out <- enet_tophits_class_CC@result
enet_tophits_class_BP_out <- enet_tophits_class_BP@result

# filter to significant pathways
enet_tophits_class_CC_filt <- filter(enet_tophits_class_CC_out, Count != 0)
enet_tophits_class_BP_filt <- filter(enet_tophits_class_BP_out, Count != 0)
```

## extract regional enet results and format
```{r}
# TI
TI_enet_param <- TI_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

TI_enet_coef <- TI_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
TI_enet_coef_filt <- TI_enet_coef[!rownames(TI_enet_coef) %in% c("(Intercept)"),]
TI_enet_coef_filt_new <- TI_enet_coef_filt[rowSums(TI_enet_coef_filt != 0) > 0,]
TI_enet_coef_filt_new <- TI_enet_coef_filt_new[,colSums(TI_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
TI_enet_genes_mic <- matrix(ncol = ncol(TI_enet_coef_filt_new), nrow = nrow(TI_enet_coef_filt_new))
colnames(TI_enet_genes_mic) <- colnames(TI_enet_coef_filt_new)

for(i in colnames(TI_enet_genes_mic)){
  TI_nonzero <- subset(TI_enet_coef_filt_new, TI_enet_coef_filt_new[[i]] != 0)
  TI_genes <- rownames(TI_nonzero)
  length(TI_genes) <- nrow(TI_enet_coef_filt_new)
  TI_enet_genes_mic[,i] <- TI_genes
}

## clean up columns and remove duplicates
TI_enet_genes_mic <- as.data.frame(TI_enet_genes_mic)
TI_enet_genes_clean <- pivot_longer(TI_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
TI_enet_genes_clean <- TI_enet_genes_clean[!is.na(TI_enet_genes_clean$gene), ]
TI_enet_genes_only <- TI_enet_genes_clean$gene
TI_enet_genes_only <- unique(TI_enet_genes_only)



# C
C_enet_param <- C_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

C_enet_coef <- C_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
C_enet_coef_filt <- C_enet_coef[!rownames(C_enet_coef) %in% c("(Intercept)"),]
C_enet_coef_filt_new <- C_enet_coef_filt[rowSums(C_enet_coef_filt != 0) > 0,]
C_enet_coef_filt_new <- C_enet_coef_filt_new[,colSums(C_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
C_enet_genes_mic <- matrix(ncol = ncol(C_enet_coef_filt_new), nrow = nrow(C_enet_coef_filt_new))
colnames(C_enet_genes_mic) <- colnames(C_enet_coef_filt_new)

for(i in colnames(C_enet_genes_mic)){
  C_nonzero <- subset(C_enet_coef_filt_new, C_enet_coef_filt_new[[i]] != 0)
  C_genes <- rownames(C_nonzero)
  length(C_genes) <- nrow(C_enet_coef_filt_new)
  C_enet_genes_mic[,i] <- C_genes
}

## clean up columns and remove duplicates
C_enet_genes_mic <- as.data.frame(C_enet_genes_mic)
C_enet_genes_clean <- pivot_longer(C_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
C_enet_genes_clean <- C_enet_genes_clean[!is.na(C_enet_genes_clean$gene), ]
C_enet_genes_only <- C_enet_genes_clean$gene
C_enet_genes_only <- unique(C_enet_genes_only)



# RC
RC_enet_param <- RC_enet %>% 
  map(~ .x[[1]]) %>%
  bind_cols()

RC_enet_coef <- RC_enet %>% 
  map(~ .x[[2]]) %>%
  bind_cols()

## remove 0s
RC_enet_coef_filt <- RC_enet_coef[!rownames(RC_enet_coef) %in% c("(Intercept)"),]
RC_enet_coef_filt_new <- RC_enet_coef_filt[rowSums(RC_enet_coef_filt != 0) > 0,]
RC_enet_coef_filt_new <- RC_enet_coef_filt_new[,colSums(RC_enet_coef_filt_new != 0) > 0]

## format into gene - microbe tables
RC_enet_genes_mic <- matrix(ncol = ncol(RC_enet_coef_filt_new), nrow = nrow(RC_enet_coef_filt_new))
colnames(RC_enet_genes_mic) <- colnames(RC_enet_coef_filt_new)

for(i in colnames(RC_enet_genes_mic)){
  RC_nonzero <- subset(RC_enet_coef_filt_new, RC_enet_coef_filt_new[[i]] != 0)
  RC_genes <- rownames(RC_nonzero)
  length(RC_genes) <- nrow(RC_enet_coef_filt_new)
  RC_enet_genes_mic[,i] <- RC_genes
}

## clean up columns and remove duplicates
RC_enet_genes_mic <- as.data.frame(RC_enet_genes_mic)
RC_enet_genes_clean <- pivot_longer(RC_enet_genes_mic, everything(), names_to = "microbe", values_to = "gene")
RC_enet_genes_clean <- RC_enet_genes_clean[!is.na(RC_enet_genes_clean$gene), ]
RC_enet_genes_only <- RC_enet_genes_clean$gene
RC_enet_genes_only <- unique(RC_enet_genes_only)
```

## identify relationships by region
```{r}
# TI
TI_enet_coef_freq <- as.data.frame(sapply(TI_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(TI_enet_coef_freq) <- rownames(TI_enet_coef_filt_new)

TI_enet_assoc <- TI_enet_coef_freq
TI_enet_assoc$gene <- rownames(TI_enet_assoc)
TI_enet_assoc <- pivot_longer(TI_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
TI_enet_assoc <- filter(TI_enet_assoc, rel == "Yes")

TI_enet_assoc_all <- paste(TI_enet_assoc$gene, TI_enet_assoc$microbe, sep = "_")

TI_enet_micro <- as.data.frame(table(TI_enet_assoc$microbe))
colnames(TI_enet_micro) <- c("microbe", "num_assoc")

ggplot(TI_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))


# C
C_enet_coef_freq <- as.data.frame(sapply(C_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(C_enet_coef_freq) <- rownames(C_enet_coef_filt_new)

C_enet_assoc <- C_enet_coef_freq
C_enet_assoc$gene <- rownames(C_enet_assoc)
C_enet_assoc <- pivot_longer(C_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
C_enet_assoc <- filter(C_enet_assoc, rel == "Yes")

C_enet_assoc_all <- paste(C_enet_assoc$gene, C_enet_assoc$microbe, sep = "_")

C_enet_micro <- as.data.frame(table(C_enet_assoc$microbe))
colnames(C_enet_micro) <- c("microbe", "num_assoc")

ggplot(C_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))



# RC
RC_enet_coef_freq <- as.data.frame(sapply(RC_enet_coef_filt_new, function(x) ifelse(x == 0, "No", "Yes")))
rownames(RC_enet_coef_freq) <- rownames(RC_enet_coef_filt_new)

RC_enet_assoc <- RC_enet_coef_freq
RC_enet_assoc$gene <- rownames(RC_enet_assoc)
RC_enet_assoc <- pivot_longer(RC_enet_assoc, !gene, names_to = "microbe", values_to = "rel")
RC_enet_assoc <- filter(RC_enet_assoc, rel == "Yes")

RC_enet_assoc_all <- paste(RC_enet_assoc$gene, RC_enet_assoc$microbe, sep = "_")

RC_enet_micro <- as.data.frame(table(RC_enet_assoc$microbe))
colnames(RC_enet_micro) <- c("microbe", "num_assoc")

ggplot(RC_enet_micro, aes(x = microbe, y = num_assoc)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95))
```

## extract genes and switch to entrez IDs regional
```{r}
# universe
all_genes_TI <- colnames(enet_uni_genes_TI)
all_genes_C <- colnames(enet_uni_genes_C)
all_genes_RC <- colnames(enet_uni_genes_RC)

# entrez ID
enet_entrez_TI <- as.character(anno_protein_coding$entrezID[match(TI_enet_genes_only, anno_protein_coding$symbol)])
enet_entrez_C <- as.character(anno_protein_coding$entrezID[match(C_enet_genes_only, anno_protein_coding$symbol)])
enet_entrez_RC <- as.character(anno_protein_coding$entrezID[match(RC_enet_genes_only, anno_protein_coding$symbol)])

all_entrez_TI <- as.character(anno_protein_coding$entrezID[match(all_genes_TI, anno_protein_coding$symbol)])
all_entrez_C <- as.character(anno_protein_coding$entrezID[match(all_genes_C, anno_protein_coding$symbol)])
all_entrez_RC <- as.character(anno_protein_coding$entrezID[match(all_genes_RC, anno_protein_coding$symbol)])
```

## enrichment
```{r}
enet_enrich_TI <- enricher(enet_entrez_TI,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez_TI,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)


go_enrich4 <- enrichGO(gene = enet_entrez_RC,
                      universe = all_entrez_RC,
                       OrgDb = org.Hs.eg.db,
                      readable = T,
                       ont = "BP",
                       pvalueCutoff = 0.05, 
                       qvalueCutoff = 0.10)

enet_enrich_C <- enricher(enet_entrez_C,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez_C,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

enet_enrich_RC <- enricher(enet_entrez_RC,
                   pvalueCutoff = 0.05,
                   pAdjustMethod = "BH",
                   universe = all_entrez_RC,
                   minGSSize = 10,
                   maxGSSize = 500,
                   qvalueCutoff = 0.2,
                   TERM2GENE = path_input)

enet_enrich_TI_out <- enet_enrich_TI@result
enet_enrich_C_out <- enet_enrich_C@result
enet_enrich_RC_out <- enet_enrich_RC@result

# filter to significant pathways
enet_enrich_TI_filt <- filter(enet_enrich_TI_out, p.adjust < 0.1 & Count >= 10)
enet_enrich_C_filt <- filter(enet_enrich_C_out, p.adjust < 0.1 & Count >= 10)
enet_enrich_RC_filt <- filter(enet_enrich_RC_out, p.adjust < 0.1 & Count >= 10)

# plot
# barplot(enet_enrich_TI, showCategory=20)
barplot(enet_enrich_C, showCategory=20)
# barplot(enet_enrich_RC, showCategory=20)

# examine KEGG pathways
enet_TI_kegg_path <- pathway_list[pathway_list$map %in% rownames(enet_enrich_TI_filt),]
enet_C_kegg_path <- pathway_list[pathway_list$map %in% rownames(enet_enrich_C_filt),]
enet_RC_kegg_path <- pathway_list[pathway_list$map %in% rownames(enet_enrich_RC_filt),]

## save
# write.csv(enet_enrich_TI_filt, "../output/enet_enrich_TI_filt.csv")
# write.csv(enet_enrich_C_filt, "../output/enet_enrich_C_filt.csv")
# write.csv(enet_enrich_RC_filt, "../output/enet_enrich_RC_filt.csv")
```

## extract top model hits regionally
```{r}
# TI
TI_enet_coef_filt_new_prep <- rownames_to_column(TI_enet_coef_filt_new, var = "gene")
TI_enet_coef_df <- pivot_longer(TI_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
TI_enet_coef_df <- filter(TI_enet_coef_df, coef != 0)

TI_enet_coef_tophits <- filter(TI_enet_coef_df, coef < -0.1 | coef > 0.1)

ggplot(TI_enet_coef_df, aes(x = coef)) +
  geom_histogram()

ggplot(TI_enet_coef_tophits, aes(x = coef)) +
  geom_histogram()

# C
C_enet_coef_filt_new_prep <- rownames_to_column(C_enet_coef_filt_new, var = "gene")
C_enet_coef_df <- pivot_longer(C_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
C_enet_coef_df <- filter(C_enet_coef_df, coef != 0)

C_enet_coef_tophits <- filter(C_enet_coef_df, coef < -0.1 | coef > 0.1)

ggplot(C_enet_coef_df, aes(x = coef)) +
  geom_histogram()

ggplot(C_enet_coef_tophits, aes(x = coef)) +
  geom_histogram()

# RC
RC_enet_coef_filt_new_prep <- rownames_to_column(RC_enet_coef_filt_new, var = "gene")
RC_enet_coef_df <- pivot_longer(RC_enet_coef_filt_new_prep, !gene, names_to = "microbe", values_to = "coef")
RC_enet_coef_df <- filter(RC_enet_coef_df, coef != 0)

RC_enet_coef_tophits <- filter(RC_enet_coef_df, coef < -0.1 | coef > 0.1)

ggplot(RC_enet_coef_df, aes(x = coef)) +
  geom_histogram()

ggplot(RC_enet_coef_tophits, aes(x = coef)) +
  geom_histogram()
```

## examine plots
```{r}
TI_df <- cbind(enet_microbe_TI, enet_uni_genes_TI)
C_df <- cbind(enet_microbe_C, enet_uni_genes_C)
RC_df <- cbind(enet_microbe_RC, enet_uni_genes_RC)

ggplot(TI_df, aes(x = PDE3B, y = Akkermansia)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "PDE3B and Akkermansia in TI")

ggplot(TI_df, aes(x = LRRC37A2, y = Endobacter)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "LRRC37A2 and Endobacter in TI")

ggplot(C_df, aes(x = 	MLH3, y = Akkermansia)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "MLH3 and Akkermansia in C")

ggplot(C_df, aes(x = 	IRF5, y = Pseudomonas)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "IRF5 and Pseudomonas in C")

ggplot(C_df, aes(x = 	PCDH7, y = Enterobacteriaceae_unclassified)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "PCDH7 and Enterobacteriaceae_unclassified in C")

ggplot(C_df, aes(x = 	NPM2, y = Christensenellaceae_R.7_group)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "NPM2 and Christensenellaceae_R.7_group in C")
```

## extract coefficient hits in prep for GSEA
```{r}
# TI
TI_enet_rank <- TI_enet_coef_df
TI_enet_rank <- subset(TI_enet_rank, select=-c(microbe))
TI_enet_rank <- TI_enet_rank[order(TI_enet_rank$coef, decreasing = TRUE),]  

# write.table(TI_enet_rank, "../output/TI_enet_rank.rnk", sep = "\t", col.names = FALSE, row.names = FALSE)


# C
C_enet_rank <- C_enet_coef_df
C_enet_rank <- subset(C_enet_rank, select=-c(microbe))
C_enet_rank <- C_enet_rank[order(C_enet_rank$coef, decreasing = TRUE),]  

# write.table(C_enet_rank, "../output/C_enet_rank.rnk", sep = "\t", col.names = FALSE, row.names = FALSE)


# RC
RC_enet_rank <- RC_enet_coef_df
RC_enet_rank <- subset(RC_enet_rank, select=-c(microbe))
RC_enet_rank <- RC_enet_rank[order(RC_enet_rank$coef, decreasing = TRUE),]  

# write.table(RC_enet_rank, "../output/RC_enet_rank.rnk", sep = "\t", col.names = FALSE, row.names = FALSE)
```

## generate LMs from enet results regionally
```{r}
# TI
TI_enet_pruned <- TI_enet_assoc

for (i in 1:nrow(TI_enet_pruned)) {
  g <- TI_enet_pruned$gene[i]
  m <- TI_enet_pruned$microbe[i]

  # Subset to the samples where both are measured
  df_subset <- TI_df[ , c(g, m), drop = FALSE]

  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = df_subset)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  TI_enet_pruned$rel[i] <- lm_p
}

TI_enet_pruned_BH <- TI_enet_pruned

TI_enet_pruned_BH$rel <- p.adjust(TI_enet_pruned_BH$rel, method = "BH")
 

# C
C_enet_pruned <- C_enet_assoc

for (i in 1:nrow(C_enet_pruned)) {
  g <- C_enet_pruned$gene[i]
  m <- C_enet_pruned$microbe[i]

  # Subset to the samples where both are measured
  df_subset <- C_df[ , c(g, m), drop = FALSE]

  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = df_subset)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  C_enet_pruned$rel[i] <- lm_p
}

C_enet_pruned_BH <- C_enet_pruned

C_enet_pruned_BH$rel <- p.adjust(C_enet_pruned_BH$rel, method = "BH")


# RC
RC_enet_pruned <- RC_enet_assoc

for (i in 1:nrow(RC_enet_pruned)) {
  g <- RC_enet_pruned$gene[i]
  m <- RC_enet_pruned$microbe[i]

  # Subset to the samples where both are measured
  df_subset <- RC_df[ , c(g, m), drop = FALSE]

  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = df_subset)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  RC_enet_pruned$rel[i] <- lm_p
}

RC_enet_pruned_BH <- RC_enet_pruned

RC_enet_pruned_BH$rel <- p.adjust(RC_enet_pruned_BH$rel, method = "BH")
```

## generate plots based on top lm results
```{r}
ggplot(C_df, aes(x = 	SLX4, y = X.Eubacterium._brachy_group)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "SLX4 and X.Eubacterium._brachy_group in C")

# not good and not present in the enet results
ggplot(C_df, aes(x = 	ACOT11, y = Desulfovibrio)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "ACOT11 and Desulfovibrio in C")

ggplot(C_df, aes(x = 	HAX1, y = Agathobacter)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "HAX1 and Agathobacter in C")
```

## classification of top hits
```{r}
enet_tophits_entrez_TI <- as.character(anno_protein_coding$entrezID[match(TI_enet_coef_tophits$gene, anno_protein_coding$symbol)])
enet_tophits_entrez_C <- as.character(anno_protein_coding$entrezID[match(C_enet_coef_tophits$gene, anno_protein_coding$symbol)])
enet_tophits_entrez_RC <- as.character(anno_protein_coding$entrezID[match(RC_enet_coef_tophits$gene, anno_protein_coding$symbol)])

# classification
TI_enet_tophits_class_CC <- groupGO(gene = enet_tophits_entrez_TI,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "CC",
                                level    = 3,
                                readable = TRUE)

TI_enet_tophits_class_BP <- groupGO(gene = enet_tophits_entrez_TI,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)

C_enet_tophits_class_CC <- groupGO(gene = enet_tophits_entrez_C,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "CC",
                                level    = 3,
                                readable = TRUE)

C_enet_tophits_class_BP <- groupGO(gene = enet_tophits_entrez_C,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)

RC_enet_tophits_class_CC <- groupGO(gene = enet_tophits_entrez_RC,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "CC",
                                level    = 3,
                                readable = TRUE)

RC_enet_tophits_class_BP <- groupGO(gene = enet_tophits_entrez_RC,
                                OrgDb    = org.Hs.eg.db,
                                ont      = "BP",
                                level    = 3,
                                readable = TRUE)

TI_enet_tophits_class_CC_out <- TI_enet_tophits_class_CC@result
TI_enet_tophits_class_BP_out <- TI_enet_tophits_class_BP@result
C_enet_tophits_class_CC_out <- C_enet_tophits_class_CC@result
C_enet_tophits_class_BP_out <- C_enet_tophits_class_BP@result
RC_enet_tophits_class_CC_out <- RC_enet_tophits_class_CC@result
RC_enet_tophits_class_BP_out <- RC_enet_tophits_class_BP@result

# filter to significant pathways
TI_enet_tophits_class_CC_filt <- filter(TI_enet_tophits_class_CC_out, Count != 0)
TI_enet_tophits_class_BP_filt <- filter(TI_enet_tophits_class_BP_out, Count != 0)
C_enet_tophits_class_CC_filt <- filter(C_enet_tophits_class_CC_out, Count != 0)
C_enet_tophits_class_BP_filt <- filter(C_enet_tophits_class_BP_out, Count != 0)
RC_enet_tophits_class_CC_filt <- filter(RC_enet_tophits_class_CC_out, Count != 0)
RC_enet_tophits_class_BP_filt <- filter(RC_enet_tophits_class_BP_out, Count != 0)
```

## generate table of associations
```{r}
enet_coef_df_prep <- enet_coef_df
TI_enet_coef_df_prep <- TI_enet_coef_df
C_enet_coef_df_prep <- C_enet_coef_df
RC_enet_coef_df_prep <- RC_enet_coef_df

enet_coef_df_prep$type <- "intestine"
TI_enet_coef_df_prep$type <- "TI"
C_enet_coef_df_prep$type <- "C"
RC_enet_coef_df_prep$type <- "RC"

all_enet_coef_df <- rbind(enet_coef_df_prep, TI_enet_coef_df_prep)
all_enet_coef_df <- rbind(all_enet_coef_df, C_enet_coef_df_prep)
all_enet_coef_df <- rbind(all_enet_coef_df, RC_enet_coef_df_prep)

# write.csv(all_enet_coef_df, "../output/all_enet_coef_df.csv")
```




## generate top hits table
```{r}
shared_rows_enet_TI <- intersect(enet_coef_tophits[ , c("gene", "microbe")], TI_enet_coef_tophits[ , c("gene", "microbe")])
shared_rows_enet_C <- intersect(enet_coef_tophits[ , c("gene", "microbe")], C_enet_coef_tophits[ , c("gene", "microbe")])
shared_rows_enet_RC <- intersect(enet_coef_tophits[ , c("gene", "microbe")], RC_enet_coef_tophits[ , c("gene", "microbe")])
shared_rows_TI_C <- intersect(TI_enet_coef_tophits[ , c("gene", "microbe")], C_enet_coef_tophits[ , c("gene", "microbe")])
shared_rows_TI_RC <- intersect(TI_enet_coef_tophits[ , c("gene", "microbe")], RC_enet_coef_tophits[ , c("gene", "microbe")])
shared_rows_C_RC <- intersect(C_enet_coef_tophits[ , c("gene", "microbe")], RC_enet_coef_tophits[ , c("gene", "microbe")])

# none shared

enet_coef_tophits_prep <- enet_coef_tophits
TI_enet_coef_tophits_prep <- TI_enet_coef_tophits
C_enet_coef_tophits_prep <- C_enet_coef_tophits
RC_enet_coef_tophits_prep <- RC_enet_coef_tophits

enet_coef_tophits_prep$type <- "intestine"
TI_enet_coef_tophits_prep$type <- "TI"
C_enet_coef_tophits_prep$type <- "C"
RC_enet_coef_tophits_prep$type <- "RC"

all_enet_coef_tophits <- rbind(enet_coef_tophits_prep, TI_enet_coef_tophits_prep)
all_enet_coef_tophits <- rbind(all_enet_coef_tophits, C_enet_coef_tophits_prep)
all_enet_coef_tophits <- rbind(all_enet_coef_tophits, RC_enet_coef_tophits_prep)

# write.csv(all_enet_coef_tophits, "../output/all_enet_coef_tophits.csv")
```

## Rsquared distribution
```{r}
intestine_wide_enet_param_Rsq <- intestine_wide_enet_param["Rsquared", ]
TI_enet_param_Rsq <- TI_enet_param["Rsquared", ]
C_enet_param_Rsq <- C_enet_param["Rsquared", ]
RC_enet_param_Rsq <- RC_enet_param["Rsquared", ]

intestine_wide_enet_param_Rsq <- intestine_wide_enet_param_Rsq[, colSums(intestine_wide_enet_param_Rsq > 0, na.rm = TRUE) > 0]
TI_enet_param_Rsq <- TI_enet_param_Rsq[, colSums(TI_enet_param_Rsq > 0, na.rm = TRUE) > 0]
C_enet_param_Rsq <- C_enet_param_Rsq[, colSums(C_enet_param_Rsq > 0, na.rm = TRUE) > 0]
RC_enet_param_Rsq <- RC_enet_param_Rsq[, colSums(RC_enet_param_Rsq > 0, na.rm = TRUE) > 0]

intestine_wide_enet_param_Rsq_df <- as.data.frame(t(intestine_wide_enet_param_Rsq))
TI_enet_param_Rsq_df <- as.data.frame(t(TI_enet_param_Rsq))
C_enet_param_Rsq_df <- as.data.frame(t(C_enet_param_Rsq))
RC_enet_param_Rsq_df <- as.data.frame(t(RC_enet_param_Rsq))

# plot distribution
ggplot(intestine_wide_enet_param_Rsq_df, aes(x = Rsquared)) +
  geom_histogram() +
  geom_vline(xintercept = 0.2, color = "red", linetype = "dashed")

ggplot(TI_enet_param_Rsq_df, aes(x = Rsquared)) +
  geom_histogram() +
  geom_vline(xintercept = 0.4, color = "red", linetype = "dashed")

ggplot(C_enet_param_Rsq_df, aes(x = Rsquared)) +
  geom_histogram() +
  geom_vline(xintercept = 0.4, color = "red", linetype = "dashed")

ggplot(RC_enet_param_Rsq_df, aes(x = Rsquared)) +
  geom_histogram() +
  geom_vline(xintercept = 0.4, color = "red", linetype = "dashed")
```

## set cutoff for models, filter tables, and save
```{r}
intestine_wide_enet_param_Rsq_cutoff <- filter(intestine_wide_enet_param_Rsq_df, Rsquared > 0.2)
TI_enet_param_Rsq_cutoff <- filter(TI_enet_param_Rsq_df, Rsquared > 0.4)
C_enet_param_Rsq_cutoff <- filter(C_enet_param_Rsq_df, Rsquared > 0.4)
RC_enet_param_Rsq_cutoff <- filter(RC_enet_param_Rsq_df, Rsquared > 0.4)

# filter intestine wide
intestine_wide_GSEA_df <- enet_coef_df[enet_coef_df$microbe %in% rownames(intestine_wide_enet_param_Rsq_cutoff), ]

intestine_wide_GSEA_list <- split(intestine_wide_GSEA_df, intestine_wide_GSEA_df$microbe)

intestine_wide_GSEA_list <- lapply(intestine_wide_GSEA_list, function(df) {
  df <- df[, !names(df) %in% "microbe"]
  df[order(df$coef), ]
})

# filter TI
TI_GSEA_df <- TI_enet_coef_df[TI_enet_coef_df$microbe %in% rownames(TI_enet_param_Rsq_cutoff), ]

TI_GSEA_list <- setNames(
  lapply(TI_GSEA_df$microbe, function(name) {
    out <- data.frame(
      gene <- TI_GSEA_df$gene,
      coef <- TI_GSEA_df$coef,
      stringsAsFactors = FALSE
    )
    # order by coef (ascending)
    out[order(out$coef, decreasing = TRUE), ]
  }),
  TI_GSEA_df$microbe
)

# filter C
C_GSEA_df <- C_enet_coef_filt[,colnames(C_enet_coef_filt) %in% rownames(C_enet_param_Rsq_cutoff)]

C_GSEA_list <- setNames(
  lapply(names(C_GSEA_df), function(name) {
    out <- data.frame(
      gene <- rownames(C_GSEA_df),
      coef <- C_GSEA_df[[name]],
      stringsAsFactors = FALSE
    )
    # order by coef (ascending)
    out[order(out$coef, decreasing = TRUE), ]
  }),
  names(C_GSEA_df)
)

# filter RC
RC_GSEA_df <- RC_enet_coef_filt[,colnames(RC_enet_coef_filt) %in% rownames(RC_enet_param_Rsq_cutoff)]

RC_GSEA_list <- setNames(
  lapply(names(RC_GSEA_df), function(name) {
    out <- data.frame(
      gene <- rownames(RC_GSEA_df),
      coef <- RC_GSEA_df[[name]],
      stringsAsFactors = FALSE
    )
    # order by coef (ascending)
    out[order(out$coef, decreasing = TRUE), ]
  }),
  names(RC_GSEA_df)
)

## save data for GSEA
for (i in seq_along(intestine_wide_GSEA_list)) {
  filename <- paste0("../output/gsea_input/intestine_wide_GSEA_", names(intestine_wide_GSEA_list)[i], ".rnk")
  write.table(intestine_wide_GSEA_list[[i]], filename, sep = "\t", col.names = FALSE, row.names = FALSE)
}

for (i in seq_along(TI_GSEA_list)) {
  filename <- paste0("../output/gsea_input/TI_GSEA_", names(TI_GSEA_list)[i], ".rnk")
  write.table(TI_GSEA_list[[i]], filename, sep = "\t", col.names = FALSE, row.names = FALSE)
}

for (i in seq_along(C_GSEA_list)) {
  filename <- paste0("../output/gsea_input/C_GSEA_", names(C_GSEA_list)[i], ".rnk")
  write.table(C_GSEA_list[[i]], filename, sep = "\t", col.names = FALSE, row.names = FALSE)
}

for (i in seq_along(RC_GSEA_list)) {
  filename <- paste0("../output/gsea_input/RC_GSEA_", names(RC_GSEA_list)[i], ".rnk")
  write.table(RC_GSEA_list[[i]], filename, sep = "\t", col.names = FALSE, row.names = FALSE)
}
```

## format GSEA output
```{r}
GSEA_output_formatted <- GSEA_output

GSEA_output_formatted$region <- sub("_.*", "", GSEA_output_formatted$parent_dir)
GSEA_output_formatted$microbe <- sub(".*GSEA_", "", GSEA_output_formatted$parent_dir) 
GSEA_output_formatted$pathway <- sub(".*HALLMARK_", "", GSEA_output_formatted$file_name)

GSEA_output_formatted$levels <- paste0(GSEA_output_formatted$microbe, "_", GSEA_output_formatted$pathway)

# how many genes per microbe/pathway
GSEA_output_count <- GSEA_output_formatted %>% 
  group_by(levels) %>%
  dplyr::count()

GSEA_output_path <- GSEA_output_formatted %>% 
  group_by(pathway) %>%
  dplyr::count()

GSEA_output_region <- GSEA_output_formatted %>% 
  group_by(region) %>%
  dplyr::count()

GSEA_output_microbe <- GSEA_output_formatted %>% 
  group_by(microbe) %>%
  dplyr::count()

GSEA_output_count
GSEA_output_path
GSEA_output_region
GSEA_output_microbe
```

## GSEA LM
```{r}
# GSEA_output_formatted_core <- filter(GSEA_output_formatted, CORE.ENRICHMENT == "Yes")

GSEA_output_formatted_TI <- filter(GSEA_output_formatted, region == "TI")
GSEA_output_formatted_C <- filter(GSEA_output_formatted, region == "C")
GSEA_output_formatted_RC <- filter(GSEA_output_formatted, region == "RC")

# TI
for (i in 1:nrow(GSEA_output_formatted_TI)) {
  g <- GSEA_output_formatted_TI$SYMBOL[i]
  m <- GSEA_output_formatted_TI$microbe[i]
  
  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = TI_df)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  GSEA_output_formatted_TI$pval[i] <- lm_p
}

GSEA_output_formatted_TI_BH <- GSEA_output_formatted_TI
GSEA_output_formatted_TI_BH$pval <- p.adjust(GSEA_output_formatted_TI_BH$pval, method = "BH")


# C
for (i in 1:nrow(GSEA_output_formatted_C)) {
  g <- GSEA_output_formatted_C$SYMBOL[i]
  m <- GSEA_output_formatted_C$microbe[i]
  
  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = C_df)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  GSEA_output_formatted_C$pval[i] <- lm_p
}

GSEA_output_formatted_C_BH <- GSEA_output_formatted_C
GSEA_output_formatted_C_BH$pval <- p.adjust(GSEA_output_formatted_C_BH$pval, method = "BH")


# RC
for (i in 1:nrow(GSEA_output_formatted_RC)) {
  g <- GSEA_output_formatted_RC$SYMBOL[i]
  m <- GSEA_output_formatted_RC$microbe[i]
  
  # Fit linear model
  fit <- lm(as.formula(get(m) ~ get(g)), data = RC_df)

  # Extract p-value of the microbe coefficient
  lm_p <- summary(fit)$coefficients[,"Pr(>|t|)"][[2]]  # coefficient p-values column

  GSEA_output_formatted_RC$pval[i] <- lm_p
}

GSEA_output_formatted_RC_BH <- GSEA_output_formatted_RC
GSEA_output_formatted_RC_BH$pval <- p.adjust(GSEA_output_formatted_RC_BH$pval, method = "BH")

# correct for multiple tests overall
GSEA_output_formatted_all <- rbind(GSEA_output_formatted_RC, GSEA_output_formatted_C)
GSEA_output_formatted_all <- rbind(GSEA_output_formatted_all, GSEA_output_formatted_TI)

GSEA_output_formatted_all_BH <- GSEA_output_formatted_all
GSEA_output_formatted_all_BH$pval <- p.adjust(GSEA_output_formatted_all_BH$pval, method = "BH")

ggplot(RC_df, aes(x = 	SYT12, y = Klebsiella)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "NPM2 and Christensenellaceae_R.7_group in C")

ggplot(C_df, aes(x = 	HBB, y = Peptostreptococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "NPM2 and Christensenellaceae_R.7_group in C")

test <- lm(Peptostreptococcus ~ HBB, data = C_df)
summary(test)

ggplot(C_df, aes(x = 	TPRKB, y = Pseudomonas)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(title = "NPM2 and Christensenellaceae_R.7_group in C")

```

# save GSEA output
```{r}
GSEA_output_formatted_save <- rbind(GSEA_output_formatted_C_BH, GSEA_output_formatted_RC_BH)
GSEA_output_formatted_save <- rbind(GSEA_output_formatted_save, GSEA_output_formatted_TI_BH)

# write.csv(GSEA_output_formatted_save, "../output/GSEA_output_formatted_tab.csv")

# write.csv(GSEA_output_formatted_all, "../output/GSEA_output_formatted_all.csv")
```

## plot associations of interest
```{r}
# TI
lacto_acat1_TI_plot <- ggplot(TI_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_TI_plot <- ggplot(TI_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

# C
lacto_acat1_C_plot <- ggplot(C_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_C_plot <- ggplot(C_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

# RC
lacto_acat1_RC_plot <- ggplot(RC_df, aes(x = ACAT1, y = Lactococcus)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

faecal_pcdh20_RC_plot <- ggplot(RC_df, aes(x = PCDH20, y = Faecalitalea)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x) +
  labs(x = NULL, y = NULL)

## correlation values
cor.test(TI_df$ACAT1, TI_df$Lactococcus, method = c("pearson"))
cor.test(TI_df$PCDH20, TI_df$Faecalitalea, method = c("pearson"))
cor.test(C_df$ACAT1, C_df$Lactococcus, method = c("pearson"))
cor.test(C_df$PCDH20, C_df$Faecalitalea, method = c("pearson"))
cor.test(RC_df$ACAT1, RC_df$Lactococcus, method = c("pearson"))
cor.test(RC_df$PCDH20, RC_df$Faecalitalea, method = c("pearson"))

## save
ggsave(file="../figures/lacto_acat1_TI_plot.pdf", plot=lacto_acat1_TI_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_TI_plot.pdf", plot=faecal_pcdh20_TI_plot, width=2.5, height=2)
ggsave(file="../figures/lacto_acat1_C_plot.pdf", plot=lacto_acat1_C_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_C_plot.pdf", plot=faecal_pcdh20_C_plot, width=2.5, height=2)
ggsave(file="../figures/lacto_acat1_RC_plot.pdf", plot=lacto_acat1_RC_plot, width=2.5, height=2)
ggsave(file="../figures/faecal_pcdh20_RC_plot.pdf", plot=faecal_pcdh20_RC_plot, width=2.5, height=2)
```
