---
title: "full_deseq2"
author: "Erica Ryu"
date: "7/23/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 4.Covariate Correction

## The purpose of this script is to correct for technical and biological covariates

## load packages
```{r}
library(tibble)
library(factoextra)
library(lme4)
library(lmtest)
library(dplyr)
library(ggplot2)
library(car)
library(DESeq2)
```

## import data
```{r}
counts <- read.table("../../1.read_processing/output/3.feature_counts/reverse_counts.txt", header = TRUE)
metadata <- read.csv("../data/RNA_metadata.csv", header = TRUE)

# subset to sequenced libraries first
metadata_lp <- metadata[!is.na(metadata$s.SequencingBatch), ]
```

## format count data
```{r}
counts_clean <- counts
# clean up count columns
for ( col in 1:ncol(counts_clean)){
    colnames(counts_clean)[col] <-  sub("_S.*", "", colnames(counts_clean)[col])
}

for ( col in 1:ncol(counts_clean)){
    colnames(counts_clean)[col] <-  sub("^output.2.star.", "", colnames(counts_clean)[col])
}

counts_clean_form <- column_to_rownames(counts_clean, var = "Geneid")

counts_clean_form <- counts_clean_form[!names(counts_clean_form) %in% c("Chr", "Start", "End", "Strand", "Length")]

# filter out genes with zero counts
counts_clean_form <- counts_clean_form[rowSums(counts_clean_form) != 0, ]

## make batches into factors
metadata_lp$s.SequencingBatch <- as.factor(metadata_lp$s.SequencingBatch)

```

## save uncorrected
```{r}
write.csv(counts_clean_form, file = "../output/counts_uncorr.csv")
```

## transform counts
```{r}
set.seed(123)
dds <- DESeqDataSetFromMatrix(countData = counts_clean_form, colData = metadata_lp, design = ~ i.Disease + i.Specimen)

# filter low expression counts
smallestGroupSize <- 15
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

# transform
set.seed(123)
vsd <- vst(dds, blind=FALSE)
mat <- assay(vsd)
```

## save normalized
```{r}
write.csv(mat, file = "../output/counts_normal_uncorr.csv")
```

## generate PCs
```{r}
counts_pc <- mat
counts_pc_t <- t(counts_pc)

set.seed(123)
pca <- prcomp(counts_pc_t)
pca_scores <- as.data.frame(pca$x)
eig.val <- get_eigenvalue(pca)
```

## check for significant associations
```{r}
# filter to PCs that explain at least 90% of variation
pca_scores_PC62 <- 1:62
pca_scores_PC62 <- paste0('PC', pca_scores_PC62)

pca_scores_vectors <- pca_scores[, pca_scores_PC62]

# remove metadata columns with NA
metadata_noNA <- metadata_lp

metadata_noNA[metadata_noNA==""] <- NA

metadata_noNA <- metadata_noNA[ , colSums(is.na(metadata_noNA))==0]

metadata_noNA <- metadata_noNA[order(metadata_noNA$p.LibraryID, decreasing = FALSE),]

# check order
identical(rownames(pca_scores_vectors), metadata_noNA$p.LibraryID)

# set up df for check
df_comb <- cbind(pca_scores_vectors, metadata_noNA)

# check factor
df_comb$i.Sex <- as.factor(df_comb$i.Sex)
df_comb$i.Age <- as.numeric(df_comb$i.Age)
df_comb$i.OR.Age <- as.numeric(df_comb$i.OR.Age)
df_comb$p.LibraryConcentration_ngul <- as.numeric(df_comb$p.LibraryConcentration_ngul)
df_comb$ex.RINscore <- as.numeric(df_comb$ex.RINscore)
df_comb$ex.Extractedby <- as.factor(df_comb$ex.Extractedby)
df_comb$s.SequencingBatch <- as.factor(df_comb$s.SequencingBatch)
df_comb$ex.ExtractionBatch <- as.factor(df_comb$ex.ExtractionBatch)
df_comb$p.LibraryBatch <- as.factor(df_comb$p.LibraryBatch)
df_comb$i.PatientID <- as.factor(df_comb$i.PatientID)

# check for associations
df_associations <- matrix(ncol = 9, nrow = ncol(pca_scores_vectors))
colnames(df_associations) <- c("i.Sex", "i.Age", "i.OR.Age", "p.LibraryConcentration_ngul", "ex.RINscore", "ex.Extractedby", "ex.ExtractionBatch", "p.LibraryBatch", "s.SequencingBatch")
rownames(df_associations) <- colnames(pca_scores_vectors)

set.seed(123)
for(i in colnames(df_associations)){
  for (j in rownames(df_associations)) {
    print(as.formula(get(j) ~ get(i)))
    obj <- print(lm(as.formula(get(j) ~ get(i)), data = df_comb))
    df_associations[j,i] <- summary(obj)$coefficients[,"Pr(>|t|)"][[2]]
  }
}

# multiple test correction
df_associations_BH <- as.data.frame(t(df_associations))

df_associations_BH <- sapply(df_associations_BH, p.adjust, method="BH", n = 9)
df_associations_BH <- t(df_associations_BH)
colnames(df_associations_BH) <- colnames(df_associations)
```

## save covariate dfs
```{r}
write.csv(df_associations_BH, "../output/covariate_rna.csv")
```

## plot covariate associations
```{r}
df_associations_BH_df <- as.data.frame(df_associations_BH)

df_associations_BH_df$axis <- rownames(df_associations_BH_df)

## convert format
df_associations_BH_long <- pivot_longer(df_associations_BH_df, !axis, names_to = "covariate", values_to = "padj")

## order by axis
df_associations_BH_long <- mutate(df_associations_BH_long, axis = factor(axis, levels = unique(rownames(df_associations_BH_df))))

ggplot(df_associations_BH_long, aes(x = covariate, y = axis, fill = padj)) +
  geom_tile() +
  scale_fill_gradient(high="white", low="#009194", limits=c(0,1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  labs(x = "Covariate", y = "Axis")
```

## check for random effects
```{r}
# patient ID
## set up vector
patient_rand_vec <- c()
n = 1

set.seed(123)
for(i in colnames(pca_scores_vectors)){
  full_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.LibraryConcentration_ngul + ex.Extractedby + ex.RINscore + s.SequencingBatch + (1 | i.PatientID)")), data = df_comb)
  reduced_model <- lm(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.LibraryConcentration_ngul + ex.Extractedby + ex.RINscore + s.SequencingBatch")), data = df_comb)
  lr <- lrtest(reduced_model, full_model)
  patient_rand_vec <- c(patient_rand_vec, lr["2", "Pr(>Chisq)"])
  n = n + 1
  print(n)
}

df_cov <- as.data.frame(df_associations_BH)

df_cov$patient_rand <- patient_rand_vec

## add variance explained
df_cov$perc_contrib <- eig.val[1:62, "variance.percent"]

## reorder
df_cov <- dplyr::select(df_cov, perc_contrib, everything())

```

## check for collinearity with variables of interest
```{r}
# set up df
collinear <- matrix(ncol = 2, nrow = ncol(pca_scores_vectors))
colnames(collinear) <- c("i.Specimen", "i.Disease")
rownames(collinear) <- colnames(pca_scores_vectors)
n = 1

# run
set.seed(123)
for(i in colnames(pca_scores_vectors)){
  spec_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Specimen + i.Sex + i.Age + i.OR.Age + p.LibraryConcentration_ngul + ex.Extractedby + s.SequencingBatch + ex.RINscore + (1 | i.PatientID)")), data = df_comb)
  vif_spec <- vif(spec_model)
  collinear[i, "i.Specimen"] <- vif_spec["i.Specimen", "GVIF^(1/(2*Df))"]
  disease_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Disease + i.Sex + i.Age + i.OR.Age + p.LibraryConcentration_ngul + ex.Extractedby + s.SequencingBatch + ex.RINscore + (1 | i.PatientID)")), data = df_comb)
  vif_disease <- vif(disease_model)
  collinear[i, "i.Disease"] <- vif_disease["i.Disease", "GVIF^(1/(2*Df))"]
  print(n)
  n = n + 1
}

# and the variables in general
cov_collinear <- matrix(nrow = 7, ncol = ncol(pca_scores_vectors))
rownames(cov_collinear) <- c("i.Sex", "i.Age", "i.OR.Age", "p.LibraryConcentration_ngul", "ex.Extractedby", "s.SequencingBatch", "ex.RINscore")
colnames(cov_collinear) <- colnames(pca_scores_vectors)

n = 1

set.seed(123)
for(i in colnames(pca_scores_vectors)){
  cov_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.LibraryConcentration_ngul + ex.Extractedby + s.SequencingBatch + ex.RINscore + (1 | i.PatientID)")), data = df_comb)
  vif_cov <- vif(cov_model)
  cov_collinear[,i] <- vif_cov[,3]
  print(n)
  n = n + 1
}

## redo without possible problematic covariates
cov_collinear_up <- matrix(nrow = 6, ncol = ncol(pca_scores_vectors))
rownames(cov_collinear_up) <- c("i.Sex", "i.Age", "p.LibraryConcentration_ngul", "ex.Extractedby", "ex.RINscore", "s.SequencingBatch")
colnames(cov_collinear_up) <- colnames(pca_scores_vectors)

n = 1

set.seed(123)
for(i in colnames(pca_scores_vectors)){
  cov_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.LibraryConcentration_ngul + ex.Extractedby + ex.RINscore + s.SequencingBatch + (1 | i.PatientID)")), data = df_comb)
  vif_cov <- vif(cov_model)
  cov_collinear_up[,i] <- vif_cov[,3]
  print(n)
  n = n + 1
}
```

## regress out
```{r}
counts_regress_prep <- as.data.frame(counts_pc_t)

# make empty dataframe
counts_regress <- matrix(ncol = ncol(counts_regress_prep), nrow = nrow(counts_regress_prep))

colnames(counts_regress) <- colnames(counts_regress_prep)
rownames(counts_regress) <- rownames(counts_regress_prep)

# order metadata
metadata_corr <- metadata_noNA
metadata_corr <- metadata_corr[order(metadata_corr$p.LibraryID),]
# check if lined up
identical(rownames(counts_regress_prep), metadata_corr$p.LibraryID)
# add variables
counts_regress_prep$sex <- as.factor(metadata_corr$i.Sex)
counts_regress_prep$age <- as.numeric(metadata_corr$i.Age)
counts_regress_prep$lpconc <- as.numeric(metadata_corr$p.LibraryConcentration_ngul)
counts_regress_prep$extractor <- as.factor(metadata_corr$ex.Extractedby)
counts_regress_prep$RINscore <- as.numeric(metadata_corr$ex.RINscore)
counts_regress_prep$sbatch <- as.factor(metadata_corr$s.SequencingBatch)
counts_regress_prep$patientID <- as.factor(metadata_corr$i.PatientID)

# run
n = 1

set.seed(123)
for(i in colnames(counts_regress)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ sex + age + lpconc + extractor + RINscore + sbatch + (1 | patientID)")),  data=counts_regress_prep, control = lmerControl(optCtrl = list(maxfun = 1000000)))
  counts_regress[, i] <- resid(obj)
  n = n + 1
  print(n)
}
```

## check if OR.age correlated with removed covariates
```{r}
counts_check <- counts_regress
set.seed(123)
pca_check <- prcomp(counts_check)
pca_check_scores <- as.data.frame(pca_check$x)
eig.val_check <- get_eigenvalue(pca_check)

# filter to PCs that explain at least 90% of variation
pca_check_scores_PC40 <- pca_check_scores[, 1:40]

# check order
identical(rownames(pca_check_scores_PC40), metadata_noNA$p.LibraryID)

# set up df for check
df_comb_check <- cbind(pca_check_scores_PC40, metadata_noNA)

# check for associations
cov_check <- data.frame(matrix(ncol = 1, nrow = 40))
rownames(cov_check) <- colnames(pca_check_scores_PC40)
colnames(cov_check) <- c("i.OR.Age")

set.seed(123)
for(i in colnames(cov_check)){
  for (j in rownames(cov_check)) {
    print(as.formula(get(j) ~ get(i)))
    obj <- print(lm(as.formula(get(j) ~ get(i)), data = df_comb_check))
    cov_check[j,i] <- summary(obj)$coefficients[,"Pr(>|t|)"][[2]]
  }
}
```


