---
title: "qiime_to_phyloseq"
author: "Erica Ryu"
date: "7/27/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 7. Remove cross contaminants

## The purpose is to benchmark the samples and then filter based on potential cross contamination

## load packages
```{r}
library(ggplot2)
library(phyloseq)
library(vegan)
library(dplyr)
library(FSA)
```

## load in objects
```{r}
phyloseq <- readRDS("../output/Hershey_ps.rds")
```

## load functions
```{r}
# calculate beta diversity
beta_ordinate <- function(physeq, beta_dist){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  ## calculate distance and ordinate
  if (beta_dist == "unweighted unifrac") {
    dist <- unifrac(beta_inp, weighted = FALSE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else if (beta_dist == "weighted unifrac") {
    dist <- unifrac(beta_inp, weighted = TRUE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else {
    ord.pcoa <- ordinate(ps.prop, method = "PCoA", distance = "bray", weighted = FALSE)
  }
}

# calculate permanova
permanova <- function(physeq, beta_dist, perm_var){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  df <- data.frame(sample_data(ps.prop))
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  if (beta_dist == "unweighted unifrac") {
    dist <- unifrac(beta_inp, weighted = FALSE, tree = ps.prop@phy_tree)
  } else if (beta_dist == "weighted unifrac") {
    dist <- unifrac(beta_inp, weighted = TRUE, tree = ps.prop@phy_tree)
  } else {
    dist = phyloseq::distance(ps.prop, method="bray", normalized=TRUE, parallel=FALSE, fast=TRUE) 
  }
  perm_var <- df[[perm_var]]
   perm <- adonis2(dist ~ perm_var, permutations = 99999)
   print(perm)
}

# calculate beta_disper
beta_disper_test <- function(physeq, disper_var){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  df <- data.frame(sample_data(ps.prop))
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  dist = phyloseq::distance(ps.prop, method="bray", normalized=TRUE, parallel=FALSE, fast=TRUE) 
  disper_var <- df[[disper_var]]
  disper <- betadisper(dist, disper_var)
  anova(disper)
}
```

## set colors
```{r}
threecolors <- c("#1E88E5", "#F0E442", "#D55E00")
twocolors <- c("#332288", "#52BD0C")
```

## check DNA conc
```{r}
df_plot <- as.data.frame(sample_data(phyloseq))
df_plot$sample_type_by_exp <- ifelse(grepl("PCRNC", df_plot$p.SourceID) & df_plot$p.SampleType == "first_NC", "PCR_first_NC", 
                                     ifelse(grepl("QNC", df_plot$p.SourceID) & df_plot$p.SampleType == "first_NC", "extract_first_NC",
                                     ifelse(grepl("QNC", df_plot$p.SourceID) & df_plot$p.SampleType == "middle_NC", "extract_mid_NC",
                                     ifelse(grepl("QNC", df_plot$p.SourceID) & df_plot$p.SampleType == "last_NC", "extract_last_NC",
                                     ifelse(grepl("PCRNC", df_plot$p.SourceID) & df_plot$p.SampleType == "last_NC", "PCR_last_NC",
                                     ifelse(df_plot$p.SampleType == "PC", "PC", "extraction_product"))))))

df_plot$sample_type_by_exp <- factor(df_plot$sample_type_by_exp, levels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"))

### show DNA conc
ggplot(df_plot, aes(x = sample_type_by_exp, y = p.CleanedPCRConcentration, group = sample_type_by_exp), color = black) +
  geom_violin(aes(fill=sample_type_by_exp)) +
  geom_boxplot(width=0.2, color="black", alpha=0.2, outlier.shape = NA) +
  scale_fill_manual(name=NULL,
                     values=c("red", "lime green", "cyan", "slate blue", "blue", "forest green", "gray"),
                    breaks=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"),
                    labels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"))+
  geom_jitter(size = 1, width=0.15, col="black", alpha = 0.8)+
  labs(title = "Comparing final DNA concentrations", y = "DNA concentration ng/ul", x = "Sample Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  theme(legend.position = "none")

# log transformed
control_conc <- ggplot(df_plot, aes(x = sample_type_by_exp, y = p.CleanedPCRConcentration, group = sample_type_by_exp), color = black) +
  geom_violin(aes(fill=sample_type_by_exp)) +
  geom_boxplot(width=0.2, color="black", alpha=0.2, outlier.shape = NA) +
  scale_fill_manual(name=NULL,
                     values=c("red", "lime green", "cyan", "slate blue", "blue", "forest green", "gray"),
                    breaks=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"),
                    labels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"))+
  geom_jitter(size = 1, width=0.15, col="black", alpha = 0.8)+
  labs(title = "Comparing final DNA concentrations", y = "Log transformed DNA concentration ng/ul", x = "Sample Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  theme(legend.position = "none") +
  scale_y_log10()

ggsave(file = "../figures/control_conc.pdf", plot = control_conc, width = 7, height = 5)

kruskal.test(p.CleanedPCRConcentration ~ sample_type_by_exp, data = as.data.frame(as.matrix(df_plot)))

sample_comparison <- dunnTest(as.numeric(p.CleanedPCRConcentration) ~ sample_type_by_exp, data = as.data.frame(as.matrix(df_plot)), method="bh")

df_sample_comparison <- sample_comparison[["res"]]
```

## beta diversity
```{r}
# add sample type info
sample_data(phyloseq)$sample_type_by_exp <- ifelse(grepl("PCRNC", phyloseq@sam_data$p.SourceID) & phyloseq@sam_data$p.SampleType == "first_NC", "PCR_first_NC", 
                                     ifelse(grepl("QNC", phyloseq@sam_data$p.SourceID) & phyloseq@sam_data$p.SampleType == "first_NC", "extract_first_NC",
                                     ifelse(grepl("QNC", phyloseq@sam_data$p.SourceID) & phyloseq@sam_data$p.SampleType == "middle_NC", "extract_mid_NC",
                                     ifelse(grepl("QNC", phyloseq@sam_data$p.SourceID) & phyloseq@sam_data$p.SampleType == "last_NC", "extract_last_NC",
                                     ifelse(grepl("PCRNC", phyloseq@sam_data$p.SourceID) & phyloseq@sam_data$p.SampleType == "last_NC", "PCR_last_NC",
                                     ifelse(phyloseq@sam_data$p.SampleType == "PC", "PC", "extraction_product"))))))

# samples and controls
set.seed(123)
bray_ordinate <- beta_ordinate(phyloseq, beta_dist = "bray")

## extract axes and format
PCOAaxes <- bray_ordinate$vectors[,c(1,2,3,4)]
sample_type_by_exp <- phyloseq@sam_data$sample_type_by_exp
PCOAaxes_meta <- cbind(PCOAaxes, sample_type_by_exp)
# check if lined up
identical(rownames(PCOAaxes), rownames(phyloseq@sam_data))
df_PCOA <- as.data.frame(as.matrix(PCOAaxes_meta))

sample_w_control <- ggplot(df_PCOA, aes(x = as.numeric(Axis.1), y = as.numeric(Axis.2))) +
  geom_point(shape = 21, color = "black", size = 3, aes(fill = sample_type_by_exp)) +
  scale_x_continuous(breaks=seq(-1,1,0.1)) +
  scale_y_continuous(breaks=seq(-1,1,0.1)) +
  scale_fill_manual(name=NULL,
                     values=c("red", "lime green", "cyan", "slate blue", "blue", "forest green", "gray"),
                    breaks=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"),
                    labels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC")) +
  labs(x = "PCoA1 (11.3%)",
       y = "PCoA2 (9.57%)", 
       title = "Sample Type")

ggsave(file="../figures/sample_w_control.pdf", plot=sample_w_control, width=6, height=4)

set.seed(123)
permanova(phyloseq, beta_dist = "bray", perm_var = "sample_type_by_exp")


wo_PC <- subset_samples(phyloseq, p.SampleType != "PC")

set.seed(123)
permanova(wo_PC, beta_dist = "bray", perm_var = "sample_type_by_exp")
```

## get sample counts before/after DNA conc removal
```{r}
samp_ps <- subset_samples(phyloseq, p.SampleType == "sample")

samp_df <- as.data.frame(as.matrix(samp_ps@sam_data))

# initial sample count by batch, location, disease
samp_df %>% dplyr::count(ex.ExtractionBatch, sort = TRUE)

samp_df %>% dplyr::count(i.Disease, sort = TRUE)

samp_df %>% dplyr::count(i.Specimen, sort = TRUE)

# subset to samples above highest neg control within batch
wo_PC_df <- as.data.frame(as.matrix(wo_PC@sam_data))
wo_PC_df$Sample_or_Control <- ifelse(wo_PC_df$p.SampleType == "sample", "sample", "control" )

df_by_batch <- split(wo_PC_df, with(wo_PC_df, p.PCRBatch))

## generate vector of max control concentration
max_control_vector <- c()

for(i in 1:length(df_by_batch)){
  df <- subset(df_by_batch[[i]], Sample_or_Control == "control")
  max_control_vector = c(max_control_vector, max(df$p.CleanedPCRConcentration))
}

## get samples above each batch control's max conc
df_samp_fin <- wo_PC_df[FALSE,]

for(i in 1:length(df_by_batch)){
  df_samp <- subset(df_by_batch[[i]], Sample_or_Control == "sample")
  df_samp_set <- subset(df_samp, p.CleanedPCRConcentration > max_control_vector[i])
  df_samp_fin <- rbind(df_samp_set, df_samp_fin)
}

df_samp_fin %>% dplyr::count(ex.ExtractionBatch, sort = TRUE)

df_samp_fin %>% dplyr::count(i.Disease, sort = TRUE)

df_samp_fin %>% dplyr::count(i.Specimen, sort = TRUE)

# keep samples more than 0.5mg
df_samp_fin2 <- subset(df_samp_fin, as.numeric(ex.TissueWeight_mg) > 0.5)
```

## get samples that were removed and remove from original ps object
```{r}
samp_ret <- intersect(df_samp_fin2$p.PCRId, samp_df$p.PCRId)
diff <- samp_df$p.PCRId[!(samp_df$p.PCRId %in% samp_ret)]
diff_vec <- diff

samp_ret_ps <- subset_samples(phyloseq, !(p.PCRId %in% diff_vec))
```

## check controls and samples
```{r}
# samples and controls
set.seed(123)
bray_ordinate_ret <- beta_ordinate(samp_ret_ps, beta_dist = "bray")

## extract axes and format
PCOAaxes_ret <- bray_ordinate_ret$vectors[,c(1,2,3,4)]
sample_type_by_exp_ret <- samp_ret_ps@sam_data$sample_type_by_exp
PCOAaxes_meta_ret <- cbind(PCOAaxes_ret, sample_type_by_exp_ret)
# check if lined up
identical(rownames(PCOAaxes_ret), rownames(samp_ret_ps@sam_data))
df_PCOA_ret <- as.data.frame(as.matrix(PCOAaxes_meta_ret))

sample_after_cross_filt <- ggplot(df_PCOA_ret, aes(x = as.numeric(Axis.1), y = as.numeric(Axis.2))) +
  geom_point(shape = 21, color = "black", size = 3, aes(fill = sample_type_by_exp_ret)) +
  scale_x_continuous(breaks=seq(-1,1,0.1)) +
  scale_y_continuous(breaks=seq(-1,1,0.1)) +
  scale_fill_manual(name=NULL,
                     values=c("red", "lime green", "cyan", "slate blue", "blue", "forest green", "gray"),
                    breaks=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"),
                    labels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC")) +
  labs(x = "PCoA1 (11.52%)",
       y = "PCoA2 (9.58%)", 
       title = "Sample Type")

ggsave(file = "../figures/sample_after_cross_contam_filt.pdf", plot = sample_after_cross_filt, width = 7, height = 5)
```

## save pruned phyloseq object
```{r}
saveRDS(samp_ret_ps, "../output/samp_ret_ps.rds")
```

