---
title: "decontam"
author: "Erica Ryu"
date: "7/18/2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 8. Remove and analyse contaminants

## The purpose of this script is to remove and analyse host and environmental contaminants

## load packages
```{r}
library(phyloseq)
library(decontam)
library(ggplot2)
library(data.table)
library(dplyr)
library(vegan)
library(plyr)
library(tidyr)
```

## load in objects
```{r}
phyloseq <- readRDS("../output/samp_ret_ps.rds")

# host contamination data
raw_name <- read.table("../../1.qc_host_decontam/output/5.read_count/raw_filename")
raw_reads <- read.table("../../1.qc_host_decontam/output/5.read_count/raw_read_count")
bowtie_unmapped_name <- read.table("../../1.qc_host_decontam/output/5.read_count/bowtie_unmapped_filename")
bowtie_unmapped_reads <- read.table("../../1.qc_host_decontam/output/5.read_count/bowtie_unmapped_read_count")
snap_unmapped_name <- read.table("../../1.qc_host_decontam/output/5.read_count/snap_unmapped_filename")
snap_unmapped_reads <- read.table("../../1.qc_host_decontam/output/5.read_count/snap_unmapped_read_count")
```

## load functions
```{r}
# calculate beta diversity
beta_ordinate <- function(physeq, beta_dist){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  ## calculate distance and ordinate
  if (beta_dist == "unweighted unifrac") {
    dist <- rbiom::unifrac(beta_inp, weighted = FALSE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else if (beta_dist == "weighted unifrac") {
    dist <- rbiom::unifrac(beta_inp, weighted = TRUE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else {
    ord.pcoa <- ordinate(ps.prop, method = "PCoA", distance = "bray", weighted = FALSE)
  }
}

# calculate permanova
permanova <- function(physeq, beta_dist){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  df <- data.frame(sample_data(ps.prop))
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  if (beta_dist == "unweighted unifrac") {
    dist <- unifrac(beta_inp, weighted = FALSE, tree = ps.prop@phy_tree)
  } else if (beta_dist == "weighted unifrac") {
    dist <- unifrac(beta_inp, weighted = TRUE, tree = ps.prop@phy_tree)
  } else {
    dist = phyloseq::distance(ps.prop, method="bray", normalized=TRUE, parallel=FALSE, fast=TRUE) 
  }
   perm <- adonis2(dist ~ p.SampleType, data = df, permutations = 99999)
   print(perm)
}
```

## host decontamination prep
```{r}
# prep tables
## transpose names
raw_name_col <- t(raw_name)
bowtie_unmapped_name_col <- t(bowtie_unmapped_name)
snap_unmapped_name_col <- t(snap_unmapped_name)

## clean up names
raw_name_col <-gsub("data/raw/","", raw_name_col)
raw_name_col <- gsub(".fastq.gz", "", raw_name_col)

bowtie_unmapped_name_col <- gsub("output/2.bowtie2/3.fastq/unmapped/", "",  bowtie_unmapped_name_col)
bowtie_unmapped_name_col <- gsub("_unmapped", "", bowtie_unmapped_name_col)
bowtie_unmapped_name_col <- gsub(".fastq", "", bowtie_unmapped_name_col)

snap_unmapped_name_col <- gsub("output/3.snap/3.fastq/unmapped/", "", snap_unmapped_name_col)
snap_unmapped_name_col <- gsub("_unmapped", "", snap_unmapped_name_col)
snap_unmapped_name_col <- gsub(".fastq", "", snap_unmapped_name_col)

# collapse into tables
raw <- cbind(raw_name_col, raw_reads)
bowtie_unmapped <- cbind(bowtie_unmapped_name_col, bowtie_unmapped_reads)
snap_unmapped <- cbind(snap_unmapped_name_col, snap_unmapped_reads)

# change column names
colnames(raw) <- c("file", "raw")
colnames(bowtie_unmapped) <- c("file", "bowtie2_unmapped")
colnames(snap_unmapped) <- c("file", "snap_unmapped")

# merge into one table
read_count_unaligned <- merge(raw, bowtie_unmapped, by = "file")
read_count_unaligned <- merge(read_count_unaligned, snap_unmapped, by = "file")
# remove samples with zero reads from the start
read_count_unaligned <- filter(read_count_unaligned, raw > 10)
# remove reverse read repeats
read_count_unaligned <- read_count_unaligned[!grepl("R2", read_count_unaligned$file),]
# fix name
read_count_unaligned$file <- gsub("238_s238__S238_R1", "238_s238_S238_R1",  read_count_unaligned$file)

# clean up names
read_count_unaligned$file <- gsub(".*_s", "s",  read_count_unaligned$file)
read_count_unaligned$file <- gsub("_S.*", "",  read_count_unaligned$file)

# filter to samples only
phyloseq_samp <- subset_samples(phyloseq, p.SampleType == "sample")
read_count_unaligned <- subset(read_count_unaligned, file %in% sample_data(phyloseq_samp)$s.SequencingID)

read_count_unaligned_saved_df <- read_count_unaligned
colnames(read_count_unaligned_saved_df) <- c("sample", "raw", "bowtie2_unmapped", "snap_unmapped")

# write.csv(read_count_unaligned_saved_df, "../output/read_count_unaligned_tab.csv", row.names = FALSE)
```

## host decontamination plot
```{r}
### counts
# prep for plotting
read_count_unaligned_long <- pivot_longer(read_count_unaligned, !file, names_to = "process", values_to = "count")

# set processes as levels
process <- c("raw", "bowtie2_unmapped", "snap_unmapped")
read_count_unaligned_plot <- mutate(read_count_unaligned_long, process = factor(process, levels = unique(process)))

ggplot(read_count_unaligned_plot, aes(x = process, y = count, group = process)) + 
  geom_line() +
  geom_jitter(width=0.15) + 
  labs(title = "Counts of Unaligned Reads", 
       y = "read count")

### unaligned percentages
read_count_unaligned_perc <- read_count_unaligned
columns_to_divide_unmapped <- c("raw", "bowtie2_unmapped", "snap_unmapped")
read_count_unaligned_perc[,columns_to_divide_unmapped] <- read_count_unaligned_perc[, columns_to_divide_unmapped]/read_count_unaligned_perc[,"raw"]

# prep for plotting
read_count_unaligned_long_perc <- pivot_longer(read_count_unaligned_perc, !file, names_to = "process", values_to = "count")

# set processes as levels
read_count_unaligned_plot_perc <- mutate(read_count_unaligned_long_perc, process = factor(process, levels = unique(process)))

ggplot(read_count_unaligned_plot_perc, aes(x = process, y = count, group = process)) +
  geom_line() +
  geom_point() + 
  labs(title = "Proportion of Retained Reads", 
       y = "proportion") + 
  theme(legend.position="none")

host_decontam <- ggplot(read_count_unaligned_plot_perc, aes(x = process, y = count, group = process)) +
  geom_jitter(width=0.15) + 
  labs(title = "Proportion of Retained Reads", 
       y = "proportion") + 
  theme(legend.position="none")

# ggsave(file = "../figures/proportion_retained_reads.pdf", plot = host_decontam, width = 7, height = 5)

# microbiome reads before/after
read_count_unaligned_plot_subset <- filter(read_count_unaligned_plot, process != "bowtie2_unmapped")

unaligned_count <- ggplot(read_count_unaligned_plot_subset, aes(x = process, y = count, group = process)) + 
  geom_jitter(width=0.15) + 
  labs(title = "Counts of Unaligned Reads", 
       y = "read count")

# ggsave(file = "../figures/unaligned_count.pdf", plot = unaligned_count, width = 7, height = 5)

# host contamination reads
read_count_mapped_perc <- read_count_unaligned_perc[, c("snap_unmapped")]
read_count_mapped_perc <- 1 - read_count_mapped_perc

reads_proportion_hist <- ggplot() +
  geom_histogram(aes(read_count_mapped_perc)) + 
  labs(title = "Proportion of Reads that Map", 
       y = "number of samples") + 
  theme(legend.position="none")

# ggsave(file = "../figures/reads_proportion_hist.pdf", plot = reads_proportion_hist, width = 7, height = 5)
```

## remove contaminating taxa using decontam "frequency" method
### In this method, the distribution of the frequency of each sequence feature as a function of the input DNA concentration is used to identify contaminants.
```{r}
set.seed(123)
contamdf.freq <- isContaminant(phyloseq, method="frequency", conc="p.CleanedPCRConcentration")
head(contamdf.freq) 

# number of contaminants
table(contamdf.freq$contaminant)

# FALSE  TRUE 
#  6081   168

# how abundant are the contaminant ASVs?
head(which(contamdf.freq$contaminant))

# plot contaminant ASV vs noncontaminant and save as pdf
# pdf("output/freq_plot.pdf")
plot_frequency(phyloseq, taxa_names(phyloseq)[c(2,865)], conc="p.CleanedPCRConcentration") +
  xlab("DNA Concentration")
# dev.off() 
```

## remove env contaminant taxa
```{r}
phyloseq.noncontam <- prune_taxa(!contamdf.freq$contaminant, phyloseq)
```

## examine which taxa are getting removed
```{r}
# frequency method
tax <- as.data.frame(as.matrix(phyloseq@tax_table))
contaminant_taxa_freq <- filter(contamdf.freq, contaminant == TRUE) # filter to just contaminants
contam_tax_freq <- merge(contaminant_taxa_freq, tax, by = "row.names")
```

## save decontamed object
```{r}
saveRDS(phyloseq.noncontam, file = "../output/Hershey_ps_noncontam.rds")
```

## sanity check
```{r}
# add sample type info
sample_data(phyloseq.noncontam)$sample_type_by_exp <- ifelse(grepl("PCRNC", phyloseq.noncontam@sam_data$p.SourceID) & phyloseq.noncontam@sam_data$p.SampleType == "first_NC", "PCR_first_NC", 
                                     ifelse(grepl("QNC", phyloseq.noncontam@sam_data$p.SourceID) & phyloseq.noncontam@sam_data$p.SampleType == "first_NC", "extract_first_NC",
                                     ifelse(grepl("QNC", phyloseq.noncontam@sam_data$p.SourceID) & phyloseq.noncontam@sam_data$p.SampleType == "middle_NC", "extract_mid_NC",
                                     ifelse(grepl("QNC", phyloseq.noncontam@sam_data$p.SourceID) & phyloseq.noncontam@sam_data$p.SampleType == "last_NC", "extract_last_NC",
                                     ifelse(grepl("PCRNC", phyloseq.noncontam@sam_data$p.SourceID) & phyloseq.noncontam@sam_data$p.SampleType == "last_NC", "PCR_last_NC",
                                     ifelse(phyloseq.noncontam@sam_data$p.SampleType == "PC", "PC", "extraction_product"))))))

# samples and controls
set.seed(123)
bray_ordinate.noncontam <- beta_ordinate(phyloseq.noncontam, beta_dist = "bray")

## extract axes and format
PCOAaxes.noncontam  <- bray_ordinate.noncontam $vectors[,c(1,2,3,4)]
sample_type_by_exp <- phyloseq.noncontam@sam_data$sample_type_by_exp
PCOAaxes_meta.noncontam <- cbind(PCOAaxes.noncontam, sample_type_by_exp)
# check if lined up
identical(rownames(phyloseq.noncontam@sam_data), rownames(PCOAaxes.noncontam))
df_PCOA.noncontam <- as.data.frame(as.matrix(PCOAaxes_meta.noncontam))

neg_v_samp <- ggplot(df_PCOA.noncontam, aes(x = as.numeric(Axis.1), y = as.numeric(Axis.2))) +
  geom_point(shape = 21, color = "black", size = 3, aes(fill = sample_type_by_exp)) +
  scale_x_continuous(breaks=seq(-1,1,0.1)) +
  scale_y_continuous(breaks=seq(-1,1,0.1)) +
  scale_fill_manual(name=NULL,
                     values=c("red", "lime green", "cyan", "slate blue", "blue", "forest green", "gray"),
                    breaks=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC"),
                    labels=c("extraction_product", "PCR_first_NC", "extract_first_NC", "extract_mid_NC", "extract_last_NC", "PCR_last_NC", "PC")) +
  labs(x = "PCoA1 (10.6%)",
       y = "PCoA2 (6.14%)")

ggsave(file = "../figures/neg_v_samp.pdf", plot = neg_v_samp, width = 7, height = 5)

set.seed(123)
permanova(phyloseq.noncontam, beta_dist = "bray")


wo_PC <- subset_samples(phyloseq.noncontam, p.SampleType != "PC")

set.seed(123)
permanova(wo_PC, beta_dist = "bray")
```