---
title: "9.0.additional_QC"
author: "Erica Ryu"
date: "8/30/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 9. Additional QC

## The purpose of this script is to conduct additional QC of the microbiome data and check covariates

## load packages
```{r}
library(phyloseq)
library(data.table)
# library(sva)
library(picante)
library(dplyr)
library(ggplot2)
library(lme4)
library(ape)
library(lmtest)
library(car)
library(tidyr)
```

## load in objects
```{r}
phyloseq <- readRDS("../output/Hershey_ps_noncontam.rds")
NOD2 <- read.csv("../data/TI_NOD2_samples_withSNPcalls.csv", header = TRUE)
```

## load functions
```{r}
# calculate beta diversity
beta_ordinate <- function(physeq, beta_dist){
  # set up data
  ps.prop <- transform_sample_counts(physeq, function(otu) otu/sum(otu))
  ps.prop@otu_table <- na.omit(ps.prop@otu_table)
  beta_inp <- t(as(otu_table(ps.prop), "matrix"))
  ## calculate distance and ordinate
  if (beta_dist == "unweighted unifrac") {
    dist <- rbiom::unifrac(beta_inp, weighted = FALSE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else if (beta_dist == "weighted unifrac") {
    dist <- rbiom::unifrac(beta_inp, weighted = TRUE, tree = ps.prop@phy_tree)
    return(pcoa(dist))
  } else {
    ord.pcoa <- ordinate(ps.prop, method = "PCoA", distance = "bray", weighted = FALSE)
  }
}
```

## set colors
```{r}
threecolors <- c("#1E88E5", "#D55E00", "#CC79A7")
twocolors <- c("#332288", "#52BD0C")
```

## remove controls
```{r}
phyloseq <- subset_samples(phyloseq, p.SampleType == "sample")
```

## prevalence filtering
```{r}
phyloseq_prunetaxa <- filter_taxa(phyloseq, function(x) sum(x > 5) >= 2, TRUE)
```

## remove chloroplasts
```{r}
phyloseq_prunetaxa <- subset_taxa(phyloseq_prunetaxa, Order!="Chloroplast")
```

## examine read depth across all samples
```{r}
reads <- data.table(as(sample_data(phyloseq_prunetaxa), "data.frame"),
                 TotalReads = sample_sums(phyloseq_prunetaxa), keep.rownames = TRUE)

reads_specimen <- subset(reads, i.Disease == "Crohns")
reads_disease <- subset(reads, i.Specimen == "Right Colon")

spec_readdepth <- ggplot(reads_specimen, aes(x = reorder(s.SequencingID, TotalReads), y=TotalReads, fill=i.Specimen)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 65466, linetype="dashed") +
  scale_fill_manual(name=NULL,
                    values=threecolors,
                    breaks=c("Terminal Ileum", "Cecum", "Right Colon"),
                    labels=c("Terminal Ileum", "Cecum", "Right Colon")) +
  labs(title = "Specimen read counts",
       x = "samples", 
       y = "reads") +
  # ylim(0,95000)+
  theme(axis.text.x = element_blank())

ggsave(file = "../figures/spec_reads.pdf", plot = spec_readdepth, width = 7, height = 4)

disease_readdepth <- ggplot(reads_disease, aes(x = reorder(s.SequencingID, TotalReads), y=TotalReads, fill=i.Disease)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 65466, linetype="dashed") +
  scale_fill_manual(name=NULL,
                    values=twocolors,
                    breaks=c("Crohns", "ColorectalCancer"),
                    labels=c("Crohns", "ColorectalCancer"))+
  labs(title = "Disease read counts",
       x = "samples", 
       y = "reads") +
  # scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  # ylim(0,95000)+
  theme(axis.text.x = element_blank())

ggsave(file = "../figures/disease_reads.pdf", plot = disease_readdepth, width = 7, height = 4)

# list samples below 65467
reads_filt <- subset(reads, TotalReads < 65467)
reads_filt_ID <- reads_filt$s.SequencingID
```

## filter out samples below set read count
```{r}
phyloseq_sampleReads <- prune_samples(!(sample_names(phyloseq_prunetaxa) %in%reads_filt_ID), phyloseq_prunetaxa)

# remove taxa with zero abundance
phyloseq_sampleReads <- prune_taxa(taxa_sums(phyloseq_sampleReads)>0, phyloseq_sampleReads)
```

## how many samples per group left after QC?
```{r}
df.mat <- as.matrix(phyloseq_sampleReads@sam_data) # Put sample_data into a ggplot-friendly data.frame
df.mat <- as.data.frame(df.mat)

# subset 
df.mat.disease <- subset(df.mat, i.Specimen == "Right Colon") 
df.mat.location <- subset(df.mat, i.Disease == "Crohns") 

df_disease <- df.mat.disease[, c("i.Disease"), drop=FALSE]
df_location <- df.mat.location[,c("i.Specimen"), drop=FALSE]

# how many samples per condition?
dplyr::count(df_disease, i.Disease)
dplyr::count(df_location, i.Specimen)
```

## save ps object
```{r}
saveRDS(phyloseq_sampleReads, "../output/phyloseq_sampleReads_unadj.rds")
```

## check for significant associations
```{r}
# bray curtis
set.seed(123)
bray_pcoa <- beta_ordinate(phyloseq_sampleReads, beta_dist = "bray")
bray_eig <- bray_pcoa[["values"]][["Relative_eig"]]
bray_eig <- as.data.frame(bray_eig)
bray_eig$cumulative <- cumsum(bray_eig$bray_eig)
### 90% of variation  is first 77 PCOs

## extract vectors
bray_vec <- bray_pcoa[["vectors"]]

## filter to PCs that explain at least 90% of variation
bray_PCOs <- 1:77
bray_PCOs <- paste0('Axis.', bray_PCOs)

bray_vec <- bray_vec[, bray_PCOs]

## remove metadata columns with NA
metadata_ps <- sample_data(phyloseq_sampleReads)

## set up df for check
df_comb_bray <- cbind(bray_vec, metadata_ps)

## set class types
df_comb_bray$i.Sex <- as.factor(df_comb_bray$i.Sex)
df_comb_bray$i.Age <- as.numeric(df_comb_bray$i.Age)
df_comb_bray$i.OR.Age <- as.numeric(df_comb_bray$i.OR.Age)
df_comb_bray$p.CleanedPCRConcentration <- as.numeric(df_comb_bray$p.CleanedPCRConcentration)
df_comb_bray$ex.ExtractionBatch <- as.factor(df_comb_bray$ex.ExtractionBatch)

## check if lined up
identical(rownames(metadata_ps), rownames(bray_vec))

## check for associations
df_associations_bray <- matrix(ncol = 5, nrow = ncol(bray_vec))
colnames(df_associations_bray) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
rownames(df_associations_bray) <- colnames(bray_vec)

set.seed(123)
for(i in colnames(df_associations_bray)){
  for (j in rownames(df_associations_bray)) {
    print(as.formula(get(j) ~ get(i)))
    obj <- print(lm(as.formula(get(j) ~ get(i)), data = df_comb_bray))
    df_associations_bray[j,i] <- summary(obj)$coefficients[,"Pr(>|t|)"][[2]]
  }
}

df_associations_bray_BH <- as.data.frame(t(df_associations_bray))

## multiple test correction
set.seed(123)
df_associations_bray_BH <- sapply(df_associations_bray_BH, p.adjust, method="BH", n = 5)
df_associations_bray_BH <- t(df_associations_bray_BH)
colnames(df_associations_bray_BH) <- colnames(df_associations_bray)
## add variance explained
df_cov_bray <- as.data.frame(df_associations_bray_BH)
df_cov_bray$perc_contrib <- bray_eig[1:77, "bray_eig"]

## reorder
df_cov_bray <- dplyr::select(df_cov_bray, perc_contrib, everything())

# unweighted unifrac
set.seed(123)
uw_pcoa <- beta_ordinate(phyloseq_sampleReads, beta_dist = "unweighted unifrac")
uw_eig <- uw_pcoa[["values"]][["Relative_eig"]]
uw_eig <- as.data.frame(uw_eig)
uw_eig$cumulative <- cumsum(uw_eig$uw_eig)
### 90% of variation  is first 67 PCOs

## extract vectors
uw_vec <- uw_pcoa[["vectors"]]

## filter to PCs that explain at least 90% of variation
uw_PCOs <- 1:67
uw_PCOs <- paste0('Axis.', uw_PCOs)

uw_vec <- uw_vec[, uw_PCOs]

## remove metadata columns with NA
metadata_ps <- sample_data(phyloseq_sampleReads)

## set up df for check
df_comb_uw <- cbind(uw_vec, metadata_ps)

## set class types
df_comb_uw$i.Sex <- as.factor(df_comb_uw$i.Sex)
df_comb_uw$i.Age <- as.numeric(df_comb_uw$i.Age)
df_comb_uw$i.OR.Age <- as.numeric(df_comb_uw$i.OR.Age)
df_comb_uw$p.CleanedPCRConcentration <- as.numeric(df_comb_uw$p.CleanedPCRConcentration)
df_comb_uw$ex.ExtractionBatch <- as.factor(df_comb_uw$ex.ExtractionBatch)

## check if lined up
identical(rownames(metadata_ps), rownames(uw_vec))

## check for associations
df_associations_uw <- matrix(ncol = 5, nrow = ncol(uw_vec))
colnames(df_associations_uw) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
rownames(df_associations_uw) <- colnames(uw_vec)

set.seed(123)
for(i in colnames(df_associations_uw)){
  for (j in rownames(df_associations_uw)) {
    print(as.formula(get(j) ~ get(i)))
    obj <- print(lm(as.formula(get(j) ~ get(i)), data = df_comb_uw))
    df_associations_uw[j,i] <- summary(obj)$coefficients[,"Pr(>|t|)"][[2]]
  }
}

df_associations_uw_BH <- as.data.frame(t(df_associations_uw))

## multiple test correction
set.seed(123)
df_associations_uw_BH <- sapply(df_associations_uw_BH, p.adjust, method="BH", n = 5)
df_associations_uw_BH <- t(df_associations_uw_BH)
colnames(df_associations_uw_BH) <- colnames(df_associations_uw)
## add variance explained
df_cov_uw <- as.data.frame(df_associations_uw_BH)
df_cov_uw$perc_contrib <- uw_eig[1:67, "uw_eig"]

## reorder
df_cov_uw <- dplyr::select(df_cov_uw, perc_contrib, everything())


# weighted unifrac
set.seed(123)
w_pcoa <- beta_ordinate(phyloseq_sampleReads, beta_dist = "weighted unifrac")
w_eig <- w_pcoa[["values"]][["Relative_eig"]]
w_eig <- as.data.frame(w_eig)
w_eig$cumulative <- cumsum(w_eig$w_eig)
### 90% of variation  is first 63 PCOs

## extract vectors
w_vec <- w_pcoa[["vectors"]]

## filter to PCs that explain at least 90% of variation
w_PCOs <- 1:18
w_PCOs <- paste0('Axis.', w_PCOs)

w_vec <- w_vec[, w_PCOs]

## remove metadata columns with NA
metadata_ps <- sample_data(phyloseq_sampleReads)

## set up df for check
df_comb_w <- cbind(w_vec, metadata_ps)

## set class types
df_comb_w$i.Sex <- as.factor(df_comb_w$i.Sex)
df_comb_w$i.Age <- as.numeric(df_comb_w$i.Age)
df_comb_w$i.OR.Age <- as.numeric(df_comb_w$i.OR.Age)
df_comb_w$p.CleanedPCRConcentration <- as.numeric(df_comb_w$p.CleanedPCRConcentration)
df_comb_w$ex.ExtractionBatch <- as.factor(df_comb_w$ex.ExtractionBatch)

## check if lined up
identical(rownames(metadata_ps), rownames(w_vec))


## check for associations
df_associations_w <- matrix(ncol = 5, nrow = ncol(w_vec))
colnames(df_associations_w) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
rownames(df_associations_w) <- colnames(w_vec)

set.seed(123)
for(i in colnames(df_associations_w)){
  for (j in rownames(df_associations_w)) {
    print(as.formula(get(j) ~ get(i)))
    obj <- print(lm(as.formula(get(j) ~ get(i)), data = df_comb_w))
    df_associations_w[j,i] <- summary(obj)$coefficients[,"Pr(>|t|)"][[2]]
  }
}

df_associations_w_BH <- as.data.frame(t(df_associations_w))

## multiple test correction
set.seed(123)
df_associations_w_BH <- sapply(df_associations_w_BH, p.adjust, method="BH", n = 5)
df_associations_w_BH <- t(df_associations_w_BH)
colnames(df_associations_w_BH) <- colnames(df_associations_w)
## add variance explained
df_cov_w <- as.data.frame(df_associations_w_BH)
df_cov_w$perc_contrib <- w_eig[1:18, "w_eig"]

## reorder
df_cov_w <- dplyr::select(df_cov_w, perc_contrib, everything())
```

## save covariate dfs
```{r}
write.csv(df_cov_bray, "../output/covariate_bray.csv")
write.csv(df_cov_uw, "../output/covariate_uw.csv")
write.csv(df_cov_w, "../output/covariate_w.csv")
```

## plot covariate associations
```{r}
# bray
df_associations_bray_BH_df <- as.data.frame(df_associations_bray_BH)

df_associations_bray_BH_df$axis <- rownames(df_associations_bray_BH_df)

## convert format
df_associations_bray_BH_long <- pivot_longer(df_associations_bray_BH_df, !axis, names_to = "covariate", values_to = "padj")

## order by axis
df_associations_bray_BH_long <- mutate(df_associations_bray_BH_long, axis = factor(axis, levels = unique(rownames(df_associations_bray_BH_df))))

ggplot(df_associations_bray_BH_long, aes(x = covariate, y = axis, fill = padj)) +
  geom_tile() +
  scale_fill_gradient(high="white", low="#009194", limits=c(0,1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  labs(x = "Covariate", y = "Axis")

# uw
df_associations_uw_BH_df <- as.data.frame(df_associations_uw_BH)

df_associations_uw_BH_df$axis <- rownames(df_associations_uw_BH_df)

## convert format
df_associations_uw_BH_long <- pivot_longer(df_associations_uw_BH_df, !axis, names_to = "covariate", values_to = "padj")

## order by axis
df_associations_uw_BH_long <- mutate(df_associations_uw_BH_long, axis = factor(axis, levels = unique(rownames(df_associations_uw_BH_df))))

ggplot(df_associations_uw_BH_long, aes(x = covariate, y = axis, fill = padj)) +
  geom_tile() +
  scale_fill_gradient(high="white", low="#009194", limits=c(0,1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  labs(x = "Covariate", y = "Axis")

# w
df_associations_w_BH_df <- as.data.frame(df_associations_w_BH)

df_associations_w_BH_df$axis <- rownames(df_associations_w_BH_df)

## convert format
df_associations_w_BH_long <- pivot_longer(df_associations_w_BH_df, !axis, names_to = "covariate", values_to = "padj")

## order by axis
df_associations_w_BH_long <- mutate(df_associations_w_BH_long, axis = factor(axis, levels = unique(rownames(df_associations_w_BH_df))))

ggplot(df_associations_w_BH_long, aes(x = covariate, y = axis, fill = padj)) +
  geom_tile() +
  scale_fill_gradient(high="white", low="#009194", limits=c(0,1))+
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95, vjust=0.95)) +
  labs(x = "Covariate", y = "Axis")
```

## check for random effects
```{r}
df_comb_bray$i.PatientID <- as.factor(df_comb_bray$i.PatientID)
df_comb_uw$i.PatientID <- as.factor(df_comb_uw$i.PatientID)
df_comb_w$i.PatientID <- as.factor(df_comb_w$i.PatientID)

# patientID
## bray
### set up vec
patient_rand_bray <- c()
n = 1

### run
set.seed(123)
for(i in colnames(bray_vec)){
  full_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_bray)
  reduced_model <- lm(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_bray)
  lr <- lrtest(full_model, reduced_model)
  patient_rand_bray <- c(patient_rand_bray, lr["2", "Pr(>Chisq)"])
  print(n)
  n = n + 1
}

df_cov_bray$patient_rand <- patient_rand_bray

## uw
### set up vec
patient_rand_uw <- c()
n = 1

### run
set.seed(123)
for(i in colnames(uw_vec)){
  full_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_uw)
  reduced_model <- lm(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_uw)
  lr <- lrtest(full_model, reduced_model)
  patient_rand_uw <- c(patient_rand_uw, lr["2", "Pr(>Chisq)"])
  print(n)
  n = n + 1
}

df_cov_uw$patient_rand <- patient_rand_uw

## w
### set up vec
patient_rand_w <- c()
n = 1

### run
set.seed(123)
for(i in colnames(w_vec)){
  full_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_w)
  reduced_model <- lm(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_w)
  lr <- lrtest(full_model, reduced_model)
  patient_rand_w <- c(patient_rand_w, lr["2", "Pr(>Chisq)"])
  print(n)
  n = n + 1
}

df_cov_w$patient_rand <- patient_rand_w
```

## check for collinearity between variables
```{r}
# covariates
## bray
### set up df
cov_collinear_bray <- matrix(nrow = 5, ncol = ncol(bray_vec))
colnames(cov_collinear_bray) <- colnames(bray_vec)
rownames(cov_collinear_bray) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
n = 1

set.seed(123)
for(i in colnames(cov_collinear_bray)){
  cov_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_bray)
  vif_cov <- as.data.frame(vif(cov_model))
  cov_collinear_bray[,i] <- vif_cov[,3]
  print(n)
  n = n + 1
}

## uw
### set up df
cov_collinear_uw <- matrix(nrow = 5, ncol = ncol(uw_vec))
colnames(cov_collinear_uw) <- colnames(uw_vec)
rownames(cov_collinear_uw) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
n = 1

### run
set.seed(123)
for(i in colnames(cov_collinear_uw)){
  cov_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_uw)
  vif_cov <- as.data.frame(vif(cov_model))
  cov_collinear_uw[,i] <- vif_cov[,3]
  print(n)
  n = n + 1
}

## w
### set up vec
cov_collinear_w <- matrix(nrow = 5, ncol = ncol(w_vec))
colnames(cov_collinear_w) <- colnames(w_vec)
rownames(cov_collinear_w) <- c("i.Sex", "i.Age", "i.OR.Age", "p.CleanedPCRConcentration", "ex.ExtractionBatch")
n = 1

### run
set.seed(123)
for(i in colnames(cov_collinear_w)){
  cov_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_w)
  vif_cov <- as.data.frame(vif(cov_model))
  cov_collinear_w[,i] <- vif_cov[,3]
  print(n)
  n = n + 1
}


# variables of interest
## bray
### set up df
collinear_bray <- matrix(ncol = 2, nrow = ncol(bray_vec))
colnames(collinear_bray) <- c("i.Specimen", "i.Disease")
rownames(collinear_bray) <- colnames(bray_vec)
n = 1

### run
set.seed(123)
for(i in colnames(bray_vec)){
  spec_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Specimen + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_bray)
  vif_spec <- vif(spec_model)
  collinear_bray[i, "i.Specimen"] <- vif_spec["i.Specimen", "GVIF^(1/(2*Df))"]
  disease_model <- lm(as.formula(paste0("`", i, "`", "~ i.Disease + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_bray)
  vif_disease <- vif(disease_model)
  collinear_bray[i, "i.Disease"] <- vif_disease["i.Disease", "GVIF^(1/(2*Df))"]
  print(n)
  n = n + 1
}


## uw
### set up df
collinear_uw <- matrix(ncol = 2, nrow = ncol(uw_vec))
colnames(collinear_uw) <- c("i.Specimen", "i.Disease")
rownames(collinear_uw) <- colnames(uw_vec)
n = 1

### run
set.seed(123)
for(i in colnames(uw_vec)){
  spec_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Specimen + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_uw)
  vif_spec <- vif(spec_model)
  collinear_uw[i, "i.Specimen"] <- vif_spec["i.Specimen", "GVIF^(1/(2*Df))"]
  disease_model <- lm(as.formula(paste0("`", i, "`", "~ i.Disease + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_uw)
  vif_disease <- vif(disease_model)
  collinear_uw[i, "i.Disease"] <- vif_disease["i.Disease", "GVIF^(1/(2*Df))"]
  print(n)
  n = n + 1
}

## w
### set up vec
collinear_w <- matrix(ncol = 2, nrow = ncol(w_vec))
colnames(collinear_w) <- c("i.Specimen", "i.Disease")
rownames(collinear_w) <- colnames(w_vec)
n = 1

### run
set.seed(123)
for(i in colnames(w_vec)){
  spec_model <- lmer(as.formula(paste0("`", i, "`", "~ i.Specimen + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")), data = df_comb_w)
  vif_spec <- vif(spec_model)
  collinear_w[i, "i.Specimen"] <- vif_spec["i.Specimen", "GVIF^(1/(2*Df))"]
  disease_model <- lm(as.formula(paste0("`", i, "`", "~ i.Disease + i.Sex + i.Age + i.OR.Age + p.CleanedPCRConcentration + ex.ExtractionBatch")), data = df_comb_w)
  vif_disease <- vif(disease_model)
  collinear_w[i, "i.Disease"] <- vif_disease["i.Disease", "GVIF^(1/(2*Df))"]
  print(n)
  n = n + 1
}
```

## correct w/o OR age and check if ORage still associated
```{r}
# bray
## set up df
bray_corr <- matrix(nrow = nrow(bray_vec), ncol = ncol(bray_vec))
colnames(bray_corr) <- colnames(bray_vec)
rownames(bray_corr) <- rownames(bray_vec)

n = 1

set.seed(123)
for(i in colnames(bray_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")),  data=df_comb_bray)
  bray_corr[, i] <- resid(obj)
  n = n + 1
  print(n)
}

## check for associations
df_comb_bray_corr <- cbind(bray_corr, metadata_ps)
## check if lined up
identical(rownames(metadata_ps), rownames(bray_corr))

bray_ORage_vector <- c()

set.seed(123)
for(i in colnames(bray_corr)){
  obj <- lm(as.formula(paste0(i, "~ i.OR.Age")),  data=df_comb_bray_corr)
  bray_ORage_vector = c(bray_ORage_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

## put results in df
bray_ORage_check <- data.frame(matrix(ncol = 1, nrow = 77))
rownames(bray_ORage_check) <- colnames(bray_corr)
colnames(bray_ORage_check) <- c("p_ORage")

bray_ORage_check$p_ORage <- bray_ORage_vector


# uw
## set up df
uw_corr <- matrix(nrow = nrow(uw_vec), ncol = ncol(uw_vec))
colnames(uw_corr) <- colnames(uw_vec)
rownames(uw_corr) <- rownames(uw_vec)

n = 1

set.seed(123)
for(i in colnames(uw_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")),  data=df_comb_uw)
  uw_corr[, i] <- resid(obj)
  n = n + 1
  print(n)
}

## check for associations
df_comb_uw_corr <- cbind(uw_corr, metadata_ps)
## check if lined up
identical(rownames(metadata_ps), rownames(uw_corr))

uw_ORage_vector <- c()

set.seed(123)
for(i in colnames(uw_corr)){
  obj <- lm(as.formula(paste0(i, "~ i.OR.Age")),  data=df_comb_uw_corr)
  uw_ORage_vector = c(uw_ORage_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

## put results in df
uw_ORage_check <- data.frame(matrix(ncol = 1, nrow = 65))
rownames(uw_ORage_check) <- colnames(uw_corr)
colnames(uw_ORage_check) <- c("p_ORage")

uw_ORage_check$p_ORage <- uw_ORage_vector


# w
## set up df
w_corr <- matrix(nrow = nrow(w_vec), ncol = ncol(w_vec))
colnames(w_corr) <- colnames(w_vec)
rownames(w_corr) <- rownames(w_vec)

n = 1

set.seed(123)
for(i in colnames(w_corr)){
  obj <- lmer(as.formula(paste0("`", i, "`", "~ i.Sex + i.Age + p.CleanedPCRConcentration + ex.ExtractionBatch + (1 | i.PatientID)")),  data=df_comb_w)
  w_corr[, i] <- resid(obj)
  n = n + 1
  print(n)
}

## check for associations
df_comb_w_corr <- cbind(w_corr, metadata_ps)
## check if lined up
identical(rownames(metadata_ps), rownames(w_corr))

w_ORage_vector <- c()

set.seed(123)
for(i in colnames(w_corr)){
  obj <- lm(as.formula(paste0(i, "~ i.OR.Age")),  data=df_comb_w_corr)
  w_ORage_vector = c(w_ORage_vector, summary(obj)$coefficients[,"Pr(>|t|)"][[2]])
}

## put results in df
w_ORage_check <- data.frame(matrix(ncol = 1, nrow = 19))
rownames(w_ORage_check) <- colnames(w_corr)
colnames(w_ORage_check) <- c("p_ORage")

w_ORage_check$p_ORage <- w_ORage_vector
```

## generate rarefaction curves
```{r}
# function for rarefaction curve
calculate_rarefaction_curves <- function(psdata, measures, depths, parallel=FALSE) {
  require('plyr') # ldply
  require('reshape2') # melt
  require('doParallel')

  # set parallel options if required
  if (parallel) {
    paropts  <- list(.packages=c("phyloseq", "reshape2"))
  } else {
    paropts  <- NULL
  }

  estimate_rarified_richness <- function(psdata, measures, depth) {
    if(max(sample_sums(psdata)) < depth) return()
    psdata <- prune_samples(sample_sums(psdata) >= depth, psdata)

    rarified_psdata <- rarefy_even_depth(psdata, depth, verbose = FALSE)

    alpha_diversity <- estimate_richness(rarified_psdata, measures = measures)

    # as.matrix forces the use of melt.array, which includes the Sample names (rownames)
    molten_alpha_diversity <- melt(as.matrix(alpha_diversity), varnames = c('Sample', 'Measure'), value.name = 'Alpha_diversity')

    molten_alpha_diversity
  }

  names(depths) <- depths # this enables automatic addition of the Depth to the output by ldply
  rarefaction_curve_data <- ldply(depths, estimate_rarified_richness, psdata = phyloseq_prunetaxa, measures = measures, .id = 'Depth', .progress = ifelse(interactive() && ! parallel, 'text', 'none'), .parallel=parallel, .paropts=paropts)

  # convert Depth from factor to numeric
  rarefaction_curve_data$Depth <- as.numeric(levels(rarefaction_curve_data$Depth))[rarefaction_curve_data$Depth]

  rarefaction_curve_data
}

# calculate rarefaction curves
rarefaction_curve_data <- calculate_rarefaction_curves(phyloseq_sampleReads, c('Observed', 'Shannon'), rep(c(1, 10, 100, 1000, 1:100 * 10000), each = 10))
summary(rarefaction_curve_data)

rarefaction_curve_data_summary <- ddply(rarefaction_curve_data, c('Depth', 'Sample', 'Measure'), summarise, Alpha_diversity_mean = mean(Alpha_diversity), Alpha_diversity_sd = sd(Alpha_diversity))

rarefaction_curve_data_summary_verbose <- merge(rarefaction_curve_data_summary, data.frame(sample_data(phyloseq_sampleReads)), by.x = 'Sample', by.y = 'row.names')

# plot rarefaction curves
rarefy <- ggplot(
  data = rarefaction_curve_data_summary_verbose,
  mapping = aes(
    x = Depth,
    y = Alpha_diversity_mean,
    ymin = Alpha_diversity_mean - Alpha_diversity_sd,
    ymax = Alpha_diversity_mean + Alpha_diversity_sd,
    colour = i.Specimen,
    group = Sample,
  )
) + geom_line(alpha = 0.5
) + geom_point(alpha = 0.5) + 
  scale_color_manual(name=NULL,
                    values=threecolors,
                    breaks=c("Terminal Ileum", "Cecum", "Right Colon"),
                    labels=c("Terminal Ileum", "Cecum", "Right Colon")) +
  facet_wrap(
  facets = ~ Measure,
  scales = 'free_y'
)

# ggsave(file = "figures/rarefy.pdf", plot = rarefy, width=8, height=6)

# function for calculating alpha diversity
 compute_alphadiv<-function(phyloseqobj,trials,rarefaction_depth){
    min_lib = rarefaction_depth
    nsamp = phyloseq::nsamples(phyloseqobj)
    trials = trials
    tree = phy_tree(phyloseqobj)
   
    richness <- matrix(nrow = nsamp, ncol = trials)
    row.names(richness) <- sample_names(phyloseqobj)
    
    evennessF <- matrix(nrow = nsamp, ncol = trials)
    row.names(evennessF) <- sample_names(phyloseqobj)
    
    evennessS <- matrix(nrow = nsamp, ncol = trials)
    row.names(evennessS) <- sample_names(phyloseqobj)
    
    evennessI <- matrix(nrow = nsamp, ncol = trials)
    row.names(evennessI) <- sample_names(phyloseqobj)
    
    faiths <- matrix(nrow = nsamp, ncol = trials)
    row.names(faiths) <- sample_names(phyloseqobj)
  
    set.seed(007)
    for (i in 1:trials) {
      # Subsample
      r <- rarefy_even_depth(phyloseqobj, sample.size = min_lib, verbose = TRUE, replace = TRUE)
      
      # Calculate richness
      rich <- as.numeric(as.matrix(estimate_richness(r, measures = "Observed")))
      richness[ ,i] <- rich
      
      # Calculate evenness Fisher
      evenF <- as.numeric(as.matrix(estimate_richness(r, measures = "Fisher")))
      evennessF[ ,i] <- evenF
    
      # Calculate evenness Shannon
      evenS <- as.numeric(as.matrix(estimate_richness(r, measures = "Shannon")))
      evennessS[ ,i] <- evenS
      
      # Calculate evenness Simpson
      evenI <- as.numeric(as.matrix(estimate_richness(r, measures = "Simpson")))
      evennessI[ ,i] <- evenI
      
      # calculate faith's PD
      pd_all <- pd(as.matrix(r@otu_table), tree, include.root=TRUE)
      pd_extract <- subset(pd_all, select = -c(SR))
      faithspd <- as.numeric(as.matrix(pd_extract))
      faiths[ ,i] <- faithspd
    }
  
    s.SequencingID <- row.names(richness)
    mean <- apply(richness, 1, mean)
    sd <- apply(richness, 1, sd)
    measure <- rep("Richness", nsamp)
    rich_stats <- data.frame(s.SequencingID, mean, sd, measure)
  
    s.SequencingID <- row.names(evennessF)
    mean <- apply(evennessF, 1, mean)
    sd <- apply(evennessF, 1, sd)
    measure <- rep("Fisher", nsamp)
    even_statsF <- data.frame(s.SequencingID, mean, sd, measure)
  
    s.SequencingID <- row.names(evennessS)
    mean <- apply(evennessS, 1, mean)
    sd <- apply(evennessS, 1, sd)
    measure <- rep("Shannon", nsamp)
    even_statsS <- data.frame(s.SequencingID, mean, sd, measure)
  
    s.SequencingID <- row.names(evennessI)
    mean <- apply(evennessI, 1, mean)
    sd <- apply(evennessI, 1, sd)
    measure <- rep("Simpson", nsamp)
    even_statsI <- data.frame(s.SequencingID, mean, sd, measure)
    
    s.SequencingID <- row.names(faiths)
    mean <- apply(faiths, 1, mean)
    sd <- apply(faiths, 1, sd)
    measure <- rep("Faiths", nsamp)
    faiths_stats <- data.frame(s.SequencingID, mean, sd, measure)
  
    alpha <- rbind(rich_stats, even_statsF, even_statsS, even_statsI, faiths_stats)
  
    s <- data.frame(sample_data(phyloseqobj))
    alphadiv <- merge(alpha, s, by = "s.SequencingID")
    alphadiv$rarefaction=rep(min_lib,dim(alphadiv)[1])
    return(alphadiv)
 } 
 
# calculate alpha diversity - include number of trials, number of sequences to subsample
set.seed(100)
rfxn <- compute_alphadiv(phyloseq_sampleReads,1000,65467) # this might take a couple hours to run
```

## save rfxn
```{r}
write.csv(rfxn, file = "../output/rfxn_uncorrect.csv")
```

## baseline stats
```{r}
# samples per location prep
df.mat_micro <- as.matrix(phyloseq_sampleReads@sam_data) # Put sample_data into a ggplot-friendly data.frame
df.mat_micro <- as.data.frame(df.mat_micro)
df.mat_micro <- subset(df.mat_micro, i.Disease == "Crohns")
df_spec_micro <- df.mat_micro[,c("i.Specimen"), drop=FALSE]

# how many samples per location?
df_spec_micro %>% 
  group_by(i.Specimen) %>%
  dplyr::count()
```

